### 2021-05-24 generics.tex
First commit

\documentclass{llncs}

\usepackage{url}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage[hidelinks]{hyperref}
\usepackage{pdfpages}
\usepackage{relsize}

% Solidity
\newcommand{\uint}{\texttt{uint}}
\newcommand{\SafeMath}{\texttt{SafeMath}}
\newcommand{\Storage}{\texttt{Storage}}
\newcommand{\Contract}{\texttt{Contract}}
% Java
\newcommand{\UBI}{\texttt{UnsignedBigInteger}}
\newcommand{\UBIshort}{\texttt{UBI}}
\newcommand{\BI}{\texttt{BigInteger}}
\newcommand{\ie}{\textit{, ie.\ }}

\def\codesize{\smaller}
\def\<#1>{\codeid{#1}}
\newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}

\begin{document}

\title{Power and Pitfalls of Generic Smart Contract}
\titlerunning{Power and Pitfalls of Generics for Smart Contracts}
\author{Andrea Benini \and Sara Migliorini \and Fausto Spoto}
\institute{Dipartimento di Informatica, Universit\`a di Verona, Italy}

\maketitle

\begin{abstract}
  TBD
  \keywords{smart contract \and generics \and blockchain}
\end{abstract}

\section{Introduction}\label{sec:introduction}

Blockchains exploit the redundant, concurrent execution of the same
transactions on a decentralized network of many machines
in order to enforce their execution in accordance to
a set of predefined rules. Namely, blockchains make it hard, for a single machine,
to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
has been the first blockchain's success story. Here
transactions are transfers of units of coins between accounts, with the specific rule that
the same inputs cannot be spent twice. Bitcoin's coins are called bitcoins
themselves and have been the first example of a blockchain \emph{token}.

A few years after Bitcoin, another blockchain, called
Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
transactions with an actual, imperative programming language, called Solidity.
Ethereum's transactions are still paid in terms of its native \emph{ether} token,
but they execute much more than token transfers. Namely,
transactions run object constructors and methods, which results in a sort
of \emph{world computer} that persists the same objects in the memory of all the
computers in the blockchain's network. These computers must execute
the transactions according to Solidity programs called
\emph{smart contracts}, or will end up being put out of consensus. Among such programs,
a few have emerged, that implement a dynamic ledger of coin transfers between accounts. These
coins are not the native ether, but rather new, derived tokens.

The rest of this paper is organized as follows.

\section{Shared Entities using Generics}\label{sec:shared_entities}

\section{An Attack to the Shared Entities Contract}\label{sec:attack}

\section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}

\section{Conclusion}\label{sec:conclusion}

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2021-05-24 generics.tex
Draft of the introduction

@@
 
-\title{Power and Pitfalls of Generic Smart Contract}
+\title{Power and Pitfalls of Generic Smart Contracts}
 \titlerunning{Power and Pitfalls of Generics for Smart Contracts}
@@
 has been the first blockchain's success story. Here
-transactions are transfers of units of coins between accounts, with the specific rule that
-the same inputs cannot be spent twice. Bitcoin's coins are called bitcoins
-themselves and have been the first example of a blockchain \emph{token}.
+transactions are programmed in a non-Turing complete bytecode language,
+almost exclusively used to implement transfers of units of coins between \emph{accounts}.
 
@@
 Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
-transactions with an actual, imperative programming language, called Solidity.
-Ethereum's transactions are still paid in terms of its native \emph{ether} token,
-but they execute much more than token transfers. Namely,
-transactions run object constructors and methods, which results in a sort
+transactions in an actual, imperative and Turing-complete programming language, called Solidity.
+Solidity's code is organized in \emph{smart contracts}, that can be seen as
+objects that control money.
+Ethereum's transactions can hence execute much more than coin transfers. Namely,
+they run object constructors and methods, which results in a sort
 of \emph{world computer} that persists the same objects in the memory of all the
-computers in the blockchain's network. These computers must execute
-the transactions according to Solidity programs called
-\emph{smart contracts}, or will end up being put out of consensus. Among such programs,
-a few have emerged, that implement a dynamic ledger of coin transfers between accounts. These
-coins are not the native ether, but rather new, derived tokens.
+computers in the blockchain's network.
+
+In Solidity's bytecode,
+non-primitive values are referenced through a very general
+\<address> type. For instance, a Solidity method
+\<child(Person p, uint256 n) returns Person> actually compiles
+into \<child(address p, uint256 n) returns address>, losing most
+type information~\cite{CrafaPZ19}.
+At run time, it is the bytecode that gets executed. Hence,
+everything can be passed for \<p>, not just a \<Person>.
+The compiler cannot even enforce strong typing
+by generating defensive type instance checks and casts, since
+values are unboxed in Ethereum: they have no attached
+type information at run time,
+they are just numerical \emph{addresses}.
+It follows that calls to \<Person>'s methods
+on \<p> might actually execute arbitrary code, if \<p> is not a \<Person>.
+In other words, Solidity is not strongly typed.
+Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
+to our methods, such as on \<p> passed to \<child>, since an attacker can pass crafted
+objects for \<p>, with arbitrary implementations for their methods,
+which can result in the unexpected execution of
+dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
+costed millions of dollars.
+
+Strong typing is one of the reasons that push towards the adoption
+of \emph{traditional} programming languages for smart contracts. For instance,
+the Cosmos blockchain~\cite{cosmos} uses Go. The
+Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
+for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
+Hyperledger~\cite{hyperldeger} allows Go and Java.
+Another reason is the availability of modern
+language features, missing in Solidity,
+such as \emph{generics}\ie the possibility of using
+type variables. Generics are powerful and very useful for programming
+smart contracts. In Java source code, they are strongly typed, if no \emph{unchecked operations}
+are used~\cite{NaftalinW06}, as it will always be the case in this paper.
+However, generics in Java are compiled by \emph{erasure}\ie replaced
+by their upwards bound, most often \<Object>. Hence, the risk
+is that generics introduce the same kind of attack to the bytecode
+as normal reference types do in Solidity.
+
+The contribution of this paper is to show a real-life
+use of generics for an actual smart contract used in the support
+library of the Takamaka language, and to show that a naive use
+of Java generics can lead to a code security vulnerability that
+allows an attacker to earn money by exploiting someone else's work.
+This paper will provide a fix to that specific issue,
+by obliging the compiler to generate defensive checks.
+More generally, this paper can be useful for the definition of
+bytecode languages for future smart contract languages, by
+learning from the weaknesses of Java bytecode.
 
 The rest of this paper is organized as follows.
+Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
+contracts that uses generics. Sec.~\ref{sec:attack} shows that a naive
+deployment of that contract leads to a code vulnerability.
+Sec.~\ref{sec:fix} shows a fix to that vulnerability.
+Sec.~\ref{sec:conclusion} concludes.
 

### 2021-06-08 generics.tex
Started describing the shared entity

@@
 
-% Solidity
-\newcommand{\uint}{\texttt{uint}}
-\newcommand{\SafeMath}{\texttt{SafeMath}}
-\newcommand{\Storage}{\texttt{Storage}}
-\newcommand{\Contract}{\texttt{Contract}}
-% Java
-\newcommand{\UBI}{\texttt{UnsignedBigInteger}}
-\newcommand{\UBIshort}{\texttt{UBI}}
-\newcommand{\BI}{\texttt{BigInteger}}
+\usepackage{svg}
+\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\includesvg[height = 2ex]{svg-inkscape/ORCID_iD}}}
+
 \newcommand{\ie}{\textit{, ie.\ }}
@@
 \newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}
+
+\usepackage{listings, xcolor}
+
+\renewcommand{\UrlFont}{\ttfamily\codesize}
+
+\definecolor{verylightgray}{rgb}{.97,.97,.97}
+
+\lstdefinelanguage{Takamaka}{
+        keywords=[1]{abstract, break, case, catch, class, continue, default, do
+, else, false, finally, for, if, final, implements, extends, import, instanceof, interface, length, new, private, protected, public, return, super, switch, this, throw, true, try, while, var, null}, % generic keywords
+        keywordstyle=[1]\color{blue}\bfseries,
+        keywords=[2]{boolean, int, long, float, double, byte, short, char, void, enum}, % types; money and time units
+        keywordstyle=[2]\color{teal}\bfseries,
+        keywords=[3]{@Override,@View,@FromContract,@Payable}, % annotations
+        keywordstyle=[3]\color{violet}\bfseries,
+        identifierstyle=\color{black},
+        sensitive=false,
+        comment=[l]{//},
+        morecomment=[s]{/*}{*/},
+        commentstyle=\color{gray}\ttfamily,
+        stringstyle=\color{red}\ttfamily,
+        morestring=[b]',
+        morestring=[b]"
+}
+
+\lstset{
+        language=Takamaka,
+        backgroundcolor=\color{verylightgray},
+        extendedchars=true,
+        basicstyle=\scriptsize\ttfamily,
+        showstringspaces=false,
+        showspaces=false,
+        numbers=none,
+        numberstyle=\scriptsize,
+        numbersep=9pt,
+        tabsize=2,
+        breaklines=true,
+        showtabs=false,
+        captionpos=b
+}
 
@@
 
+A \emph{shared entity} is something divided into \emph{shares}. Participants,
+that hold shares, are called \emph{shareholders} and can dynamically
+sell and buy shares. An example of a shared entity is a comporation,
+where shares represent units of possess of the company. Another example is
+a voting community, where shares represent the voting power of each given voter.
+A further example are the validator nodes of a proof of stake blockchain,
+where shares represent their voting power and remuneration percentage.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
+{
+  @View
+  BigInteger sharesOf(S shareholder);
+
+  @FromContract(PayableContract.class) @Payable
+  void place(BigInteger amount, O offer);
+
+  @FromContract(PayableContract.class) @Payable
+  void accept(BigInteger amount, S buyer, O offer);
+
+  class Offer<S extends PayableContract> extends Storage {
+    public final S seller;
+    public final BigInteger sharesOnSale;
+    public final BigInteger cost;
+    public final long expiration;
+
+    public Offer
+        (S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
+
+      this.seller = seller;
+      this.sharesOnSale = sharesOnSale;
+      this.cost = cost;
+      this.expiration = now() + duration;
+    }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{A simplified subset of our shared entity interface.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
+\end{figure}
+
+In general, two concepts are specific to each implementation of shared entities:
+who are the potential shareholders and how offers for selling shares work.
+Therefore, we have parameterized the interface of a shared entity with two type variables:
+\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
+Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
+It includes an inner class \<Offer> that models sale offers:
+it specifies who is the seller of the shares,
+how many shares are being sold, the requested price and the expiration of the sale offer.
+Implementations might subclass \<Offer> if they need further properties in a sale offer.
+By using this class \<Offer>, the \<SharedEntity> interface specifies three methods.
+Method \<sharesOf> allows one to know how many shares a potential shareholder (of type \<S>) holds.
+It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
+\emph{for free}, without paying gas, since it has no side-effects. Method \<place>
+activates a new sale offer of shares. This method is annotated as \<@Payable> since
+implementations are allowed to require a payment of \<amount> coins for managing the sale.
+Method \<accept> is called by who accepts an offer, hence becoming a new shareholder or increasing
+its total amount of shares (if it was already a shareholder).
+Also this method is \<@Payable>, since its caller must pay the price imposed by the seller
+and specified in the \<Offer> instance. Method \<accept> forwards this price to the seller,
+which is why shareholders must be \<PayableContract>s: in Takamaka, this means that they
+can receive payments.
+
+The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
+contracts can call these methods. This is important to attest, explicity, the will
+of sharehodlers to sell or buy shares. Therefore,
+the callers of both methods should have type \<S>. In Takamaka, this could be written
+as \<@FromContract(S.class)>. However, Java does not allow a generic type variable \<S>
+in \<S.class>. Because of this limitation,
+the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
+which allows \emph{any} \<PayableContract> to call these methods, not just those of type \<S>.
+Since the syntax of the language does not support our abstraction, we will have to
+implement explicit dynamic checks in code, as shown in a moment.
+
 \section{An Attack to the Shared Entities Contract}\label{sec:attack}

### 2021-06-08 generics.tex
Added implementation of the shared entities

@@
 \usepackage{relsize}
+\usepackage{tcolorbox}
 
@@
     \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
-{
+public interface SharedEntity<S extends PayableContract,O extends Offer<S>> {
+
   @View
@@
     }
+
+    @View
+    public boolean isOngoing() {
+      return now() <= expiration;
+    }
+  }
 }
@@
   \end{center}
-  \caption{A simplified subset of our shared entity interface.
+  \caption{A simplified part of our shared entity interface.
   Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
@@
 it specifies who is the seller of the shares,
-how many shares are being sold, the requested price and the expiration of the sale offer.
-Implementations might subclass \<Offer> if they need further properties in a sale offer.
-By using this class \<Offer>, the \<SharedEntity> interface specifies three methods.
-Method \<sharesOf> allows one to know how many shares a potential shareholder (of type \<S>) holds.
+how many shares are being sold, the requested price and the expiration of the offer.
+Method \<isOngoing> checks if an offer has not expired yet.
+Implementations can subclass \<Offer> if they need more specific offers.
+By using class \<Offer>, the \<SharedEntity> interface specifies three methods.
+Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
 It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
-\emph{for free}, without paying gas, since it has no side-effects. Method \<place>
-activates a new sale offer of shares. This method is annotated as \<@Payable> since
+\emph{for free}, without paying gas, since it has no side-effects. Who wants to sell
+(some of) its shares calls method \<place> with a sale \<offer>.
+This method is annotated as \<@Payable> since
 implementations are allowed to require a payment of \<amount> coins for managing the sale.
-Method \<accept> is called by who accepts an offer, hence becoming a new shareholder or increasing
-its total amount of shares (if it was already a shareholder).
-Also this method is \<@Payable>, since its caller must pay the price imposed by the seller
-and specified in the \<Offer> instance. Method \<accept> forwards this price to the seller,
-which is why shareholders must be \<PayableContract>s: in Takamaka, this means that they
-can receive payments.
+Who buys the shares calls method \<accept> with the accepted \<offer>
+and with itself as \<buyer> (the reason will be explained soon)
+and becomes a new shareholder or increases
+its cumulative number of shares (if it was a shareholder already).
+Also this method is \<@Payable>, since its caller must pay \<amount> $\ge$ \<offer.cost>
+coins to the seller.
+This means that shareholders must be able to receive payments and that
+is why \<S extends PayableContract>: the \<PayableContract>s are those that can receive
+payments in Takamaka.
 
 The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This is important to attest, explicity, the will
-of sharehodlers to sell or buy shares. Therefore,
-the callers of both methods should have type \<S>. In Takamaka, this could be written
-as \<@FromContract(S.class)>. However, Java does not allow a generic type variable \<S>
+contracts can call these methods. This attests, explicity, their will
+to sell or buy shares. The callers must be (old or new) shareholders,
+hence of type \<S>. In Takamaka, this could be written
+as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
 in \<S.class>. Because of this limitation,
@@
 Since the syntax of the language does not support our abstraction, we will have to
-implement explicit dynamic checks in code, as shown in a moment.
+program explicit dynamic checks in code, as shown in a moment.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class SimpleSharedEntity
+      <S extends PayableContract,O extends Offer<S>>
+      extends Contract
+      implements SharedEntity<S,O> {
+
+  private final StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
+  private final StorageSet<O> offers = new StorageTreeSet<>();
+
+  public SimpleSharedEntity(S[] shareholders, BigInteger[] shares) {
+    require(shareholders.length == shares.length, "length mismatch");
+    for (int pos = 0; pos < shareholders.length; pos++)
+      addShares(shareholders[pos], shares[pos]);
+  }
+
+  @Override @View
+  public final BigInteger sharesOf(S shareholder) {
+    return shares.getOrDefault(shareholder, BigInteger.ZERO);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void place(BigInteger amount, O offer) {
+    require(offer.seller == caller(), "not authorized to sell");
+    require(shares.containsKey(offer.seller), "only shareholders can sell");
+    require(sharesOf(offer.seller).subtract(sharesOnSaleOf(offer.seller))
+        .compareTo(offer.sharesOnSale) >= 0, "not enough shares to sell");
+    offers.add(offer);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void accept(BigInteger amount, S buyer, O offer) {
+    require(caller() == buyer, "only the future owner can buy the shares");
+    require(offers.contains(offer), "unknown offer");
+    require(offer.isOngoing(), "the sale offer is not ongoing anymore");
+    require(offer.cost.compareTo(amount) <= 0, "not enough money");
+    offers.remove(offer);
+    removeShares(offer.seller, offer.sharesOnSale);
+    addShares(buyer, offer.sharesOnSale);
+    offer.seller.receive(offer.cost);
+  }
+
+  private BigInteger sharesOnSaleOf(S shareholder) {
+    return offers.stream()
+      .filter(offer -> offer.seller == shareholder && offer.isOngoing())
+      .map(offer -> offer.sharesOnSale)
+      .reduce(ZERO, BigInteger::add);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{A simplified part of our implementation of the shared entity interface.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}.}\label{fig:simple_shared_entity}
+\end{figure}
+
+Fig.~\ref{fig:simple_shared_entity} shows our implementation of the \<SharedEntity> interface
+in Fig.~\ref{fig:shared_entity}
+by using two fields: \<shares> maps each shareholder to its amount of shares and
+\<offers> collects the offers that have been placed.
+The constructor populates the map \<shares>, initially.
+Method \<sharesOf> simply accesses
+\<shares>, by using zero as default. Method \<place> requires its \<caller()> to be
+the seller identified in the \<offer>. This forbids shareholders to sell shares on behalf of others.
+Moreover, this guarantees that the caller has type \<S>, like \<offer.seller>.
+As we said before, this cannot be expressed with the syntax of the language.
+Method \<place> further requires the seller to be a shareholder with at least \<offer.sharesOnSale>
+shares not yet placed on sale. This forbids to oversell more shares
+than one owns. At the end, \<place> adds the \<offer> to the set of \<offers>.
+Method \<accept> requires that who calls the method must be \<buyer>. Hence, successful
+calls to \<accept> can only pass the same caller for \<buyer>. This is a trick to enforce the
+caller to have type \<S>, since the syntax of the language does not allow one to express it,
+as we explained before. Then \<accept> requires the \<offer> to exists, to be still ongoing
+and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,
+the \<offer> is removed from the \<offers>, shares are moved from seller to buyer (code not
+shown in Fig.~\ref{fig:simple_shared_entity}) and the seller of the \<offer>
+receives the required price \<offer.cost>.
+
+Let us state an important property about this code:
+
+\begin{tcolorbox}
+  \begin{center}\emph{Consistency of Shareholders}\end{center}
+  If \<sse> is a {\codesize\texttt{SimpleSharedEntity<S,O>}}
+  then the domain of its map \<shares> consists of objects of type \<S> only.
+\end{tcolorbox}
+
+\noindent
+It turns out that this property holds.
+Namely, the code does not use unchecked casts, hence it is strongly-typed and
+the map \<shares> actually holds values of type \<S> in its domain, only.
+For this consistency result, we had to pay the price
+of the dummy \<buyer> argument for method \<accept>. Without that argument,
+\emph{Consistency of Shareholders} would not hold, since we could only write
+\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
+Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
+not strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
+and become shareholders.
+
+There is, however, a problem with the previous paragraph. Namely, absence of unchecked
+operations guarantees strong typing \emph{of the source code}. But what is installed
+and executed in blockchain is the Java bytecode from Fig.~\ref{fig:simple_shared_entity}.
+It is important to investigate whether it is also strongly typed or not.
 

### 2021-06-22 generics.tex
Reorganized the consistency of shareholders property

@@
 
+  List<S> getShareholders();
+
   class Offer<S extends PayableContract> extends Storage {
@@
 Implementations can subclass \<Offer> if they need more specific offers.
-By using class \<Offer>, the \<SharedEntity> interface specifies three methods.
+By using class \<Offer>, the \<SharedEntity> interface specifies four methods.
 Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
@@
 payments in Takamaka.
+Method \<getShareholders> yields the list of the current shareholders of the entity.
+It is not \<@View>, since it creates a new list.
 
@@
 Since the syntax of the language does not support our abstraction, we will have to
-program explicit dynamic checks in code, as shown in a moment.
+program explicit dynamic checks in code, as shown later.
+
+Let us state an important property about shared entities:
+
+\begin{tcolorbox}
+  \begin{center}\emph{Consistency of Shareholders}\end{center}
+  If \<sse> is a {\codesize\texttt{SharedEntity<S,O>}}
+  then the elements in the list \<sse.getShareholders()> have type \<S>.
+\end{tcolorbox}
+
+\noindent
+This property is important since it states that we can trust the type \<S> of
+the shareholders: if we create a \<SharedEntity> and fix a specific type \<S>
+for its shareholders, then only instances of \<S> will manage to become shareholders.
 
@@
   }
+
+  @Override
+  public List<S> getShareholders() {
+    return shares.keyList();
+  }
 }
@@
 receives the required price \<offer.cost>.
+Method \<getShareholders()> yields the list of the keys in the domain of map \<shares>.
 
-Let us state an important property about this code:
-
-\begin{tcolorbox}
-  \begin{center}\emph{Consistency of Shareholders}\end{center}
-  If \<sse> is a {\codesize\texttt{SimpleSharedEntity<S,O>}}
-  then the domain of its map \<shares> consists of objects of type \<S> only.
-\end{tcolorbox}
-
-\noindent
-It turns out that this property holds.
-Namely, the code does not use unchecked casts, hence it is strongly-typed and
+It turns out that the \emph{Consistency of Shareholders} property holds if Java code
+creates and populates \<SimpleSharedEntity>s.
+Namely, the code in Fig.~\ref{fig:simple_shared_entity}
+does not use unchecked casts, hence it is strongly-typed and
 the map \<shares> actually holds values of type \<S> in its domain, only.
@@
 Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-not strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
+non-strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
 and become shareholders.
 
-There is, however, a problem with the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing \emph{of the source code}. But what is installed
-and executed in blockchain is the Java bytecode from Fig.~\ref{fig:simple_shared_entity}.
-It is important to investigate whether it is also strongly typed or not.
+There is, however, a problem with the reasoning
+in the previous paragraph. Namely, absence of unchecked
+operations guarantees strong typing \emph{of source code} in Java. But what is installed
+and executed in blockchain is the Java bytecode derived from
+the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
+Malicious users might install in blockchain some crafted bytecode that
+calls the methods of \<SimpleSharedEntity>s in order to attack
+the contract. In particular,
+the signature of method \<accept> declares a parameter \<buyer> of type \<S> at source code level, but
+its compilation into Java bytecode declares a parameter \<buyer> of type \<PayableContract> instead.
+This is because of the so called \emph{erasure} of generic types in the compilation of Java.
+It entails that an attacker might install in blockchain a snippet of bytecode that calls
+\<accept> and passes \emph{any} \<PayableContract>, not only those that are instances of \<S>:
+the \emph{Consistency of Shareholders} property is easily violated at bytecode level.
+
+In general, the problem arises whenever a method declares a formal parameter of generic type,
+such as \<buyer> in method \<accept> in Fig.~\ref{fig:shared_entity}. Next section shows
+the actual significance of the consequent security risk.
 

### 2021-06-22 generics.tex
Started explaining the attack on the validators set

@@
 
+This paper originated from an actual security issue that we found in our code
+that used shared entities in order to model the validators of a blockchain.
+Namely, the Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
+generic engine for replicating an application over a network of nodes. In our case,
+the application is the executor of smart contracts in Java, such as that in
+Fig.~\ref{fig:simple_shared_entity}. Tendermint is based on a proof of stake
+consensus, which means that a selected dynamic subset of the nodes is in charge of
+validating the transactions and voting their acceptance. Validator nodes are
+represented by \<Validator> objects, that are externally owned accounts
+with an extra identifier derived from their public key (see Fig.~\ref{fig:validator}).
+This identifier is public information, reported in the blocks or easily eavesdropped.
+At each block, the reward for the block is distributed to the validators.
+Tendermint applications can implement their own
+policy for changing the validator set dynamically. In our case, the validator set
+and the distribution of the reward is implemented by a subclass
+\<Validators> of \<SimpleSharedEntity> whose type \<S> for the shareholders
+has been set to \<Validator>. Shares are voting power in this case.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public final class Validator extends ExternallyOwnedAccount {
+  private final String id;
+
+  public Validator(String publicKey) {
+    super(publicKey);
+    this.id = /* derived from publicKey as specified in Tendermint's spec */
+  }
+
+  public @View String id() {
+    return id;
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{The representation of a validator of a Tendermint blockchain.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintED25519Validator.java}.}\label{fig:validator}
+\end{figure}
+
+Since shares of a \<SimpleSharedEntity> can be sold and bought, the set of validators
+is dynamic. Users can sell or buy voting power in order to invest in the blockchain
+and earn rewards. At each block creation, Hotmoka calls method \<getShareholders>
+on the shared entity and informs the
+underlying Tendermint engine about the identifiers of the validator nodes for the next blocks.
+Tendermint expects such validators to mine and vote the subsequent blocks, until a change in the
+validators set occurs.
+
+It is important that the shareholders be instances of \<Validator>, since that class enforces the
+match between their public key, that identifies who can use the reward sent to the validator,
+and the Tendermint identifier of the validator, that identifies which node of the blockchain
+must do the validation work (Fig.~\ref{fig:validator}).
+If it were possible to add a shareholder of another
+type \<Attacker>, the latter could declare the identifier of a node that does not correspond to its
+public key (see Fig.~\ref{fig:attacker}):
+the node (belonging to the \emph{victim}) would do the work while the owner
+of the private key of the \<Attacker> could just wait and earn the reward.
+A sort of validator's slavery.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class Attacker extends ExternallyOwnedAccount {
+
+  public Validator(String publicKey) {
+    super(publicKey);
+  }
+
+  public @View String id() {
+    return /* id of the victim node of the blockchain */
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the reward of its work.}\label{fig:attacker}
+\end{figure}
+
 \section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}

### 2021-06-28 010_introduction.tex
struttura file latex

--- 2021-06-08 generics.tex
@@
-\documentclass{llncs}
-
-\usepackage{url}
-\usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage[hidelinks]{hyperref}
-\usepackage{pdfpages}
-\usepackage{relsize}
-\usepackage{tcolorbox}
-
-\usepackage{svg}
-\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\includesvg[height = 2ex]{svg-inkscape/ORCID_iD}}}
-
-\newcommand{\ie}{\textit{, ie.\ }}
-
-\def\codesize{\smaller}
-\def\<#1>{\codeid{#1}}
-\newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}
-
-\usepackage{listings, xcolor}
-
-\renewcommand{\UrlFont}{\ttfamily\codesize}
-
-\definecolor{verylightgray}{rgb}{.97,.97,.97}
-
-\lstdefinelanguage{Takamaka}{
-        keywords=[1]{abstract, break, case, catch, class, continue, default, do
-, else, false, finally, for, if, final, implements, extends, import, instanceof, interface, length, new, private, protected, public, return, super, switch, this, throw, true, try, while, var, null}, % generic keywords
-        keywordstyle=[1]\color{blue}\bfseries,
-        keywords=[2]{boolean, int, long, float, double, byte, short, char, void, enum}, % types; money and time units
-        keywordstyle=[2]\color{teal}\bfseries,
-        keywords=[3]{@Override,@View,@FromContract,@Payable}, % annotations
-        keywordstyle=[3]\color{violet}\bfseries,
-        identifierstyle=\color{black},
-        sensitive=false,
-        comment=[l]{//},
-        morecomment=[s]{/*}{*/},
-        commentstyle=\color{gray}\ttfamily,
-        stringstyle=\color{red}\ttfamily,
-        morestring=[b]',
-        morestring=[b]"
-}
-
-\lstset{
-        language=Takamaka,
-        backgroundcolor=\color{verylightgray},
-        extendedchars=true,
-        basicstyle=\scriptsize\ttfamily,
-        showstringspaces=false,
-        showspaces=false,
-        numbers=none,
-        numberstyle=\scriptsize,
-        numbersep=9pt,
-        tabsize=2,
-        breaklines=true,
-        showtabs=false,
-        captionpos=b
-}
-
-\begin{document}
-
-\title{Power and Pitfalls of Generic Smart Contracts}
-\titlerunning{Power and Pitfalls of Generics for Smart Contracts}
-\author{Andrea Benini \and Sara Migliorini \and Fausto Spoto}
-\institute{Dipartimento di Informatica, Universit\`a di Verona, Italy}
-
-\maketitle
-
-\begin{abstract}
-  TBD
-  \keywords{smart contract \and generics \and blockchain}
-\end{abstract}
-
 \section{Introduction}\label{sec:introduction}
@@
 Sec.~\ref{sec:conclusion} concludes.
-
-\section{Shared Entities using Generics}\label{sec:shared_entities}
-
-A \emph{shared entity} is something divided into \emph{shares}. Participants,
-that hold shares, are called \emph{shareholders} and can dynamically
-sell and buy shares. An example of a shared entity is a comporation,
-where shares represent units of possess of the company. Another example is
-a voting community, where shares represent the voting power of each given voter.
-A further example are the validator nodes of a proof of stake blockchain,
-where shares represent their voting power and remuneration percentage.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract,O extends Offer<S>> {
-
-  @View
-  BigInteger sharesOf(S shareholder);
-
-  @FromContract(PayableContract.class) @Payable
-  void place(BigInteger amount, O offer);
-
-  @FromContract(PayableContract.class) @Payable
-  void accept(BigInteger amount, S buyer, O offer);
-
-  class Offer<S extends PayableContract> extends Storage {
-    public final S seller;
-    public final BigInteger sharesOnSale;
-    public final BigInteger cost;
-    public final long expiration;
-
-    public Offer
-        (S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
-
-      this.seller = seller;
-      this.sharesOnSale = sharesOnSale;
-      this.cost = cost;
-      this.expiration = now() + duration;
-    }
-
-    @View
-    public boolean isOngoing() {
-      return now() <= expiration;
-    }
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
-\end{figure}
-
-In general, two concepts are specific to each implementation of shared entities:
-who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
-Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
-It includes an inner class \<Offer> that models sale offers:
-it specifies who is the seller of the shares,
-how many shares are being sold, the requested price and the expiration of the offer.
-Method \<isOngoing> checks if an offer has not expired yet.
-Implementations can subclass \<Offer> if they need more specific offers.
-By using class \<Offer>, the \<SharedEntity> interface specifies three methods.
-Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
-It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
-\emph{for free}, without paying gas, since it has no side-effects. Who wants to sell
-(some of) its shares calls method \<place> with a sale \<offer>.
-This method is annotated as \<@Payable> since
-implementations are allowed to require a payment of \<amount> coins for managing the sale.
-Who buys the shares calls method \<accept> with the accepted \<offer>
-and with itself as \<buyer> (the reason will be explained soon)
-and becomes a new shareholder or increases
-its cumulative number of shares (if it was a shareholder already).
-Also this method is \<@Payable>, since its caller must pay \<amount> $\ge$ \<offer.cost>
-coins to the seller.
-This means that shareholders must be able to receive payments and that
-is why \<S extends PayableContract>: the \<PayableContract>s are those that can receive
-payments in Takamaka.
-
-The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This attests, explicity, their will
-to sell or buy shares. The callers must be (old or new) shareholders,
-hence of type \<S>. In Takamaka, this could be written
-as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Because of this limitation,
-the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
-which allows \emph{any} \<PayableContract> to call these methods, not just those of type \<S>.
-Since the syntax of the language does not support our abstraction, we will have to
-program explicit dynamic checks in code, as shown in a moment.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class SimpleSharedEntity
-      <S extends PayableContract,O extends Offer<S>>
-      extends Contract
-      implements SharedEntity<S,O> {
-
-  private final StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private final StorageSet<O> offers = new StorageTreeSet<>();
-
-  public SimpleSharedEntity(S[] shareholders, BigInteger[] shares) {
-    require(shareholders.length == shares.length, "length mismatch");
-    for (int pos = 0; pos < shareholders.length; pos++)
-      addShares(shareholders[pos], shares[pos]);
-  }
-
-  @Override @View
-  public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, BigInteger.ZERO);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "not authorized to sell");
-    require(shares.containsKey(offer.seller), "only shareholders can sell");
-    require(sharesOf(offer.seller).subtract(sharesOnSaleOf(offer.seller))
-        .compareTo(offer.sharesOnSale) >= 0, "not enough shares to sell");
-    offers.add(offer);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "only the future owner can buy the shares");
-    require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the sale offer is not ongoing anymore");
-    require(offer.cost.compareTo(amount) <= 0, "not enough money");
-    offers.remove(offer);
-    removeShares(offer.seller, offer.sharesOnSale);
-    addShares(buyer, offer.sharesOnSale);
-    offer.seller.receive(offer.cost);
-  }
-
-  private BigInteger sharesOnSaleOf(S shareholder) {
-    return offers.stream()
-      .filter(offer -> offer.seller == shareholder && offer.isOngoing())
-      .map(offer -> offer.sharesOnSale)
-      .reduce(ZERO, BigInteger::add);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}.}\label{fig:simple_shared_entity}
-\end{figure}
-
-Fig.~\ref{fig:simple_shared_entity} shows our implementation of the \<SharedEntity> interface
-in Fig.~\ref{fig:shared_entity}
-by using two fields: \<shares> maps each shareholder to its amount of shares and
-\<offers> collects the offers that have been placed.
-The constructor populates the map \<shares>, initially.
-Method \<sharesOf> simply accesses
-\<shares>, by using zero as default. Method \<place> requires its \<caller()> to be
-the seller identified in the \<offer>. This forbids shareholders to sell shares on behalf of others.
-Moreover, this guarantees that the caller has type \<S>, like \<offer.seller>.
-As we said before, this cannot be expressed with the syntax of the language.
-Method \<place> further requires the seller to be a shareholder with at least \<offer.sharesOnSale>
-shares not yet placed on sale. This forbids to oversell more shares
-than one owns. At the end, \<place> adds the \<offer> to the set of \<offers>.
-Method \<accept> requires that who calls the method must be \<buyer>. Hence, successful
-calls to \<accept> can only pass the same caller for \<buyer>. This is a trick to enforce the
-caller to have type \<S>, since the syntax of the language does not allow one to express it,
-as we explained before. Then \<accept> requires the \<offer> to exists, to be still ongoing
-and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,
-the \<offer> is removed from the \<offers>, shares are moved from seller to buyer (code not
-shown in Fig.~\ref{fig:simple_shared_entity}) and the seller of the \<offer>
-receives the required price \<offer.cost>.
-
-Let us state an important property about this code:
-
-\begin{tcolorbox}
-  \begin{center}\emph{Consistency of Shareholders}\end{center}
-  If \<sse> is a {\codesize\texttt{SimpleSharedEntity<S,O>}}
-  then the domain of its map \<shares> consists of objects of type \<S> only.
-\end{tcolorbox}
-
-\noindent
-It turns out that this property holds.
-Namely, the code does not use unchecked casts, hence it is strongly-typed and
-the map \<shares> actually holds values of type \<S> in its domain, only.
-For this consistency result, we had to pay the price
-of the dummy \<buyer> argument for method \<accept>. Without that argument,
-\emph{Consistency of Shareholders} would not hold, since we could only write
-\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
-Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-not strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
-and become shareholders.
-
-There is, however, a problem with the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing \emph{of the source code}. But what is installed
-and executed in blockchain is the Java bytecode from Fig.~\ref{fig:simple_shared_entity}.
-It is important to investigate whether it is also strongly typed or not.
-
-\section{An Attack to the Shared Entities Contract}\label{sec:attack}
-
-\section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}
-
-\section{Conclusion}\label{sec:conclusion}
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2021-06-28 020_shared_entities_with_generics.tex
struttura file latex

--- 2021-06-22 generics.tex
@@
-\documentclass{llncs}
-
-\usepackage{url}
-\usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage[hidelinks]{hyperref}
-\usepackage{pdfpages}
-\usepackage{relsize}
-\usepackage{tcolorbox}
-
-\usepackage{svg}
-\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\includesvg[height = 2ex]{svg-inkscape/ORCID_iD}}}
-
-\newcommand{\ie}{\textit{, ie.\ }}
-
-\def\codesize{\smaller}
-\def\<#1>{\codeid{#1}}
-\newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}
-
-\usepackage{listings, xcolor}
-
-\renewcommand{\UrlFont}{\ttfamily\codesize}
-
-\definecolor{verylightgray}{rgb}{.97,.97,.97}
-
-\lstdefinelanguage{Takamaka}{
-        keywords=[1]{abstract, break, case, catch, class, continue, default, do
-, else, false, finally, for, if, final, implements, extends, import, instanceof, interface, length, new, private, protected, public, return, super, switch, this, throw, true, try, while, var, null}, % generic keywords
-        keywordstyle=[1]\color{blue}\bfseries,
-        keywords=[2]{boolean, int, long, float, double, byte, short, char, void, enum}, % types; money and time units
-        keywordstyle=[2]\color{teal}\bfseries,
-        keywords=[3]{@Override,@View,@FromContract,@Payable}, % annotations
-        keywordstyle=[3]\color{violet}\bfseries,
-        identifierstyle=\color{black},
-        sensitive=false,
-        comment=[l]{//},
-        morecomment=[s]{/*}{*/},
-        commentstyle=\color{gray}\ttfamily,
-        stringstyle=\color{red}\ttfamily,
-        morestring=[b]',
-        morestring=[b]"
-}
-
-\lstset{
-        language=Takamaka,
-        backgroundcolor=\color{verylightgray},
-        extendedchars=true,
-        basicstyle=\scriptsize\ttfamily,
-        showstringspaces=false,
-        showspaces=false,
-        numbers=none,
-        numberstyle=\scriptsize,
-        numbersep=9pt,
-        tabsize=2,
-        breaklines=true,
-        showtabs=false,
-        captionpos=b
-}
-
-\begin{document}
-
-\title{Power and Pitfalls of Generic Smart Contracts}
-\titlerunning{Power and Pitfalls of Generics for Smart Contracts}
-\author{Andrea Benini \and Sara Migliorini \and Fausto Spoto}
-\institute{Dipartimento di Informatica, Universit\`a di Verona, Italy}
-
-\maketitle
-
-\begin{abstract}
-  TBD
-  \keywords{smart contract \and generics \and blockchain}
-\end{abstract}
-
-\section{Introduction}\label{sec:introduction}
-
-Blockchains exploit the redundant, concurrent execution of the same
-transactions on a decentralized network of many machines
-in order to enforce their execution in accordance to
-a set of predefined rules. Namely, blockchains make it hard, for a single machine,
-to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
-gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
-has been the first blockchain's success story. Here
-transactions are programmed in a non-Turing complete bytecode language,
-almost exclusively used to implement transfers of units of coins between \emph{accounts}.
-
-A few years after Bitcoin, another blockchain, called
-Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
-transactions in an actual, imperative and Turing-complete programming language, called Solidity.
-Solidity's code is organized in \emph{smart contracts}, that can be seen as
-objects that control money.
-Ethereum's transactions can hence execute much more than coin transfers. Namely,
-they run object constructors and methods, which results in a sort
-of \emph{world computer} that persists the same objects in the memory of all the
-computers in the blockchain's network.
-
-In Solidity's bytecode,
-non-primitive values are referenced through a very general
-\<address> type. For instance, a Solidity method
-\<child(Person p, uint256 n) returns Person> actually compiles
-into \<child(address p, uint256 n) returns address>, losing most
-type information~\cite{CrafaPZ19}.
-At run time, it is the bytecode that gets executed. Hence,
-everything can be passed for \<p>, not just a \<Person>.
-The compiler cannot even enforce strong typing
-by generating defensive type instance checks and casts, since
-values are unboxed in Ethereum: they have no attached
-type information at run time,
-they are just numerical \emph{addresses}.
-It follows that calls to \<Person>'s methods
-on \<p> might actually execute arbitrary code, if \<p> is not a \<Person>.
-In other words, Solidity is not strongly typed.
-Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
-to our methods, such as on \<p> passed to \<child>, since an attacker can pass crafted
-objects for \<p>, with arbitrary implementations for their methods,
-which can result in the unexpected execution of
-dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
-costed millions of dollars.
-
-Strong typing is one of the reasons that push towards the adoption
-of \emph{traditional} programming languages for smart contracts. For instance,
-the Cosmos blockchain~\cite{cosmos} uses Go. The
-Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
-for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
-Hyperledger~\cite{hyperldeger} allows Go and Java.
-Another reason is the availability of modern
-language features, missing in Solidity,
-such as \emph{generics}\ie the possibility of using
-type variables. Generics are powerful and very useful for programming
-smart contracts. In Java source code, they are strongly typed, if no \emph{unchecked operations}
-are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics in Java are compiled by \emph{erasure}\ie replaced
-by their upwards bound, most often \<Object>. Hence, the risk
-is that generics introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity.
-
-The contribution of this paper is to show a real-life
-use of generics for an actual smart contract used in the support
-library of the Takamaka language, and to show that a naive use
-of Java generics can lead to a code security vulnerability that
-allows an attacker to earn money by exploiting someone else's work.
-This paper will provide a fix to that specific issue,
-by obliging the compiler to generate defensive checks.
-More generally, this paper can be useful for the definition of
-bytecode languages for future smart contract languages, by
-learning from the weaknesses of Java bytecode.
-
-The rest of this paper is organized as follows.
-Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contracts that uses generics. Sec.~\ref{sec:attack} shows that a naive
-deployment of that contract leads to a code vulnerability.
-Sec.~\ref{sec:fix} shows a fix to that vulnerability.
-Sec.~\ref{sec:conclusion} concludes.
-
 \section{Shared Entities using Generics}\label{sec:shared_entities}
@@
 that hold shares, are called \emph{shareholders} and can dynamically
-sell and buy shares. An example of a shared entity is a comporation,
+sell and buy shares. An example of a shared entity is a corporation,
 where shares represent units of possess of the company. Another example is
@@
 the actual significance of the consequent security risk.
-
-\section{An Attack to the Shared Entities Contract}\label{sec:attack}
-
-\section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}
-
-\section{Conclusion}\label{sec:conclusion}
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2021-06-28 030_generics_attack.tex
struttura file latex

--- 2021-06-22 generics.tex
@@
-\documentclass{llncs}
-
-\usepackage{url}
-\usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage[hidelinks]{hyperref}
-\usepackage{pdfpages}
-\usepackage{relsize}
-\usepackage{tcolorbox}
-
-\usepackage{svg}
-\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\includesvg[height = 2ex]{svg-inkscape/ORCID_iD}}}
-
-\newcommand{\ie}{\textit{, ie.\ }}
-
-\def\codesize{\smaller}
-\def\<#1>{\codeid{#1}}
-\newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}
-
-\usepackage{listings, xcolor}
-
-\renewcommand{\UrlFont}{\ttfamily\codesize}
-
-\definecolor{verylightgray}{rgb}{.97,.97,.97}
-
-\lstdefinelanguage{Takamaka}{
-        keywords=[1]{abstract, break, case, catch, class, continue, default, do
-, else, false, finally, for, if, final, implements, extends, import, instanceof, interface, length, new, private, protected, public, return, super, switch, this, throw, true, try, while, var, null}, % generic keywords
-        keywordstyle=[1]\color{blue}\bfseries,
-        keywords=[2]{boolean, int, long, float, double, byte, short, char, void, enum}, % types; money and time units
-        keywordstyle=[2]\color{teal}\bfseries,
-        keywords=[3]{@Override,@View,@FromContract,@Payable}, % annotations
-        keywordstyle=[3]\color{violet}\bfseries,
-        identifierstyle=\color{black},
-        sensitive=false,
-        comment=[l]{//},
-        morecomment=[s]{/*}{*/},
-        commentstyle=\color{gray}\ttfamily,
-        stringstyle=\color{red}\ttfamily,
-        morestring=[b]',
-        morestring=[b]"
-}
-
-\lstset{
-        language=Takamaka,
-        backgroundcolor=\color{verylightgray},
-        extendedchars=true,
-        basicstyle=\scriptsize\ttfamily,
-        showstringspaces=false,
-        showspaces=false,
-        numbers=none,
-        numberstyle=\scriptsize,
-        numbersep=9pt,
-        tabsize=2,
-        breaklines=true,
-        showtabs=false,
-        captionpos=b
-}
-
-\begin{document}
-
-\title{Power and Pitfalls of Generic Smart Contracts}
-\titlerunning{Power and Pitfalls of Generics for Smart Contracts}
-\author{Andrea Benini \and Sara Migliorini \and Fausto Spoto}
-\institute{Dipartimento di Informatica, Universit\`a di Verona, Italy}
-
-\maketitle
-
-\begin{abstract}
-  TBD
-  \keywords{smart contract \and generics \and blockchain}
-\end{abstract}
-
-\section{Introduction}\label{sec:introduction}
-
-Blockchains exploit the redundant, concurrent execution of the same
-transactions on a decentralized network of many machines
-in order to enforce their execution in accordance to
-a set of predefined rules. Namely, blockchains make it hard, for a single machine,
-to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
-gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
-has been the first blockchain's success story. Here
-transactions are programmed in a non-Turing complete bytecode language,
-almost exclusively used to implement transfers of units of coins between \emph{accounts}.
-
-A few years after Bitcoin, another blockchain, called
-Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
-transactions in an actual, imperative and Turing-complete programming language, called Solidity.
-Solidity's code is organized in \emph{smart contracts}, that can be seen as
-objects that control money.
-Ethereum's transactions can hence execute much more than coin transfers. Namely,
-they run object constructors and methods, which results in a sort
-of \emph{world computer} that persists the same objects in the memory of all the
-computers in the blockchain's network.
-
-In Solidity's bytecode,
-non-primitive values are referenced through a very general
-\<address> type. For instance, a Solidity method
-\<child(Person p, uint256 n) returns Person> actually compiles
-into \<child(address p, uint256 n) returns address>, losing most
-type information~\cite{CrafaPZ19}.
-At run time, it is the bytecode that gets executed. Hence,
-everything can be passed for \<p>, not just a \<Person>.
-The compiler cannot even enforce strong typing
-by generating defensive type instance checks and casts, since
-values are unboxed in Ethereum: they have no attached
-type information at run time,
-they are just numerical \emph{addresses}.
-It follows that calls to \<Person>'s methods
-on \<p> might actually execute arbitrary code, if \<p> is not a \<Person>.
-In other words, Solidity is not strongly typed.
-Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
-to our methods, such as on \<p> passed to \<child>, since an attacker can pass crafted
-objects for \<p>, with arbitrary implementations for their methods,
-which can result in the unexpected execution of
-dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
-costed millions of dollars.
-
-Strong typing is one of the reasons that push towards the adoption
-of \emph{traditional} programming languages for smart contracts. For instance,
-the Cosmos blockchain~\cite{cosmos} uses Go. The
-Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
-for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
-Hyperledger~\cite{hyperldeger} allows Go and Java.
-Another reason is the availability of modern
-language features, missing in Solidity,
-such as \emph{generics}\ie the possibility of using
-type variables. Generics are powerful and very useful for programming
-smart contracts. In Java source code, they are strongly typed, if no \emph{unchecked operations}
-are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics in Java are compiled by \emph{erasure}\ie replaced
-by their upwards bound, most often \<Object>. Hence, the risk
-is that generics introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity.
-
-The contribution of this paper is to show a real-life
-use of generics for an actual smart contract used in the support
-library of the Takamaka language, and to show that a naive use
-of Java generics can lead to a code security vulnerability that
-allows an attacker to earn money by exploiting someone else's work.
-This paper will provide a fix to that specific issue,
-by obliging the compiler to generate defensive checks.
-More generally, this paper can be useful for the definition of
-bytecode languages for future smart contract languages, by
-learning from the weaknesses of Java bytecode.
-
-The rest of this paper is organized as follows.
-Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contracts that uses generics. Sec.~\ref{sec:attack} shows that a naive
-deployment of that contract leads to a code vulnerability.
-Sec.~\ref{sec:fix} shows a fix to that vulnerability.
-Sec.~\ref{sec:conclusion} concludes.
-
-\section{Shared Entities using Generics}\label{sec:shared_entities}
-
-A \emph{shared entity} is something divided into \emph{shares}. Participants,
-that hold shares, are called \emph{shareholders} and can dynamically
-sell and buy shares. An example of a shared entity is a comporation,
-where shares represent units of possess of the company. Another example is
-a voting community, where shares represent the voting power of each given voter.
-A further example are the validator nodes of a proof of stake blockchain,
-where shares represent their voting power and remuneration percentage.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract,O extends Offer<S>> {
-
-  @View
-  BigInteger sharesOf(S shareholder);
-
-  @FromContract(PayableContract.class) @Payable
-  void place(BigInteger amount, O offer);
-
-  @FromContract(PayableContract.class) @Payable
-  void accept(BigInteger amount, S buyer, O offer);
-
-  List<S> getShareholders();
-
-  class Offer<S extends PayableContract> extends Storage {
-    public final S seller;
-    public final BigInteger sharesOnSale;
-    public final BigInteger cost;
-    public final long expiration;
-
-    public Offer
-        (S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
-
-      this.seller = seller;
-      this.sharesOnSale = sharesOnSale;
-      this.cost = cost;
-      this.expiration = now() + duration;
-    }
-
-    @View
-    public boolean isOngoing() {
-      return now() <= expiration;
-    }
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
-\end{figure}
-
-In general, two concepts are specific to each implementation of shared entities:
-who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
-Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
-It includes an inner class \<Offer> that models sale offers:
-it specifies who is the seller of the shares,
-how many shares are being sold, the requested price and the expiration of the offer.
-Method \<isOngoing> checks if an offer has not expired yet.
-Implementations can subclass \<Offer> if they need more specific offers.
-By using class \<Offer>, the \<SharedEntity> interface specifies four methods.
-Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
-It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
-\emph{for free}, without paying gas, since it has no side-effects. Who wants to sell
-(some of) its shares calls method \<place> with a sale \<offer>.
-This method is annotated as \<@Payable> since
-implementations are allowed to require a payment of \<amount> coins for managing the sale.
-Who buys the shares calls method \<accept> with the accepted \<offer>
-and with itself as \<buyer> (the reason will be explained soon)
-and becomes a new shareholder or increases
-its cumulative number of shares (if it was a shareholder already).
-Also this method is \<@Payable>, since its caller must pay \<amount> $\ge$ \<offer.cost>
-coins to the seller.
-This means that shareholders must be able to receive payments and that
-is why \<S extends PayableContract>: the \<PayableContract>s are those that can receive
-payments in Takamaka.
-Method \<getShareholders> yields the list of the current shareholders of the entity.
-It is not \<@View>, since it creates a new list.
-
-The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This attests, explicity, their will
-to sell or buy shares. The callers must be (old or new) shareholders,
-hence of type \<S>. In Takamaka, this could be written
-as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Because of this limitation,
-the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
-which allows \emph{any} \<PayableContract> to call these methods, not just those of type \<S>.
-Since the syntax of the language does not support our abstraction, we will have to
-program explicit dynamic checks in code, as shown later.
-
-Let us state an important property about shared entities:
-
-\begin{tcolorbox}
-  \begin{center}\emph{Consistency of Shareholders}\end{center}
-  If \<sse> is a {\codesize\texttt{SharedEntity<S,O>}}
-  then the elements in the list \<sse.getShareholders()> have type \<S>.
-\end{tcolorbox}
-
-\noindent
-This property is important since it states that we can trust the type \<S> of
-the shareholders: if we create a \<SharedEntity> and fix a specific type \<S>
-for its shareholders, then only instances of \<S> will manage to become shareholders.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class SimpleSharedEntity
-      <S extends PayableContract,O extends Offer<S>>
-      extends Contract
-      implements SharedEntity<S,O> {
-
-  private final StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private final StorageSet<O> offers = new StorageTreeSet<>();
-
-  public SimpleSharedEntity(S[] shareholders, BigInteger[] shares) {
-    require(shareholders.length == shares.length, "length mismatch");
-    for (int pos = 0; pos < shareholders.length; pos++)
-      addShares(shareholders[pos], shares[pos]);
-  }
-
-  @Override @View
-  public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, BigInteger.ZERO);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "not authorized to sell");
-    require(shares.containsKey(offer.seller), "only shareholders can sell");
-    require(sharesOf(offer.seller).subtract(sharesOnSaleOf(offer.seller))
-        .compareTo(offer.sharesOnSale) >= 0, "not enough shares to sell");
-    offers.add(offer);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "only the future owner can buy the shares");
-    require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the sale offer is not ongoing anymore");
-    require(offer.cost.compareTo(amount) <= 0, "not enough money");
-    offers.remove(offer);
-    removeShares(offer.seller, offer.sharesOnSale);
-    addShares(buyer, offer.sharesOnSale);
-    offer.seller.receive(offer.cost);
-  }
-
-  private BigInteger sharesOnSaleOf(S shareholder) {
-    return offers.stream()
-      .filter(offer -> offer.seller == shareholder && offer.isOngoing())
-      .map(offer -> offer.sharesOnSale)
-      .reduce(ZERO, BigInteger::add);
-  }
-
-  @Override
-  public List<S> getShareholders() {
-    return shares.keyList();
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}.}\label{fig:simple_shared_entity}
-\end{figure}
-
-Fig.~\ref{fig:simple_shared_entity} shows our implementation of the \<SharedEntity> interface
-in Fig.~\ref{fig:shared_entity}
-by using two fields: \<shares> maps each shareholder to its amount of shares and
-\<offers> collects the offers that have been placed.
-The constructor populates the map \<shares>, initially.
-Method \<sharesOf> simply accesses
-\<shares>, by using zero as default. Method \<place> requires its \<caller()> to be
-the seller identified in the \<offer>. This forbids shareholders to sell shares on behalf of others.
-Moreover, this guarantees that the caller has type \<S>, like \<offer.seller>.
-As we said before, this cannot be expressed with the syntax of the language.
-Method \<place> further requires the seller to be a shareholder with at least \<offer.sharesOnSale>
-shares not yet placed on sale. This forbids to oversell more shares
-than one owns. At the end, \<place> adds the \<offer> to the set of \<offers>.
-Method \<accept> requires that who calls the method must be \<buyer>. Hence, successful
-calls to \<accept> can only pass the same caller for \<buyer>. This is a trick to enforce the
-caller to have type \<S>, since the syntax of the language does not allow one to express it,
-as we explained before. Then \<accept> requires the \<offer> to exists, to be still ongoing
-and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,
-the \<offer> is removed from the \<offers>, shares are moved from seller to buyer (code not
-shown in Fig.~\ref{fig:simple_shared_entity}) and the seller of the \<offer>
-receives the required price \<offer.cost>.
-Method \<getShareholders()> yields the list of the keys in the domain of map \<shares>.
-
-It turns out that the \emph{Consistency of Shareholders} property holds if Java code
-creates and populates \<SimpleSharedEntity>s.
-Namely, the code in Fig.~\ref{fig:simple_shared_entity}
-does not use unchecked casts, hence it is strongly-typed and
-the map \<shares> actually holds values of type \<S> in its domain, only.
-For this consistency result, we had to pay the price
-of the dummy \<buyer> argument for method \<accept>. Without that argument,
-\emph{Consistency of Shareholders} would not hold, since we could only write
-\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
-Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-non-strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
-and become shareholders.
-
-There is, however, a problem with the reasoning
-in the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing \emph{of source code} in Java. But what is installed
-and executed in blockchain is the Java bytecode derived from
-the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
-Malicious users might install in blockchain some crafted bytecode that
-calls the methods of \<SimpleSharedEntity>s in order to attack
-the contract. In particular,
-the signature of method \<accept> declares a parameter \<buyer> of type \<S> at source code level, but
-its compilation into Java bytecode declares a parameter \<buyer> of type \<PayableContract> instead.
-This is because of the so called \emph{erasure} of generic types in the compilation of Java.
-It entails that an attacker might install in blockchain a snippet of bytecode that calls
-\<accept> and passes \emph{any} \<PayableContract>, not only those that are instances of \<S>:
-the \emph{Consistency of Shareholders} property is easily violated at bytecode level.
-
-In general, the problem arises whenever a method declares a formal parameter of generic type,
-such as \<buyer> in method \<accept> in Fig.~\ref{fig:shared_entity}. Next section shows
-the actual significance of the consequent security risk.
-
 \section{An Attack to the Shared Entities Contract}\label{sec:attack}
@@
 \end{figure}
-
-\section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}
-
-\section{Conclusion}\label{sec:conclusion}
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2021-06-28 040_solution.tex
struttura file latex

--- 2021-06-08 generics.tex
@@
-\documentclass{llncs}
-
-\usepackage{url}
-\usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage[hidelinks]{hyperref}
-\usepackage{pdfpages}
-\usepackage{relsize}
-\usepackage{tcolorbox}
-
-\usepackage{svg}
-\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\includesvg[height = 2ex]{svg-inkscape/ORCID_iD}}}
-
-\newcommand{\ie}{\textit{, ie.\ }}
-
-\def\codesize{\smaller}
-\def\<#1>{\codeid{#1}}
-\newcommand{\codeid}[1]{\ifmmode{\mbox{\codesize\ttfamily{#1}}}\else{\codesize\ttfamily #1}\fi}
-
-\usepackage{listings, xcolor}
-
-\renewcommand{\UrlFont}{\ttfamily\codesize}
-
-\definecolor{verylightgray}{rgb}{.97,.97,.97}
-
-\lstdefinelanguage{Takamaka}{
-        keywords=[1]{abstract, break, case, catch, class, continue, default, do
-, else, false, finally, for, if, final, implements, extends, import, instanceof, interface, length, new, private, protected, public, return, super, switch, this, throw, true, try, while, var, null}, % generic keywords
-        keywordstyle=[1]\color{blue}\bfseries,
-        keywords=[2]{boolean, int, long, float, double, byte, short, char, void, enum}, % types; money and time units
-        keywordstyle=[2]\color{teal}\bfseries,
-        keywords=[3]{@Override,@View,@FromContract,@Payable}, % annotations
-        keywordstyle=[3]\color{violet}\bfseries,
-        identifierstyle=\color{black},
-        sensitive=false,
-        comment=[l]{//},
-        morecomment=[s]{/*}{*/},
-        commentstyle=\color{gray}\ttfamily,
-        stringstyle=\color{red}\ttfamily,
-        morestring=[b]',
-        morestring=[b]"
-}
-
-\lstset{
-        language=Takamaka,
-        backgroundcolor=\color{verylightgray},
-        extendedchars=true,
-        basicstyle=\scriptsize\ttfamily,
-        showstringspaces=false,
-        showspaces=false,
-        numbers=none,
-        numberstyle=\scriptsize,
-        numbersep=9pt,
-        tabsize=2,
-        breaklines=true,
-        showtabs=false,
-        captionpos=b
-}
-
-\begin{document}
-
-\title{Power and Pitfalls of Generic Smart Contracts}
-\titlerunning{Power and Pitfalls of Generics for Smart Contracts}
-\author{Andrea Benini \and Sara Migliorini \and Fausto Spoto}
-\institute{Dipartimento di Informatica, Universit\`a di Verona, Italy}
-
-\maketitle
-
-\begin{abstract}
-  TBD
-  \keywords{smart contract \and generics \and blockchain}
-\end{abstract}
-
-\section{Introduction}\label{sec:introduction}
-
-Blockchains exploit the redundant, concurrent execution of the same
-transactions on a decentralized network of many machines
-in order to enforce their execution in accordance to
-a set of predefined rules. Namely, blockchains make it hard, for a single machine,
-to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
-gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
-has been the first blockchain's success story. Here
-transactions are programmed in a non-Turing complete bytecode language,
-almost exclusively used to implement transfers of units of coins between \emph{accounts}.
-
-A few years after Bitcoin, another blockchain, called
-Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
-transactions in an actual, imperative and Turing-complete programming language, called Solidity.
-Solidity's code is organized in \emph{smart contracts}, that can be seen as
-objects that control money.
-Ethereum's transactions can hence execute much more than coin transfers. Namely,
-they run object constructors and methods, which results in a sort
-of \emph{world computer} that persists the same objects in the memory of all the
-computers in the blockchain's network.
-
-In Solidity's bytecode,
-non-primitive values are referenced through a very general
-\<address> type. For instance, a Solidity method
-\<child(Person p, uint256 n) returns Person> actually compiles
-into \<child(address p, uint256 n) returns address>, losing most
-type information~\cite{CrafaPZ19}.
-At run time, it is the bytecode that gets executed. Hence,
-everything can be passed for \<p>, not just a \<Person>.
-The compiler cannot even enforce strong typing
-by generating defensive type instance checks and casts, since
-values are unboxed in Ethereum: they have no attached
-type information at run time,
-they are just numerical \emph{addresses}.
-It follows that calls to \<Person>'s methods
-on \<p> might actually execute arbitrary code, if \<p> is not a \<Person>.
-In other words, Solidity is not strongly typed.
-Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
-to our methods, such as on \<p> passed to \<child>, since an attacker can pass crafted
-objects for \<p>, with arbitrary implementations for their methods,
-which can result in the unexpected execution of
-dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
-costed millions of dollars.
-
-Strong typing is one of the reasons that push towards the adoption
-of \emph{traditional} programming languages for smart contracts. For instance,
-the Cosmos blockchain~\cite{cosmos} uses Go. The
-Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
-for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
-Hyperledger~\cite{hyperldeger} allows Go and Java.
-Another reason is the availability of modern
-language features, missing in Solidity,
-such as \emph{generics}\ie the possibility of using
-type variables. Generics are powerful and very useful for programming
-smart contracts. In Java source code, they are strongly typed, if no \emph{unchecked operations}
-are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics in Java are compiled by \emph{erasure}\ie replaced
-by their upwards bound, most often \<Object>. Hence, the risk
-is that generics introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity.
-
-The contribution of this paper is to show a real-life
-use of generics for an actual smart contract used in the support
-library of the Takamaka language, and to show that a naive use
-of Java generics can lead to a code security vulnerability that
-allows an attacker to earn money by exploiting someone else's work.
-This paper will provide a fix to that specific issue,
-by obliging the compiler to generate defensive checks.
-More generally, this paper can be useful for the definition of
-bytecode languages for future smart contract languages, by
-learning from the weaknesses of Java bytecode.
-
-The rest of this paper is organized as follows.
-Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contracts that uses generics. Sec.~\ref{sec:attack} shows that a naive
-deployment of that contract leads to a code vulnerability.
-Sec.~\ref{sec:fix} shows a fix to that vulnerability.
-Sec.~\ref{sec:conclusion} concludes.
-
-\section{Shared Entities using Generics}\label{sec:shared_entities}
-
-A \emph{shared entity} is something divided into \emph{shares}. Participants,
-that hold shares, are called \emph{shareholders} and can dynamically
-sell and buy shares. An example of a shared entity is a comporation,
-where shares represent units of possess of the company. Another example is
-a voting community, where shares represent the voting power of each given voter.
-A further example are the validator nodes of a proof of stake blockchain,
-where shares represent their voting power and remuneration percentage.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract,O extends Offer<S>> {
-
-  @View
-  BigInteger sharesOf(S shareholder);
-
-  @FromContract(PayableContract.class) @Payable
-  void place(BigInteger amount, O offer);
-
-  @FromContract(PayableContract.class) @Payable
-  void accept(BigInteger amount, S buyer, O offer);
-
-  class Offer<S extends PayableContract> extends Storage {
-    public final S seller;
-    public final BigInteger sharesOnSale;
-    public final BigInteger cost;
-    public final long expiration;
-
-    public Offer
-        (S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
-
-      this.seller = seller;
-      this.sharesOnSale = sharesOnSale;
-      this.cost = cost;
-      this.expiration = now() + duration;
-    }
-
-    @View
-    public boolean isOngoing() {
-      return now() <= expiration;
-    }
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
-\end{figure}
-
-In general, two concepts are specific to each implementation of shared entities:
-who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
-Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
-It includes an inner class \<Offer> that models sale offers:
-it specifies who is the seller of the shares,
-how many shares are being sold, the requested price and the expiration of the offer.
-Method \<isOngoing> checks if an offer has not expired yet.
-Implementations can subclass \<Offer> if they need more specific offers.
-By using class \<Offer>, the \<SharedEntity> interface specifies three methods.
-Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
-It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
-\emph{for free}, without paying gas, since it has no side-effects. Who wants to sell
-(some of) its shares calls method \<place> with a sale \<offer>.
-This method is annotated as \<@Payable> since
-implementations are allowed to require a payment of \<amount> coins for managing the sale.
-Who buys the shares calls method \<accept> with the accepted \<offer>
-and with itself as \<buyer> (the reason will be explained soon)
-and becomes a new shareholder or increases
-its cumulative number of shares (if it was a shareholder already).
-Also this method is \<@Payable>, since its caller must pay \<amount> $\ge$ \<offer.cost>
-coins to the seller.
-This means that shareholders must be able to receive payments and that
-is why \<S extends PayableContract>: the \<PayableContract>s are those that can receive
-payments in Takamaka.
-
-The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This attests, explicity, their will
-to sell or buy shares. The callers must be (old or new) shareholders,
-hence of type \<S>. In Takamaka, this could be written
-as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Because of this limitation,
-the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
-which allows \emph{any} \<PayableContract> to call these methods, not just those of type \<S>.
-Since the syntax of the language does not support our abstraction, we will have to
-program explicit dynamic checks in code, as shown in a moment.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class SimpleSharedEntity
-      <S extends PayableContract,O extends Offer<S>>
-      extends Contract
-      implements SharedEntity<S,O> {
-
-  private final StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private final StorageSet<O> offers = new StorageTreeSet<>();
-
-  public SimpleSharedEntity(S[] shareholders, BigInteger[] shares) {
-    require(shareholders.length == shares.length, "length mismatch");
-    for (int pos = 0; pos < shareholders.length; pos++)
-      addShares(shareholders[pos], shares[pos]);
-  }
-
-  @Override @View
-  public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, BigInteger.ZERO);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "not authorized to sell");
-    require(shares.containsKey(offer.seller), "only shareholders can sell");
-    require(sharesOf(offer.seller).subtract(sharesOnSaleOf(offer.seller))
-        .compareTo(offer.sharesOnSale) >= 0, "not enough shares to sell");
-    offers.add(offer);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "only the future owner can buy the shares");
-    require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the sale offer is not ongoing anymore");
-    require(offer.cost.compareTo(amount) <= 0, "not enough money");
-    offers.remove(offer);
-    removeShares(offer.seller, offer.sharesOnSale);
-    addShares(buyer, offer.sharesOnSale);
-    offer.seller.receive(offer.cost);
-  }
-
-  private BigInteger sharesOnSaleOf(S shareholder) {
-    return offers.stream()
-      .filter(offer -> offer.seller == shareholder && offer.isOngoing())
-      .map(offer -> offer.sharesOnSale)
-      .reduce(ZERO, BigInteger::add);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}.}\label{fig:simple_shared_entity}
-\end{figure}
-
-Fig.~\ref{fig:simple_shared_entity} shows our implementation of the \<SharedEntity> interface
-in Fig.~\ref{fig:shared_entity}
-by using two fields: \<shares> maps each shareholder to its amount of shares and
-\<offers> collects the offers that have been placed.
-The constructor populates the map \<shares>, initially.
-Method \<sharesOf> simply accesses
-\<shares>, by using zero as default. Method \<place> requires its \<caller()> to be
-the seller identified in the \<offer>. This forbids shareholders to sell shares on behalf of others.
-Moreover, this guarantees that the caller has type \<S>, like \<offer.seller>.
-As we said before, this cannot be expressed with the syntax of the language.
-Method \<place> further requires the seller to be a shareholder with at least \<offer.sharesOnSale>
-shares not yet placed on sale. This forbids to oversell more shares
-than one owns. At the end, \<place> adds the \<offer> to the set of \<offers>.
-Method \<accept> requires that who calls the method must be \<buyer>. Hence, successful
-calls to \<accept> can only pass the same caller for \<buyer>. This is a trick to enforce the
-caller to have type \<S>, since the syntax of the language does not allow one to express it,
-as we explained before. Then \<accept> requires the \<offer> to exists, to be still ongoing
-and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,
-the \<offer> is removed from the \<offers>, shares are moved from seller to buyer (code not
-shown in Fig.~\ref{fig:simple_shared_entity}) and the seller of the \<offer>
-receives the required price \<offer.cost>.
-
-Let us state an important property about this code:
-
-\begin{tcolorbox}
-  \begin{center}\emph{Consistency of Shareholders}\end{center}
-  If \<sse> is a {\codesize\texttt{SimpleSharedEntity<S,O>}}
-  then the domain of its map \<shares> consists of objects of type \<S> only.
-\end{tcolorbox}
-
-\noindent
-It turns out that this property holds.
-Namely, the code does not use unchecked casts, hence it is strongly-typed and
-the map \<shares> actually holds values of type \<S> in its domain, only.
-For this consistency result, we had to pay the price
-of the dummy \<buyer> argument for method \<accept>. Without that argument,
-\emph{Consistency of Shareholders} would not hold, since we could only write
-\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
-Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-not strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
-and become shareholders.
-
-There is, however, a problem with the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing \emph{of the source code}. But what is installed
-and executed in blockchain is the Java bytecode from Fig.~\ref{fig:simple_shared_entity}.
-It is important to investigate whether it is also strongly typed or not.
-
-\section{An Attack to the Shared Entities Contract}\label{sec:attack}
-
 \section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}
-
-\section{Conclusion}\label{sec:conclusion}
-
-\bibliographystyle{plain}
-\bibliography{biblio}
-
-\end{document}

### 2021-06-28 030_generics_attack.tex
Reorganized Section 3

@@
 consensus, which means that a selected dynamic subset of the nodes is in charge of
-validating the transactions and voting their acceptance. Validator nodes are
+validating the transactions and voting their acceptance. In general, validator nodes are
 represented by \<Validator> objects, that are externally owned accounts
-with an extra identifier derived from their public key (see Fig.~\ref{fig:validator}).
+with an extra identifier. In the specific case of Tendermint,
+validators are \<TendermintED25519Validator> objects whose
+identifier is derived from their ed25519 public key (see Fig.~\ref{fig:validator}).
 This identifier is public information, reported in the blocks or easily eavesdropped.
-At each block, the reward for the block is distributed to the validators.
 Tendermint applications can implement their own
-policy for changing the validator set dynamically. In our case, the validator set
-and the distribution of the reward is implemented by a subclass
-\<Validators> of \<SimpleSharedEntity> whose type \<S> for the shareholders
-has been set to \<Validator>. Shares are voting power in this case.
+policy for changing the validator set dynamically.
+The \<AbstractValidators> class implements the validator set
+and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
+(see Fig.~\ref{fig:tendermint_validators}).
+Shares are voting power in this case.
+Its subclass \<TendermintValidators> restricts the type \<S> of the shareholders
+to \<TendermintED25519Validator>.
+At each block committed, Hotmoka calls the \<reward> method of \<AbstractValidators>
+in order to reward the validators that behaved correctly
+and punish those that misbehaved. They are specified by two strings
+that contain the identifiers of the validators, as provided by the underlying
+Tendermint engine.
 
@@
     \begin{lstlisting}[language=Takamaka]
-public final class Validator extends ExternallyOwnedAccount {
-  private final String id;
+public abstract class Validator extends ExternallyOwnedAccount {
 
@@
     super(publicKey);
-    this.id = /* derived from publicKey as specified in Tendermint's spec */
   }
 
-  public @View String id() {
+  public abstract @View String id();
+}
+
+public final class TendermintED25519Validator extends Validator {
+
+  // the first 40 characters of the hexadecimal representation
+  // of the sha256 hashing of the public key bytes
+  private final String id;
+
+  public TendermintED25519Validator(String publicKey) {
+    super(publicKey);
+    this.id = /* derived from publicKey as in the comment for id above */
+  }
+
+  public final @Override @View String id() {
     return id;
@@
 
-Since shares of a \<SimpleSharedEntity> can be sold and bought, the set of validators
-is dynamic. Users can sell or buy voting power in order to invest in the blockchain
-and earn rewards. At each block creation, Hotmoka calls method \<getShareholders>
-on the shared entity and informs the
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public abstract class AbstractValidators<V extends Validator>
+    extends SimpleSharedEntity<V, Offer<V>> {
+
+  public AbstractValidators(V[] validators, BigInteger[] powers) {
+    super(validators, powers);
+  }
+
+  public void reward
+    (String behaving, String misbehaving, BigInteger gasConsumed) { ... }
+}
+
+public class TendermintValidators
+    extends AbstractValidators<TendermintED25519Validator> {
+
+  public TendermintValidators
+        (TendermintED25519validator[] validators, BigInteger[] powers) {
+    super(validators, powers);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{The shared entity of the validators of a Hotmoka blockchain built over Tendermint.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}.}\label{fig:tendermint_validators}
+\end{figure}
+
+Since \<SimpleSharedEntity> allows shares to be sold and bought, this holds for
+its \<TendermintValidators> subclass as well: the set of validators
+is dynamic and it is possible to sell and buy voting power in order to invest in the blockchain
+and earn rewards at each block committed. At block creation time,
+Hotmoka calls method \<getShareholders> inherited from
+\<SimpleSharedEntity> and informs the
 underlying Tendermint engine about the identifiers of the validator nodes for the next blocks.
@@
 
-It is important that the shareholders be instances of \<Validator>, since that class enforces the
-match between their public key, that identifies who can use the reward sent to the validator,
-and the Tendermint identifier of the validator, that identifies which node of the blockchain
+It is important that the \emph{Consistency of Shareholders} property holds
+for \<TendermintValidators>, since
+the shareholders of a \<TendermintValidators>
+must be instances of \<TendermintED25519Validator> (as declared in the generic
+signature of \<TendermintValidators>): class \<TendermintED25519Validator> enforces a
+match between its public key, that identifies who can spend the rewards sent to the validator,
+and its Tendermint identifier, that identifies which node of the blockchain
 must do the validation work (Fig.~\ref{fig:validator}).
@@
 public key (see Fig.~\ref{fig:attacker}):
-the node (belonging to the \emph{victim}) would do the work while the owner
-of the private key of the \<Attacker> could just wait and earn the reward.
+Tendermint would expect the node (belonging to the \emph{victim}) to do
+the validation work while the owner
+of the private key of the \<Attacker> could just wait for accrued awards to spend.
 A sort of validator's slavery.
+Sec.~\ref{sec:shared_entities} asserted that the \emph{Consistency of Shareholders}
+property holds, at source level.
+Namely, an \<attacker> (of type \<Attacker>) can only become shareholder
+by accepting an ongoing sale \<offer> of shares through a call to
+\<tv.accept(offer.cost, attacker, offer)> (see Fig.~\ref{fig:simple_shared_entity}).
+This is impossible at source level, where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
+is not an instance of \<S>, which has been set to \<TendermintED25519Validator>.
+But a Hotmoka blockchain contains only the bytecode of \<SimpleSharedEntity>,
+where the signature of \<accept> has been erased into
+\<accept(BigInteger amount, PayableContract buyer, Offer offer)>.
+Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
+\emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
+are \<PayableContract>s.
+That transaction adds \<attacker> to the shareholders of \<tv>,
+therefore violating the \emph{Consistency of Shareholders} property and allowing
+validator's slavery.
 
@@
   public @View String id() {
-    return /* id of the victim node of the blockchain */
+    return /* id of the victim blockchain node, unrelated to publicKey */
   }
@@
   \end{center}
-  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the reward of its work.}\label{fig:attacker}
+  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the rewards of its work.}\label{fig:attacker}
 \end{figure}

### 2021-06-28 030_generics_attack.tex
Small edits

@@
 
-  public abstract @View String id();
+  @View
+  public abstract String id();
 }
@@
 
-  public final @Override @View String id() {
+  @Override @View
+  public final String id() {
     return id;
@@
 
-  public @View String id() {
+  @View
+  public String id() {
     return /* id of the victim blockchain node, unrelated to publicKey */

### 2021-06-28 030_generics_attack.tex
Improved sentences

@@
 It is important that the \emph{Consistency of Shareholders} property holds
-for \<TendermintValidators>, since
-the shareholders of a \<TendermintValidators>
-must be instances of \<TendermintED25519Validator> (as declared in the generic
-signature of \<TendermintValidators>): class \<TendermintED25519Validator> enforces a
-match between its public key, that identifies who can spend the rewards sent to the validator,
-and its Tendermint identifier, that identifies which node of the blockchain
+for \<TendermintValidators>: its shareholders
+must be \<TendermintED25519Validator>s (as declared in the generic
+signature of \<TendermintValidators>) that enforce a
+match between their public key, that identifies who can spend the rewards sent to the validator,
+and their Tendermint identifier, that identifies which node of the blockchain
 must do the validation work (Fig.~\ref{fig:validator}).

### 2021-06-28 020_shared_entities_with_generics.tex

@@
   If \<sse> is a {\codesize\texttt{SharedEntity<S,O>}}
-  then the elements in the list \<sse.getShareholders()> have type \<S>.
+  then the elements contained in the list \<sse.getShareholders()> have type \<S>.
 \end{tcolorbox}

### 2021-06-28 050_related_work.tex
Related work continue

\section{Related Work}\label{sec:related_work}

\todo{Forse questa parte va spostata.}

\textbf{Generics Compilation --} Generics is a way for writing code parametrized by type, in other words a same piece of code can be written by using a type placehorder (ie. parameter) and then instantiated it for a specific type arguments. There exists two common ways to implement generics in a programming language that are often described in literature\,\cite{generics_categories} as \textit{heterogeneous} and \textit{homogeneous}. In the heterogeneous approach the code is specialized for each instance of the generic parameter, this is the approach adopted by the C++ \emph{templates}. Conversely, the homogeneous approach is the one provided by Java or .Net, in this case only one instance of the code is maintained that is shared by all generic instances. This implementation is based on the type \emph{erasure} mechanism in which the generic parameter is replaced by the upwards bound of each instance, most often \<Object>.
%
Even if the heterogeneous approach is the most safe one, it is rarely applied in particular in resource-constrained applications, because in this case the code size  may dramatically increase as some code is duplicated \cite{generics_embedded_systems}. Conversely, the latter approach ensures a smaller consumption of resources but can lead to the problems described in this paper. \\

\noindent
\textbf{Smart contracts vulnerabilities --} It has been estimated that on average, software developers make 100 to 150 errors for every thousand lines \cite{software_engineering}.
In 2002, the National Institute of Standards and Technology (NIST) estimates that the economic costs of faulty software in the US is about tens of billions of dollars per year and represent approximately just under 1 percent of the nation's gross domestic product (GDP). The effects induced by errors in software development are even worse when such piece of software represents a smart contract for two main reasons. First of all, it is usually impossible to change a smart contract once it has been deployed, the immutability is one of its main characteristics and errors are treated as intended behaviours. Secondly, smart contracts often store and
manage critical information such as digital assets and identities. For these reason several analysis of smart contracts vulnerabilities have been performed and some static analysis tools have been developed.

\begin{itemize}
\item Solythesis \cite{solythesis_solidity_validation}: A compilation tool for smart contracts which provides an expressive language for specifying desired safety invariants. Given a smart contract and a set of user defined invariants, it is able to produce a new enriched contract which will reject all transactions that violate such invariants.

\item Another solutions which is based on bytecode rewriting is presented in \cite{bytecode_rewriting}, where the authors propose the enforcement of security polices through the enhancement of EVM bytecode. More specifically, the disassembled bytecode is instrumented through new security guard code that enforces the desired policy. Their initial efforts are mainly focused on the  verification of arithmetic operations, such as the prevention of overflows; while in the future they will focus on
adding the verification of memory access operations.

\item SafeVM \cite{safevm}:  A  verification tool for Ethereum smart contracts which directly works on EVM bytecode and exploits the state-of-the-art verification engines already available for C programs. The basic idea is to take as input a smart contract in compiled EVM\ bytecode which can eventually contain some \<assert> or \<require> annotations, then such bytecode is decompiled and converted into a C program with \<ERROR> annotations. This produced C program can be verified by using existing verification tools.

\item In \cite{hol_smart_constracts} the authors propose the use of the existing Isabelle/HOL tool with the specification of a formal logic for EVM bytecode, in order to provide a verification tool for Ethereum smart contracts. More specifically, the desired properties  of contracts are stated in a pre/postcondition style, while the verification is done by recursively structuring contracts as a set of basic blocks
down to the level of instructions.

\item Scilla \cite{scilla} is a new programming language for safe smart contracts which as been tailored taking   System F as a foundational calculus. It is able to provide strong safety guarantees by means of type soundness. Thanks to its minimalistic nature, it has been possible to define also a generic and extensible framework for lightweight verification of smart contracts by means of user-defined domain-specific analyse. \todo{Ci sono i generics oppure essendo un linguaggio minimale non si pongono il problema??}

\end{itemize}

Given the plurality of verification tools that have been developed for indentifying security bugs in smart contracts, a unified and standard approach for evaluating their accuracy and performances have been developed in \cite{effectiveness_analysis_tools}. This automated and systematic approach, called SolidiFI, consists in injecting bugs into all potential locations in a smart contract, in order to introduce different kinds of vulnerabilities, and then using several static analysis tools to identify which bugs each of them is able to detect.

### 2021-06-29 050_related_work.tex
Related work completed

@@
 \noindent
-\textbf{Smart contracts vulnerabilities --} It has been estimated that on average, software developers make 100 to 150 errors for every thousand lines \cite{software_engineering}.
-In 2002, the National Institute of Standards and Technology (NIST) estimates that the economic costs of faulty software in the US is about tens of billions of dollars per year and represent approximately just under 1 percent of the nation's gross domestic product (GDP).
-The effects induced by errors in software development are even worse when such piece of software represents a smart contract for two main reasons.
-First of all, it is usually impossible to change a smart contract once it has been deployed, the immutability is one of its main characteristics and errors are treated as intended behaviours.
-Secondly, smart contracts often store and
+\textbf{Smart contracts vulnerabilities --} It has been estimated that on average, software developers make from 100 to 150 errors for every thousand lines of code \cite{software_engineering}.
+In 2002, the National Institute of Standards and Technology (NIST) estimates that the economic costs of faulty software in the US is about tens of billions of dollars per year and represent approximately just under 1 percent of the Nation's Gross Domestic Product (GDP).
+The effects induced by errors in software development are even worse when such pieces of software represents a smart contract.
+Indeed, it is usually impossible to change a smart contract once it has been deployed, the immutability is one of its main characteristics and errors are treated as intended behaviours.
+Moreover, smart contracts often store and
 manage critical information such as digital assets and identities.
-For these reason several analysis of smart contracts vulnerabilities have been performed and some static analysis tools have been developed.
+For this reason smart contract vulnerabilities and correctness is becoming an important topic in literature.
+Possible solutions can be classified into three main categories: (i) static analysis of EVM bytecode, (ii) automatic rectification of EVM bytecode and (iii) development of new languages for smart contract definition.
 
-\begin{itemize}
-\item Solythesis \cite{solythesis_solidity_validation}: A compilation tool for smart contracts which provides an expressive language for specifying desired safety invariants.
+Given the plurality of languages currently available for the design of smart contracts, static analysis is usually performed directly on the EVM bytecode, in order to make the solution general enough and promote its adoption.
+At this regards, SafeVM \cite{safevm} is s verification tool for Ethereum smart contracts which directly works on the EVM bytecode and exploits the state-of-the-art verification engines already available for C programs.
+The basic idea is to take as input a smart contract in compiled EVM\ bytecode which can eventually contain some \<assert> or \<require> annotations, then such bytecode is decompiled and converted into a C program with \<ERROR> annotations.
+This produced C program can be verified by using existing verification tools.
+%
+In \cite{hol_smart_constracts} the authors propose a verification tool for Ethereum smart contracts which is based on the use of the existing Isabelle/HOL tool together with the specification of a formal logic for EVM bytecode.
+More specifically, the desired properties  of contracts are stated in a pre/postcondition style, while the verification is done by recursively structuring contracts as a set of basic blocks
+down to the level of instructions.
+%
+Finally, another tool for the analysis of properties in  EVM bytecode is called EthIR\,\cite{ethir}.
+This open-source tool allows the precise decompilation of EVM bytecode into a high-level representation in a rule-based form.
+Given such representation, properties can be inferred throuth available state-of-the-art analysis tools developed for high-level languages.
+More specifically, it relies on an extension of Oyente, a tool that generates control-flow graphs of the code in order to derive a rule-based representation of the bytecode.
+
+Relatively to the automatic rectification of smart contracts, Solythesis \cite{solythesis_solidity_validation} is compilation tool for smart contracts which provides an expressive language for specifying desired safety invariants.
 Given a smart contract and a set of user defined invariants, it is able to produce a new enriched contract which will reject all transactions that violate such invariants.
-
-\item Another solutions which is based on bytecode rewriting is presented in \cite{bytecode_rewriting}, where the authors propose the enforcement of security polices through the enhancement of EVM bytecode.
+%
+Another solutions which is based on bytecode rewriting is presented in \cite{bytecode_rewriting}, where the authors propose the enforcement of security polices through the enhancement of EVM bytecode.
 More specifically, the disassembled bytecode is instrumented through new security guard code that enforces the desired policy.
@@
 adding the verification of memory access operations.
-
-\item SafeVM \cite{safevm}:  A  verification tool for Ethereum smart contracts which directly works on EVM bytecode and exploits the state-of-the-art verification engines already available for C programs.
-The basic idea is to take as input a smart contract in compiled EVM\ bytecode which can eventually contain some \<assert> or \<require> annotations, then such bytecode is decompiled and converted into a C program with \<ERROR> annotations.
-This produced C program can be verified by using existing verification tools.
-
-\item In \cite{hol_smart_constracts} the authors propose the use of the existing Isabelle/HOL tool with the specification of a formal logic for EVM bytecode, in order to provide a verification tool for Ethereum smart contracts.
-More specifically, the desired properties  of contracts are stated in a pre/postcondition style, while the verification is done by recursively structuring contracts as a set of basic blocks
-down to the level of instructions.
-
-\item Scilla \cite{scilla} is a new programming language for safe smart contracts which as been tailored taking   System F as a foundational calculus.
-It is able to provide strong safety guarantees by means of type soundness.
-Thanks to its minimalistic nature, it has been possible to define also a generic and extensible framework for lightweight verification of smart contracts by means of user-defined domain-specific analyse.
-\todo{Ci sono i generics oppure essendo un linguaggio minimale non si pongono il problema??}
-
-\end{itemize}
+%
+SMARTSHIELD \cite{smartshield} is another tool for automatically rectifying bytecode with the aim to fix three typical security bugs in smart contracts:
+(i) state changes after external calls, (ii) missing checks for out-of-bound arithmetic operations, and (iii) missing checks for failing external calls.
+More specifically, given an identified issue, the tool performs a semantic-preserving code transformation to ensure that only the insecure code patterns are revised, eventually  sending the rectification suggestions back to the developers when the eventual fixes can lead to side effects.
+The tool not only guarantees that the rectified contracts are immune to certain attacks but also they are gas-friendly.
+Indeed, it adopts heuristics to optimize gas consumption.
 
@@
 This automated and systematic approach, called SolidiFI, consists in injecting bugs into all potential locations in a smart contract, in order to introduce different kinds of vulnerabilities, and then using several static analysis tools to identify which bugs each of them is able to detect.
+
+Finally, as regards to the definition of new programming languages for safe smart contracs,  Scilla \cite{scilla} is one of them.
+It has been tailored taking   System F as a foundational calculus and it is able to provide strong safety guarantees by means of type soundness.
+Thanks to its minimalistic nature, it has been possible to define also a generic and extensible framework for lightweight verification of smart contracts by means of user-defined domain-specific analyse.
+\todo{Ci sono i generics oppure essendo un linguaggio minimale non si pongono il problema??}

### 2021-07-02 030_generics_attack.tex
Added Section 4 on fixing the issue

@@
 and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
-(see Fig.~\ref{fig:tendermint_validators}).
+(see Fig.~\ref{fig:abstract_validators}).
 Shares are voting power in this case.
@@
   \end{center}
-  \caption{The shared entity of the validators of a Hotmoka blockchain built over Tendermint.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}.}\label{fig:tendermint_validators}
+  \caption{The shared entity of the validators of a Hotmoka blockchain.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/AbstractValidators.java}.}\label{fig:abstract_validators}
 \end{figure}

### 2021-07-02 040_solution.tex
Reorganized Section 3
Added Section 4 on fixing the issue

@@
-\section{Fixing the Compilation of the Shared Entities Contract} \label{sec:fix}
+\section{Fixing the Compilation of the Contract} \label{sec:fix}
+
+The security issue in Sec.~\ref{sec:attack} is due to the
+over-permissive erasure of the signature of method \<accept>,
+where \<buyer> has type \<PayableContract>.
+Therefore, a solution is to oblige the compiler to generate a more
+restrictive signature where, in particular, the parameter \<buyer>
+has type \<TendermintED25519Validator>: only that type of accounts
+must be accepted for the validators.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Java]
+public class TendermintValidators
+    extends AbstractValidators<TendermintED25519Validator> {
+
+  public TendermintValidators
+        (TendermintED25519validator[] validators, BigInteger[] powers) {
+    super(validators, powers);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void accept(BigInteger amount, TendermintED25519Validator buyer, Offer<TendermintED25519Validator> offer) {
+    super.accept(amount, buyer, offer);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{The shared entity of the validators of a Hotmoka blockchain built over Tendermint.
+  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}.}\label{fig:solution}
+\end{figure}
+
+The fixed code is in Fig.~\ref{fig:solution}. The only difference is that method
+\<accept> has been redefined to force the correct type for \<buyer>. For the rest, that method
+delegates to its implementation in \<AbstractValidators>.
+It is important to investigate which is the Java bytecode generated from
+the code in Fig.~\ref{fig:solution}. Since Java bytecode does not allow one to redefine a method
+and modify its argument types, the compiled bytecode actually contains \emph{two}
+\<accept> methods, as follows:
+
+\begin{lstlisting}[language=JavaBytecode]
+public class TendermintValidators extends AbstractValidators {
+  ...
+
+  public void accept(BigInteger,TendermintED25519Validator,Offer)
+   aload_0
+   aload_1
+   aload_2
+   aload_3
+   invokespecial AbstractValidators.accept(BigInteger,PayableContract,Offer)
+   return
+
+  // synthetic bridge method
+  public void accept(BigInteger,PayableContract,Offer)
+   aload_0
+   aload_1
+   aload_2
+   checkcast TendermintED25519Validator
+   aload_3
+   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
+   return
+}
+\end{lstlisting}
+
+\noindent
+The first \<accept> method is the compilation of that from Fig.~\ref{fig:solution}:
+it delegates to the \<accept> method of the superclass \<AbstractValidators>. The second
+\<accept> method is a \emph{bridge method} that the compiler generates in order to guarantee
+that all calls to the erased signature \<accept(BigInteger,PayableContract,Offer)> actually
+get forwarded to the first, redefined \<accept>. It casts its \<buyer> argument
+into \<TendermintED25519Validator> and calls the first \<accept>. This
+bridge method and its checked cast guarantee that only \<TendermintED25519Validator>s
+can become validators. For instance, an instance of \<Attacker> (Fig.~\ref{fig:attacker})
+cannot be passed to the first \<accept> (type mismatch) and makes the second \<accept>
+fail with a class cast exception. The \emph{Consistency of Shareholders} holds
+for instances of \<TendermintValidators> now and the attack in Sec.~\ref{sec:attack} cannot occur
+anymore.

### 2021-07-02 090_conclusion.tex
Added conclusion

\section{Conclusion}\label{sec:conclusion}

This paper has shown that generic types can be useful in the definition
of smart contracts, but introduce risks of security since in many
programming languages, including Java, they get erased at compile time
into types that might be too permissive for low level calls, such as
those that start blockchain transactions. Note that using a programming
language without generics is not the solution: Solidity has no generics
and consequently erases \emph{all} reference types into \<address>.
That is the worst possible erasure.

The solution in this paper has been to redefine a method with an argument
of generic type in such a way to call its superclass
(see \<accept> in Fig.~\ref{fig:solution}). This fixes the security risk,
but cannot be regarded as an elegant solution. It is just a trick that
works because it forces the compiler to generate some kind of bytecode.
A \emph{smarter} compiler might recognize the redefined \<accept> as
\emph{useless} and just remove it. This would recreate the issue that
we have just solved. That is, our solution works only for the way
compilers compile \emph{today}.

With hindsight, it is questionable to have implemented generics by erasure
and code instrumentation (bridge methods). If generics would be present
and checked at bytecode level, the attack in Sec.~\ref{sec:attack} would just
be impossible. Currently, generics can only exist as bytecode annotations,
but they are not mandatory and are not used by the Java virtual machine
that runs the bytecode. The same consideration might be applied beyond
generics: many features of modern programming
languages have no direct low-level
counterpart but are implemented via instrumentation, starting from
inner classes and closures. This is fine at source level, but allows
low-level calls to easily circumvent
the encapsulation guarantees of the language. When immerged in a permissionless
blockchain, such features become dangerous attack surfaces.

### 2021-07-02 020_shared_entities_with_generics.tex
Small changes in the related works

--- 2021-06-28 020_shared_entities_with_generics.tex
@@
   \begin{center}\emph{Consistency of Shareholders}\end{center}
-  If \<sse> is a {\codesize\texttt{SharedEntity<S,O>}}
-  then the elements in the list \<sse.getShareholders()> have type \<S>.
+  If \<se> is a {\codesize\texttt{SharedEntity<S,O>}}
+  then the elements contained in the list \<se.getShareholders()> have type \<S>.
 \end{tcolorbox}

### 2021-07-02 050_related_work.tex
Small changes in the related works

--- 2021-06-28 050_related_work.tex
@@
 
-\textbf{Generics Compilation --} Generics is a way for writing code parametrized by type, in other words a same piece of code can be written by using a type placehorder (ie.
-parameter) and then instantiated it for a specific type arguments.
-There exists two common ways to implement generics in a programming language that are often described in literature\,\cite{generics_categories} as \textit{heterogeneous} and \textit{homogeneous}.
-In the heterogeneous approach the code is specialized for each instance of the generic parameter, this is the approach adopted by the C++ \emph{templates}.
-Conversely, the homogeneous approach is the one provided by Java or .Net, in this case only one instance of the code is maintained that is shared by all generic instances.
-This implementation is based on the type \emph{erasure} mechanism in which the generic parameter is replaced by the upwards bound of each instance, most often \<Object>.
+\textbf{Compilation of generics --}
+Generics is a way for writing code parametrized by type, in other words a same piece of code can
+be written by using a type placehorder (a \emph{type parameter}) and then instantiated for specific,
+concrete types. There exists two common ways to implement generics in a programming language,
+that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
+and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
+of the generic parameter; this is the approach adopted by C++ \emph{templates}.
+Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
+only one instance of the code is maintained and shared by all generic instances.
+This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
+is replaced by the upwards bound of each instance, most often \<Object>.
+Even if the heterogeneous approach is the safest, it is rarely applied in particular
+in resource-constrained applications, because the code size may dramatically increase
+as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
+with extra costs of gas.
+Conversely, the homogeneous approach ensures a smaller consumption
+of resources but can lead to the problems described in this paper. \\
+
+\textbf{Smart contracts vulnerabilities --}
+It has been estimated that, on average, software developers make from 100 to 150 errors
+for every thousand lines of code~\cite{software_engineering}.
+In 2002, the National Institute of Standards and Technology (NIST) estimates that the economic
+costs of faulty software in the US is about tens of billions of dollars per year and represent
+approximately just under 1 percent of the Nation's gross domestic product.
+The effects induced by errors in software development are even worse when such pieces of software
+are smart contracts. Indeed, it is usually impossible to change a smart contract once
+it has been deployed, the immutability being one of its main characteristics, so that
+errors are treated as intended behaviours. Moreover, smart contracts often store and
+manage critical data such as money, digital assets and identities. For this reason, smart contracts
+vulnerabilities and correctness are becoming important in literature. Possible solutions can be
+classified into three main categories: (i) static analysis of EVM bytecode,
+(ii) automatic rectification of EVM bytecode and (iii) development of new languages
+for smart contracts.
+
+Given the plurality of languages currently available for the design of smart contracts,
+static analysis is usually performed directly on the Ethereum bytecode, in order to make
+the solution general enough and promote its adoption. At this regards, SafeVM~\cite{safevm}
+is a verification tool for Ethereum smart contracts that works on bytecode and exploits
+the state-of-the-art verification engines already available for C programs.
+The basic idea is to take as input a smart contract in compiled bytecode, that
+can possibly contain some \<assert> or \<require> annotations, decompile it and convert it
+into a C program with \<ERROR> annotations. This C program can be verified by using
+existing verification tools.
 %
-Even if the heterogeneous approach is the most safe one, it is rarely applied in particular in resource-constrained applications, because in this case the code size  may dramatically increase as some code is duplicated \cite{generics_embedded_systems}.
-Conversely, the latter approach ensures a smaller consumption of resources but can lead to the problems described in this paper.
-\\
+In~\cite{hol_smart_constracts}, the authors propose a verification tool for Ethereum
+smart contracts based on the use of the existing Isabelle/HOL tool, together with the
+specification of a formal logic for Ethereum bytecode. More specifically, the desired properties
+of the contracts are stated in pre/postcondition style, while the verification
+is done by recursively structuring contracts as a set of basic blocks
+down to the level of instructions.
+%
+Another tool for the analysis of Ethereum bytecode is EthIR~\cite{ethir}. This open-source tool
+allows the precise decompilation into a high-level, rule-based representation. Given such
+representation, properties can be inferred through available state-of-the-art analysis tools
+for high-level languages. More specifically, EthIR relies on an extension of Oyente,
+a tool that generates code control-flow graphs in order to derive
+a rule-based representation of the bytecode.
 
-\noindent
-\textbf{Smart contracts vulnerabilities --} It has been estimated that on average, software developers make 100 to 150 errors for every thousand lines \cite{software_engineering}.
-In 2002, the National Institute of Standards and Technology (NIST) estimates that the economic costs of faulty software in the US is about tens of billions of dollars per year and represent approximately just under 1 percent of the nation's gross domestic product (GDP).
-The effects induced by errors in software development are even worse when such piece of software represents a smart contract for two main reasons.
-First of all, it is usually impossible to change a smart contract once it has been deployed, the immutability is one of its main characteristics and errors are treated as intended behaviours.
-Secondly, smart contracts often store and
-manage critical information such as digital assets and identities.
-For these reason several analysis of smart contracts vulnerabilities have been performed and some static analysis tools have been developed.
+Relatively to the automatic certification of smart contracts,
+Solythesis~\cite{solythesis_solidity_validation} is a compilation tool for smart contracts
+that provides an expressive language for specifying desired safety invariants.
+Given a smart contract and a set of user defined invariants,
+it is able to produce a new enriched contract that will reject all transactions
+violating the invariants.
+%
+Another solutions, based on bytecode rewriting, is presented in~\cite{bytecode_rewriting},
+where the authors propose the enforcement of security policies through the enhancement of bytecode.
+More specifically, the disassembled bytecode is instrumented through new security guard code
+that enforces the desired policy. Their initial efforts are mainly focused on the verification
+of arithmetic operations, such as the prevention of overflows. In the future, they plan to focus on
+verifying memory access operations.
+%
+SMARTSHIELD~\cite{smartshield} is another tool for automatically
+rectifying bytecode with the aim to fix three typical security bugs in smart contracts:
+(i) state changes after external calls, (ii) missing checks for out-of-bound arithmetic operations,
+and (iii) missing checks for failing external calls. More specifically, given an identified issue,
+the tool performs a semantic-preserving code transformation to ensure that only the insecure code
+patterns are revised, eventually sending the rectification suggestions back to the developers
+when the eventual fixes can lead to side effects. The tool not only guarantees that the rectified
+contracts are immune to certain attacks but also that they are gas-friendly.
+Indeed, it adopts heuristics to optimize gas consumption.
 
-\begin{itemize}
-\item Solythesis \cite{solythesis_solidity_validation}: A compilation tool for smart contracts which provides an expressive language for specifying desired safety invariants.
-Given a smart contract and a set of user defined invariants, it is able to produce a new enriched contract which will reject all transactions that violate such invariants.
+Given the plurality of verification tools that have been developed for indentifying security bugs
+in smart contracts, a unified and standard approach for evaluating their accuracy and performances
+have been developed in~\cite{effectiveness_analysis_tools}. This automated and systematic approach,
+called SolidiFI, consists in injecting bugs into all potential locations in a smart contract,
+in order to introduce different kinds of vulnerabilities, and then using several static analysis
+tools to identify which bugs each of them is able to detect.
 
-\item Another solutions which is based on bytecode rewriting is presented in \cite{bytecode_rewriting}, where the authors propose the enforcement of security polices through the enhancement of EVM bytecode.
-More specifically, the disassembled bytecode is instrumented through new security guard code that enforces the desired policy.
-Their initial efforts are mainly focused on the  verification of arithmetic operations, such as the prevention of overflows; while in the future they will focus on
-adding the verification of memory access operations.
-
-\item SafeVM \cite{safevm}:  A  verification tool for Ethereum smart contracts which directly works on EVM bytecode and exploits the state-of-the-art verification engines already available for C programs.
-The basic idea is to take as input a smart contract in compiled EVM\ bytecode which can eventually contain some \<assert> or \<require> annotations, then such bytecode is decompiled and converted into a C program with \<ERROR> annotations.
-This produced C program can be verified by using existing verification tools.
-
-\item In \cite{hol_smart_constracts} the authors propose the use of the existing Isabelle/HOL tool with the specification of a formal logic for EVM bytecode, in order to provide a verification tool for Ethereum smart contracts.
-More specifically, the desired properties  of contracts are stated in a pre/postcondition style, while the verification is done by recursively structuring contracts as a set of basic blocks
-down to the level of instructions.
-
-\item Scilla \cite{scilla} is a new programming language for safe smart contracts which as been tailored taking   System F as a foundational calculus.
+Finally, as regards to the definition of new programming languages for safe smart contracs,
+Scilla~\cite{scilla} has been tailored by taking System~F as a foundational calculus.
 It is able to provide strong safety guarantees by means of type soundness.
-Thanks to its minimalistic nature, it has been possible to define also a generic and extensible framework for lightweight verification of smart contracts by means of user-defined domain-specific analyse.
-\todo{Ci sono i generics oppure essendo un linguaggio minimale non si pongono il problema??}
-
-\end{itemize}
-
-Given the plurality of verification tools that have been developed for indentifying security bugs in smart contracts, a unified and standard approach for evaluating their accuracy and performances have been developed in \cite{effectiveness_analysis_tools}.
-This automated and systematic approach, called SolidiFI, consists in injecting bugs into all potential locations in a smart contract, in order to introduce different kinds of vulnerabilities, and then using several static analysis tools to identify which bugs each of them is able to detect.
+Thanks to its minimalistic nature, it has been possible to define also a generic and extensible
+framework for lightweight verification of smart contracts by means of user-defined domain-specific
+analyses. Scilla has generic types. \textbf{Si riesce a scrivere alcuni dei nostri contratti in Scilla per vedere se il codice compilato ha gli stessi problemi che Java?}

### 2021-07-07 001_abstract.tex
Added abstract

Generics allow the use of type placeholders in order
to write parametrized code, that can be instantiated for each
concrete type provided for the placeholders. In many
programming languages, such as Java, they are implemented by
\emph{erasure}\ie replaced by their upper bound type.
Generics are a powerful feature of programming languages that allows one
to write highly reusable code. Namely, this paper shows their use in the
context of smart contracts for blockchain, in order to implement
a contract for \emph{shared entities} (such as a company shared
by its shareholders),
for the Hotmoka blockchain whose contracts are written in Java.
It shows that a special case of shared entities is the validators set of
the blockchain itself. However, this power comes at the risk of a too permissive
typing of the compiled code, where generics have been erased.
This paper shows
that this allows an attack to the validators set and proposes a solution
that forces the compiler to generate more precise type information than
those arising by erasure.

### 2021-07-07 010_introduction.tex
Introduction review

@@
 objects that control money.
+\todo{A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary\,\cite{ebp}.}
 Ethereum's transactions can hence execute much more than coin transfers. Namely,
@@
 type information~\cite{CrafaPZ19}.
-At run time, it is the bytecode that gets executed. Hence,
-everything can be passed for \<p>, not just a \<Person>.
+Since at run time, it is the bytecode that gets executed, everything can be passed for \<p>, not just a \<Person> instance.
 The compiler cannot even enforce strong typing
-by generating defensive type instance checks and casts, since
+by generating defensive type instance checks and casts, because
 values are unboxed in Ethereum: they have no attached
@@
 they are just numerical \emph{addresses}.
-It follows that calls to \<Person>'s methods
-on \<p> might actually execute arbitrary code, if \<p> is not a \<Person>.
+It follows that \todo{inside the \<child> method, an eventual} call to a \<Person>'s method
+on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
 In other words, Solidity is not strongly typed.
@@
 Another reason is the availability of modern
-language features, missing in Solidity,
+language features, which are missed in Solidity,
 such as \emph{generics}\ie the possibility of using
-type variables. Generics are powerful and very useful for programming
-smart contracts. In Java source code, they are strongly typed, if no \emph{unchecked operations}
+type variables. Generics are a powerful and very useful facility for programming
+smart contracts, \todo{since they allow one to personalize the behaviour of such contracts and partially overcome the inherent incompleteness of contracts \cite{ebp}}.
+In Java source code, generics are strongly typed, if no \emph{unchecked operations}
 are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics in Java are compiled by \emph{erasure}\ie replaced
+However,  in Java they are compiled by \emph{erasure}\ie replaced
 by their upwards bound, most often \<Object>. Hence, the risk
@@
 use of generics for an actual smart contract used in the support
-library of the Takamaka language, and to show that a naive use
+library of the Takamaka language, and to show that a na\"{i}ve use
 of Java generics can lead to a code security vulnerability that
@@
 This paper will provide a fix to that specific issue,
-by obliging the compiler to generate defensive checks.
+by forcing the compiler to generate defensive checks.
 More generally, this paper can be useful for the definition of
@@
 Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contracts that uses generics. Sec.~\ref{sec:attack} shows that a naive
+contracts that uses generics. Sec.~\ref{sec:attack} shows that a na\"{i}ve
 deployment of that contract leads to a code vulnerability.
 Sec.~\ref{sec:fix} shows a fix to that vulnerability.
+Sec.~\ref{sec:related_work} discusses some related work and finally,
 Sec.~\ref{sec:conclusion} concludes.

### 2021-07-07 020_shared_entities_with_generics.tex
Introduction review

@@
 who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
+Therefore, we have parameterized the interface of a shared entity with two type variables\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
 Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
@@
 The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This attests, explicity, their will
+contracts can call these methods. This annotation explicitly states that the methods will
 to sell or buy shares. The callers must be (old or new) shareholders,
@@
 as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Because of this limitation,
+in \<S.class>. Due of this limitation,
 the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
@@
 \<offers> collects the offers that have been placed.
-The constructor populates the map \<shares>, initially.
+The constructor initially populates the map \<shares>.
 Method \<sharesOf> simply accesses
@@
 operations guarantees strong typing \emph{of source code} in Java. But what is installed
-and executed in blockchain is the Java bytecode derived from
+and executed in blockchain is the Java bytecode which has been derived from
 the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
@@
 In general, the problem arises whenever a method declares a formal parameter of generic type,
-such as \<buyer> in method \<accept> in Fig.~\ref{fig:shared_entity}. Next section shows
+such as \<buyer> in method \<accept> of Fig.~\ref{fig:shared_entity}. Next section shows
 the actual significance of the consequent security risk.

### 2021-07-07 040_solution.tex
Introduction review
Compilation fix

@@
 
-The fixed code is in Fig.~\ref{fig:solution}. The only difference is that method
+The fixed code is shown in Fig.~\ref{fig:solution}. The only difference is that method
 \<accept> has been redefined to force the correct type for \<buyer>. For the rest, that method
@@
 anymore.
+
+\todo{Si potrebbe dire che la soluzione e' simile ad una gestione dei generics eterogenea?
+che pero' deve essere fatta a mano ridefinendo i metodi, invece che essere generata in automatico dal compilatore?
+}

### 2021-07-14 010_introduction.tex
Small edits

@@
 objects that control money.
-\todo{A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary\,\cite{ebp}.}
+\todo{A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.}
 Ethereum's transactions can hence execute much more than coin transfers. Namely,
@@
 type information~\cite{CrafaPZ19}.
-Since at run time, it is the bytecode that gets executed, everything can be passed for \<p>, not just a \<Person> instance.
+Since, at run time, it is the bytecode that gets executed,
+everything can be passed for \<p>, not just a \<Person> instance.
 The compiler cannot even enforce strong typing
@@
 Another reason is the availability of modern
-language features, which are missed in Solidity,
+language features, that are missed in Solidity,
 such as \emph{generics}\ie the possibility of using
@@
 Sec.~\ref{sec:fix} shows a fix to that vulnerability.
-Sec.~\ref{sec:related_work} discusses some related work and finally,
+Sec.~\ref{sec:related_work} discusses some related work and, finally,
 Sec.~\ref{sec:conclusion} concludes.

### 2021-07-14 020_shared_entities_with_generics.tex
Small edits

@@
 who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
+Therefore, we have parameterized the interface of a shared entity with two type variables:
+\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
 Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
@@
 The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
-contracts can call these methods. This annotation explicitly states that the methods will
-to sell or buy shares. The callers must be (old or new) shareholders,
-hence of type \<S>. In Takamaka, this could be written
+contracts can call these methods.
+These callers must be (old or new) shareholders,
+hence they must have type \<S>. In Takamaka, this could be written
 as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Due of this limitation,
+in \<S.class>. Due to this limitation,
 the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
@@
 Since the syntax of the language does not support our abstraction, we will have to
-program explicit dynamic checks in code, as shown later.
+program explicit dynamic checks in code, as shown later, and this will be the reason of the
+parameter \<buyer> in \<accept>.
 
@@
 Namely, the code in Fig.~\ref{fig:simple_shared_entity}
-does not use unchecked casts, hence it is strongly-typed and
+does not use unchecked casts, hence it is strongly-typed~\cite{NaftalinW06} and
 the map \<shares> actually holds values of type \<S> in its domain, only.
@@
 This is because of the so called \emph{erasure} of generic types in the compilation of Java.
-It entails that an attacker might install in blockchain a snippet of bytecode that calls
+It follows that an attacker can install in blockchain a snippet of bytecode that calls
 \<accept> and passes \emph{any} \<PayableContract>, not only those that are instances of \<S>:

### 2021-07-14 030_generics_attack.tex
Small edits

@@
 If it were possible to add a shareholder of another
-type \<Attacker>, the latter could declare the identifier of a node that does not correspond to its
+type \<Attacker>, the latter could decouple the node identifier from its
 public key (see Fig.~\ref{fig:attacker}):
@@
 Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
+at bytecode level
 \emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
-are \<PayableContract>s.
+are instances of \<PayableContract>.
 That transaction adds \<attacker> to the shareholders of \<tv>,

### 2021-07-14 050_related_work.tex
Introduction review
Small edits

@@
 manage critical data such as money, digital assets and identities. For this reason, smart contracts
-vulnerabilities and correctness are becoming important in literature. Possible solutions can be
+vulnerabilities and correctness are becoming important in literature\,\cite{smart_contracts_verification}. Possible solutions can be
 classified into three main categories: (i) static analysis of EVM bytecode,
@@
 framework for lightweight verification of smart contracts by means of user-defined domain-specific
-analyses. Scilla has generic types. \textbf{Si riesce a scrivere alcuni dei nostri contratti in Scilla per vedere se il codice compilato ha gli stessi problemi che Java?}
+analyses. The type variables of the functional foundational calculus can be seen as
+generic types. We do not know how they are compiled and if the strong typing guarantee of the
+source code extends to the compiled code as well. Scilla contracts are developed with
+the Neo Savant online IDE. Currently, neither Neo Savant IDE nor the block explorer
+allow one to inspect the compiled bytecode.

### 2021-07-14 090_conclusion.tex
Small edits

@@
 and checked at bytecode level, the attack in Sec.~\ref{sec:attack} would just
-be impossible. Currently, generics can only exist as bytecode annotations,
-but they are not mandatory and are not used by the Java virtual machine
+be impossible. Currently, generics can only exist as bytecode annotations
+that are not mandatory and are ignored by the Java virtual machine
 that runs the bytecode. The same consideration might be applied beyond
@@
 low-level calls to easily circumvent
-the encapsulation guarantees of the language. When immerged in a permissionless
+the encapsulation guarantees of the language. When embedded in a permissionless
 blockchain, such features become dangerous attack surfaces.

### 2021-07-19 001_abstract.tex
new latex style and review of introduction

@@
-Generics allow the use of type placeholders in order
-to write parametrized code, that can be instantiated for each
-concrete type provided for the placeholders. In many
-programming languages, such as Java, they are implemented by
-\emph{erasure}\ie replaced by their upper bound type.
 Generics are a powerful feature of programming languages that allows one
-to write highly reusable code. Namely, this paper shows their use in the
+to write highly reusable code.
+%
+More specifically, they are based on the use of type placeholders in order
+to produce parametrized code, that can be instantiated for each
+concrete type provided for them. 
+%
+In many programming languages, such as Java, they are implemented by
+\emph{erasure}\ie replaced by their upper bound type during bytecode compilation.
+%
+This paper investigates the use of generics in the
 context of smart contracts for blockchain, in order to implement
@@
 by its shareholders),
-for the Hotmoka blockchain whose contracts are written in Java.
-It shows that a special case of shared entities is the validators set of
-the blockchain itself. However, this power comes at the risk of a too permissive
-typing of the compiled code, where generics have been erased.
-This paper shows
-that this allows an attack to the validators set and proposes a solution
+by using the Hotmoka blockchain whose contracts are written in Java.
+%
+The considered case study is particular important since
+a special case of shared entities could be the validators' set of
+the blockchain itself. The analysis shows that, the power of generics comes at the risk of a too permissive
+typing of the compiled code, due to the erasure mechanism, with a consequent possible attack
+to the validators' set. This paper proposes a solution
 that forces the compiler to generate more precise type information than

### 2021-07-19 010_introduction.tex
new latex style and review of introduction

@@
 
-Blockchains exploit the redundant, concurrent execution of the same
-transactions on a decentralized network of many machines
-in order to enforce their execution in accordance to
+\IEEEPARstart{B}{lockchains} exploit the redundant, concurrent execution of the same
+transactions on a decentralized network of many machines,
+in order to enforce their execution in accordance with
 a set of predefined rules. Namely, blockchains make it hard, for a single machine,
@@
 they run object constructors and methods, which results in a sort
-of \emph{world computer} that persists the same objects in the memory of all the
-computers in the blockchain's network.
+of \emph{world computer} that persists the same objects inside the memory of all the
+computers composing the blockchain's network.
 
@@
 Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
-to our methods, such as on \<p> passed to \<child>, since an attacker can pass crafted
+to another method, such as on \<p> passed to \<child>, since an attacker can pass crafted
 objects for \<p>, with arbitrary implementations for their methods,
@@
 type variables. Generics are a powerful and very useful facility for programming
-smart contracts, \todo{since they allow one to personalize the behaviour of such contracts and partially overcome the inherent incompleteness of contracts \cite{ebp}}.
+smart contracts, \todo{since they allow one to personalize the behaviour of such contracts and partially overcome the their inherent incompleteness\,\cite{ebp}}.
 In Java source code, generics are strongly typed, if no \emph{unchecked operations}
 are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However,  in Java they are compiled by \emph{erasure}\ie replaced
-by their upwards bound, most often \<Object>. Hence, the risk
-is that generics introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity.
+
+There exists two common ways to implement generics in a programming language,
+that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
+and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
+of the generic parameter; this is the approach adopted by C++ \emph{templates}.
+Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
+only one instance of the code is maintained and shared by all generic instances.
+This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
+is replaced by the upwards bound of each instance, mostly often \<Object>.
+Even if the heterogeneous approach is the safest, it is rarely applied in particular
+in resource-constrained applications, because the code size may dramatically increase
+as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
+with extra costs of gas.
+Conversely, the homogeneous approach ensures a smaller consumption
+of resources but it can introduce the same kind of attack to the bytecode
+as normal reference types do in Solidity, as we will describe in this paper.
 
 The contribution of this paper is to show a real-life
-use of generics for an actual smart contract used in the support
-library of the Takamaka language, and to show that a na\"{i}ve use
-of Java generics can lead to a code security vulnerability that
-allows an attacker to earn money by exploiting someone else's work.
+use of generics for an actual smart contract contained in the support
+library of the Takamaka language, and to demonstrate that a na\"{i}ve use
+of Java generics can lead to a code security vulnerability which
+allows an attacker to earn money by exploiting someone else's work, \todo{with both economical and legal side effects}.
 This paper will provide a fix to that specific issue,
-by forcing the compiler to generate defensive checks.
+by \todo{proposing a re-engineering of the code that forces} the compiler to generate defensive checks.
 More generally, this paper can be useful for the definition of
@@
 
-The rest of this paper is organized as follows.
+The remainder of this paper is organized as follows.
 Sec.~\ref{sec:shared_entities} introduces our real-life Java smart

### 2021-07-19 020_shared_entities_with_generics.tex
new latex style and review of introduction

@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
   \begin{center}
     \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract,O extends Offer<S>> {
+public interface SharedEntity<S extends PayableContract,
+                              O extends Offer<S>> {
 
@@
     public Offer
-        (S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
+        (S seller, BigInteger sharesOnSale, 
+         BigInteger cost, long duration) {
 
@@
 
-  private final StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private final StorageSet<O> offers = new StorageTreeSet<>();
+  private final StorageTreeMap<S,BigInteger> shares = 
+     new StorageTreeMap<>();
+  private final StorageSet<O> offers = 
+     new StorageTreeSet<>();
 
-  public SimpleSharedEntity(S[] shareholders, BigInteger[] shares) {
-    require(shareholders.length == shares.length, "length mismatch");
+  public SimpleSharedEntity
+    (S[] shareholders, BigInteger[] shares) {
+    require(shareholders.length == shares.length, 
+            "length mismatch");
     for (int pos = 0; pos < shareholders.length; pos++)
@@
   public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, BigInteger.ZERO);
+    return shares.getOrDefault
+      (shareholder, BigInteger.ZERO);
   }
@@
   public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "not authorized to sell");
-    require(shares.containsKey(offer.seller), "only shareholders can sell");
-    require(sharesOf(offer.seller).subtract(sharesOnSaleOf(offer.seller))
-        .compareTo(offer.sharesOnSale) >= 0, "not enough shares to sell");
+    require(offer.seller == caller(), 
+            "not authorized to sell");
+    require(shares.containsKey(offer.seller), 
+            "only shareholders can sell");
+    require(sharesOf(offer.seller)
+        .subtract(sharesOnSaleOf(offer.seller))
+        .compareTo(offer.sharesOnSale) >= 0, 
+        "not enough shares to sell");
     offers.add(offer);
@@
   public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "only the future owner can buy the shares");
+    require(caller() == buyer, 
+            "only the future owner can buy the shares");
     require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the sale offer is not ongoing anymore");
-    require(offer.cost.compareTo(amount) <= 0, "not enough money");
+    require(offer.isOngoing(), 
+            "the sale offer is not ongoing anymore");
+    require(offer.cost.compareTo(amount) <= 0, 
+            "not enough money");
     offers.remove(offer);
@@
     return offers.stream()
-      .filter(offer -> offer.seller == shareholder && offer.isOngoing())
+      .filter(offer -> offer.seller == shareholder && 
+              offer.isOngoing())
       .map(offer -> offer.sharesOnSale)

### 2021-07-19 040_solution.tex
Small edits
new latex style and review of introduction

@@
   public TendermintValidators
-        (TendermintED25519validator[] validators, BigInteger[] powers) {
+        (TendermintED25519validator[] validators, 
+         BigInteger[] powers) {
     super(validators, powers);
@@
   @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, TendermintED25519Validator buyer, Offer<TendermintED25519Validator> offer) {
+  public void accept
+    (BigInteger amount, 
+     TendermintED25519Validator buyer, 
+     Offer<TendermintED25519Validator> offer) {
     super.accept(amount, buyer, offer);
@@
 The fixed code is shown in Fig.~\ref{fig:solution}. The only difference is that method
-\<accept> has been redefined to force the correct type for \<buyer>. For the rest, that method
+\<accept> has been redefined to enforce the correct type for \<buyer>. For the rest, that method
 delegates to its implementation in \<AbstractValidators>.
@@
 
-  public void accept(BigInteger,TendermintED25519Validator,Offer)
+  public void accept
+   (BigInteger,TendermintED25519Validator,Offer)
    aload_0
@@
    aload_3
-   invokespecial AbstractValidators.accept(BigInteger,PayableContract,Offer)
+   invokespecial AbstractValidators
+     .accept(BigInteger,PayableContract,Offer)
    return
@@
    aload_3
-   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
+   invokevirtual accept
+      (BigInteger,TendermintED25519Validator,Offer)
    return

### 2021-07-19 050_related_work.tex
new latex style and review of introduction

@@
 
-\todo{Forse questa parte va spostata.}
+% \todo{Forse questa parte va spostata.}
+% 
+% 
+% \textbf{Compilation of generics --}
+% Generics is a way for writing code parametrized by type, in other words 
+% a same piece of code can be written by using a type placehorder 
+% (a \emph{type parameter}) and then instantiated for specific,
+% concrete types. There exists two common ways to implement generics in 
+% a programming language, that are often described in literature~\cite{generics_categories}
+%  as \emph{heterogeneous} and \emph{homogeneous}. In the heterogeneous 
+% approach the code is specialized for each instance
+% of the generic parameter; this is the approach adopted by C++ \emph{templates}.
+% Conversely, the homogeneous approach is the one provided by Java and .Net; 
+% in this case, only one instance of the code is maintained and shared by 
+% all generic instances.
+% This implementation is based on the type \emph{erasure} mechanism, where the 
+% generic parameter is replaced by the upwards bound of each instance, most 
+% often \<Object>.
+% Even if the heterogeneous approach is the safest, it is rarely applied in particular
+% in resource-constrained applications, because the code size may dramatically increase
+% as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+% the heterogeneous approach obliges one to reinstall all instantiations of the 
+% generic code, with extra costs of gas.
+% Conversely, the homogeneous approach ensures a smaller consumption
+% of resources but can lead to the problems described in this paper. \\
+% 
+% \textbf{Smart contracts vulnerabilities --}
 
-\textbf{Compilation of generics --}
-Generics is a way for writing code parametrized by type, in other words a same piece of code can
-be written by using a type placehorder (a \emph{type parameter}) and then instantiated for specific,
-concrete types. There exists two common ways to implement generics in a programming language,
-that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
-and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
-of the generic parameter; this is the approach adopted by C++ \emph{templates}.
-Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
-only one instance of the code is maintained and shared by all generic instances.
-This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
-is replaced by the upwards bound of each instance, most often \<Object>.
-Even if the heterogeneous approach is the safest, it is rarely applied in particular
-in resource-constrained applications, because the code size may dramatically increase
-as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
-the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
-with extra costs of gas.
-Conversely, the homogeneous approach ensures a smaller consumption
-of resources but can lead to the problems described in this paper. \\
-
-\textbf{Smart contracts vulnerabilities --}
 It has been estimated that, on average, software developers make from 100 to 150 errors

### 2021-07-21 010_introduction.tex
bibliography style and removed red text

@@
 Solidity's code is organized in \emph{smart contracts}, that can be seen as
-objects that control money.
-\todo{A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.}
+objects that control money. A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.
 Ethereum's transactions can hence execute much more than coin transfers. Namely,
@@
 they are just numerical \emph{addresses}.
-It follows that \todo{inside the \<child> method, an eventual} call to a \<Person>'s method
+It follows that inside the \<child> method, an eventual call to a \<Person>'s method
 on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
@@
 type variables. Generics are a powerful and very useful facility for programming
-smart contracts, \todo{since they allow one to personalize the behaviour of such contracts and partially overcome the their inherent incompleteness\,\cite{ebp}}.
+smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome the their inherent incompleteness\,\cite{ebp}.
 In Java source code, generics are strongly typed, if no \emph{unchecked operations}
@@
 of Java generics can lead to a code security vulnerability which
-allows an attacker to earn money by exploiting someone else's work, \todo{with both economical and legal side effects}.
+allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
 This paper will provide a fix to that specific issue,
-by \todo{proposing a re-engineering of the code that forces} the compiler to generate defensive checks.
+by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
 More generally, this paper can be useful for the definition of

### 2021-07-29 001_abstract.tex
Added omments on our solution

@@
 In many programming languages, such as Java, they are implemented by
-\emph{erasure}\ie replaced by their upper bound type during bytecode compilation.
+\emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
 %
@@
 %
-The considered case study is particular important since
+The considered case study is particularly important since
 a special case of shared entities could be the validators' set of
-the blockchain itself. The analysis shows that, the power of generics comes at the risk of a too permissive
+the blockchain itself. The analysis shows that the power of generics comes at the risk of a too permissive
 typing of the compiled code, due to the erasure mechanism, with a consequent possible attack

### 2021-07-29 040_solution.tex
Added omments on our solution

@@
 
-\todo{Si potrebbe dire che la soluzione e' simile ad una gestione dei generics eterogenea?
-che pero' deve essere fatta a mano ridefinendo i metodi, invece che essere generata in automatico dal compilatore?
-}
+The solution of redefining method \<accept> can be seen as a limited form of heterogeneous
+compilation of generics, restricted to a specific method and forced manually, rather than
+automatically performed by the compiler. It is interesting to consider which methods
+would need that redefinition, in general. They are those that have a parameter of generic type and
+that generic type is restricted in the subclass. For instance, method \<accept> in
+Fig.~\ref{fig:shared_entity} has parameters \<buyer> and \<offer> of generic type
+\<S> and \<O>, respectively. The subclass in Fig.~\ref{fig:abstract_validators} restricts
+\<S> to be a \<TendermintED25519Validator> and \<O> to be
+an {\codesize\texttt{Offer<TendermintED25519Validator>}}. Hence we must redefine
+\<accept> in the subclass with the more specific types for the \<buyer> and \<offer>
+parameters. In general, a compiler might perform this automatically or a static analysis
+tool might issue a warning when such redefinition is needed. Currently, however, this is
+left to the programmer, which might be the origin of security issues, as shown
+in Sec.~\ref{sec:attack}.

### 2021-07-29 010_introduction.tex

@@
 
-There exists two common ways to implement generics in a programming language,
-that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
-and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
-of the generic parameter; this is the approach adopted by C++ \emph{templates}.
-Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
-only one instance of the code is maintained and shared by all generic instances.
-This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
-is replaced by the upwards bound of each instance, mostly often \<Object>.
-Even if the heterogeneous approach is the safest, it is rarely applied in particular
-in resource-constrained applications, because the code size may dramatically increase
-as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
-the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
-with extra costs of gas.
-Conversely, the homogeneous approach ensures a smaller consumption
-of resources but it can introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity, as we will describe in this paper.
-
 The contribution of this paper is to show a real-life
@@
 The remainder of this paper is organized as follows.
+Sec..~\ref{sec:java_generics} discusses the management of generics in Java.
 Sec.~\ref{sec:shared_entities} introduces our real-life Java smart

### 2021-07-29 015_generics_in_java.tex

\section{Generics Management in Java}\label{sec:java_generics}

There exists two common ways to implement generics in a programming language,
that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
of the generic parameter; this is the approach adopted by C++ \emph{templates}.
Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
only one instance of the code is maintained and shared by all generic instances.
This implementation is based on the type \emph{erasure} mechanism, where the generic parameter
is replaced by the upwards bound of each instance, mostly often \<Object>.
Even if the heterogeneous approach is the safest, it is rarely applied in particular
in resource-constrained applications, because the code size may dramatically increase
as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
with extra costs of gas.
Conversely, the homogeneous approach ensures a smaller consumption
of resources but it can introduce the same kind of attack to the bytecode
as normal reference types do in Solidity, as we will describe in this paper.

### 2021-07-29 010_introduction.tex
Final revision

@@
 Another reason is the availability of modern
-language features, that are missed in Solidity,
+language features, that are missing in Solidity,
 such as \emph{generics}\ie the possibility of using
 type variables. Generics are a powerful and very useful facility for programming
-smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome the their inherent incompleteness\,\cite{ebp}.
+smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness\,\cite{ebp}.
 In Java source code, generics are strongly typed, if no \emph{unchecked operations}
@@
 Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contracts that uses generics. Sec.~\ref{sec:attack} shows that a na\"{i}ve
-deployment of that contract leads to a code vulnerability.
-Sec.~\ref{sec:fix} shows a fix to that vulnerability.
+contract that uses generics. Sec.~\ref{sec:attack} shows that a na\"{i}ve
+deployment of a subclass of that contract leads to a code vulnerability.
+Sec.~\ref{sec:fix} presents a fix to that vulnerability.
 Sec.~\ref{sec:related_work} discusses some related work and, finally,

### 2021-07-29 015_generics_in_java.tex
Final revision

@@
 There exists two common ways to implement generics in a programming language,
-that are often described in literature~\cite{generics_categories} as \emph{heterogeneous}
-and \emph{homogeneous}. In the heterogeneous approach the code is specialized for each instance
-of the generic parameter; this is the approach adopted by C++ \emph{templates}.
-Conversely, the homogeneous approach is the one provided by Java and .Net; in this case,
+that are often described in literature as \emph{heterogeneous}
+and \emph{homogeneous}~\cite{generics_categories}.
+In the heterogeneous approach, the code is duplicated and specialized for each instance
+of the generic parameters; this is the approach adopted by C++ \emph{templates}.
+Conversely, the homogeneous approach is that provided by Java and .Net; in this case,
 only one instance of the code is maintained and shared by all generic instances.
@@
 in resource-constrained applications, because the code size may dramatically increase
-as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+because of duplicated~\cite{generics_embedded_systems}. For code in blockchain,
 the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
-with extra costs of gas.
-Conversely, the homogeneous approach ensures a smaller consumption
-of resources but it can introduce the same kind of attack to the bytecode
-as normal reference types do in Solidity, as we will describe in this paper.
+with extra costs of gas, which makes it impractical.
+Conversely, the homogeneous approach ensures a smaller consumption of resources.
+
+In order to understand the mechanism of erasure, consider for instance
+the interface \<SharedEntity> in Fig.~\ref{fig:shared_entity} and its method \<accept>.
+The functionality of \<SharedEntity> will be discussed later (Sec.~\ref{sec:shared_entities}).
+Here, we only consider how its generic type parameters get compiled.
+Namely, \<SharedEntity> uses two generic type parameters \<S> and \<O>, that must be provided
+whenever a client creates a concrete implementation of the interface.
+Such generic parameters have an upper bound: \<S> can only be a subtype of \<PayableContract>,
+while \<O> can only be a subtype of {\codesize\texttt{Offer<S>}}. If we check the bytecode
+generated for \<SharedEntity>, we will see that \<accept> is declared, in bytecode, as
+\<void accept(BigInteger amount, PayableContract buyer, Offer offer)>, that is,
+the two type variables \<S> and \<O> have been \emph{erased} and replaced with their
+respective upper bound.
+
+Erasure weakens the type information of the compiled code. It is the responsibility of the
+compiler to guarantee that types are still respected, in all implementations of \<SharedEntity>.
+In Java, the compiler guarantees type correctness and the Java language remains strongly-typed,
+also in the presence of generic types,
+if no \emph{unchecked operations} are performed~\cite{NaftalinW06} (such as casts to generic types,
+that are unchecked for a limitation of the Java bytecode).
+However, this guarantee applies to Java source code compiled by the Java compiler,
+not to bytecode that can be generated manually, in order to attack class \<SharedEntity>,
+as we will show later.

### 2021-07-29 020_shared_entities_with_generics.tex
Final revision

@@
 
-A \emph{shared entity} is something divided into \emph{shares}. Participants,
+A \emph{shared entity} is concept that often arises in blockchain
+applications. Namely, a it is something divided into \emph{shares}. Participants,
 that hold shares, are called \emph{shareholders} and can dynamically
@@
 who are the potential shareholders and how offers for selling shares work.
-Therefore, we have parameterized the interface of a shared entity with two type variables:
+Therefore, one can parameterize the interface of a shared entity with two type variables:
 \<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
-Fig.~\ref{fig:shared_entity} shows a simplification of our interface.
+
+We are developing for the Hotmoka blockchain, that uses smart contracts written in a subset
+of Java called Takamaka.
+Fig.~\ref{fig:shared_entity} shows a simplification of our interface for shared entities.
 It includes an inner class \<Offer> that models sale offers:
@@
 This means that shareholders must be able to receive payments and that
-is why \<S extends PayableContract>: the \<PayableContract>s are those that can receive
-payments in Takamaka.
+is why \<S extends PayableContract>: the \<PayableContract>s, in Takamaka, are those that can receive
+payments.
 Method \<getShareholders> yields the list of the current shareholders of the entity.
-It is not \<@View>, since it creates a new list.
+It is not \<@View>, since it creates a new list, which is a side-effect.
 
@@
 as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Due to this limitation,
+in \<S.class>. Due to this syntactical limitation of Java,
 the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
@@
 the shareholders: if we create a \<SharedEntity> and fix a specific type \<S>
-for its shareholders, then only instances of \<S> will manage to become shareholders.
+for its shareholders, then only instances of \<S> will actually manage to become shareholders.
 
@@
 caller to have type \<S>, since the syntax of the language does not allow one to express it,
-as we explained before. Then \<accept> requires the \<offer> to exists, to be still ongoing
+as we explained before. Then \<accept> requires the \<offer> to exist, to be still ongoing
 and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,
@@
 
-It turns out that the \emph{Consistency of Shareholders} property holds if Java code
+It turns out that the \emph{Consistency of Shareholders} property holds if Java source code
 creates and populates \<SimpleSharedEntity>s.
@@
 Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-non-strongly-typed. Therefore, also contracts not of type \<S> could call \<accept>
+non-strongly-typed. In that case, also contracts not of type \<S> could call \<accept>
 and become shareholders.
@@
 in the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing \emph{of source code} in Java. But what is installed
-and executed in blockchain is the Java bytecode which has been derived from
+operations guarantees strong typing of Java \emph{source} code. But what is installed
+and executed in blockchain is the Java bytecode that has been derived from
 the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
-Malicious users might install in blockchain some crafted bytecode that
-calls the methods of \<SimpleSharedEntity>s in order to attack
-the contract. In particular,
+Malicious users might install in blockchain some manually crafted bytecode,
+not derived from its Java source code compiled together with the source code
+in Fig.~\ref{fig:simple_shared_entity}.
+That crafted code might
+call the methods of \<SimpleSharedEntity>s in order to attack
+that contract. In particular,
 the signature of method \<accept> declares a parameter \<buyer> of type \<S> at source code level, but
-its compilation into Java bytecode declares a parameter \<buyer> of type \<PayableContract> instead.
-This is because of the so called \emph{erasure} of generic types in the compilation of Java.
+its compilation into Java bytecode declares an erased
+parameter \<buyer> of type \<PayableContract> instead.
 It follows that an attacker can install in blockchain a snippet of bytecode that calls

### 2021-07-29 030_generics_attack.tex
new latex style and review of introduction
Final revision

@@
 consensus, which means that a selected dynamic subset of the nodes is in charge of
-validating the transactions and voting their acceptance. In general, validator nodes are
-represented by \<Validator> objects, that are externally owned accounts
-with an extra identifier. In the specific case of Tendermint,
+validating the transactions and voting their acceptance. In general, Hotmoka
+models validator nodes as \<Validator> objects, that are externally owned accounts
+with an extra identifier. In the specific case of a Hotmoka blockchain built over Tendermint,
 validators are \<TendermintED25519Validator> objects whose
@@
 Tendermint applications can implement their own
-policy for changing the validator set dynamically.
-The \<AbstractValidators> class implements the validator set
+policy for rewarding or changing the validators' set dynamically.
+The \<AbstractValidators> class implements the validators' set
 and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
@@
 in order to reward the validators that behaved correctly
-and punish those that misbehaved. They are specified by two strings
+and punish those that misbehaved, possibly removing them from the validators' set.
+They are specified by two strings
 that contain the identifiers of the validators, as provided by the underlying
@@
     \begin{lstlisting}[language=Takamaka]
-public abstract class Validator extends ExternallyOwnedAccount {
+public abstract class Validator 
+       extends ExternallyOwnedAccount {
 
@@
 
-public final class TendermintED25519Validator extends Validator {
+public final class TendermintED25519Validator 
+       extends Validator {
 
@@
   \end{center}
-  \caption{The representation of a validator of a Tendermint blockchain.
+  \caption{The representation, in Hotmoka, of a validator of a Hotmoka blockchain based on Tendermint.
   Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintED25519Validator.java}.}\label{fig:validator}
@@
     \begin{lstlisting}[language=Takamaka]
-public abstract class AbstractValidators<V extends Validator>
+public abstract class 
+    AbstractValidators<V extends Validator>
     extends SimpleSharedEntity<V, Offer<V>> {
 
-  public AbstractValidators(V[] validators, BigInteger[] powers) {
+  public AbstractValidators
+    (V[] validators, BigInteger[] powers) {
     super(validators, powers);
@@
   public void reward
-    (String behaving, String misbehaving, BigInteger gasConsumed) { ... }
+    (String behaving, String misbehaving, 
+     BigInteger gasConsumed) { ... }
 }
@@
 Tendermint expects such validators to mine and vote the subsequent blocks, until a change in the
-validators set occurs.
+validators' set occurs.
 
@@
 and their Tendermint identifier, that identifies which node of the blockchain
-must do the validation work (Fig.~\ref{fig:validator}).
+must do the validation work (see how the constructor initializes \<this.id> in
+Fig.~\ref{fig:validator}).
 If it were possible to add a shareholder of another
-type \<Attacker>, the latter could decouple the node identifier from its
+type \<Attacker>, the code of \<Attacker> could decouple the node identifier from its
 public key (see Fig.~\ref{fig:attacker}):
@@
 the validation work while the owner
-of the private key of the \<Attacker> could just wait for accrued awards to spend.
+of the private key of the \<Attacker> could just wait for accrued rewards to spend.
 A sort of validator's slavery.
@@
   \end{center}
-  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the rewards of its work.}\label{fig:attacker}
+  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the rewards of the work of that node.}\label{fig:attacker}
 \end{figure}

### 2021-07-29 040_solution.tex
Final revision

@@
 over-permissive erasure of the signature of method \<accept>,
-where \<buyer> has type \<PayableContract>.
+where \<buyer> is given type \<PayableContract>.
 Therefore, a solution is to oblige the compiler to generate a more
@@
 has type \<TendermintED25519Validator>: only that type of accounts
-must be accepted for the validators.
+must be accepted for the validators, consequently banning instances of \<Attacker>.
 
@@
 \<accept> has been redefined to enforce the correct type for \<buyer>. For the rest, that method
-delegates to its implementation in \<AbstractValidators>.
+delegates to its implementation inherited from \<AbstractValidators>, through a call
+to \<super.accept>.
 It is important to investigate which is the Java bytecode generated from
@@
 The solution of redefining method \<accept> can be seen as a limited form of heterogeneous
-compilation of generics, restricted to a specific method and forced manually, rather than
-automatically performed by the compiler. It is interesting to consider which methods
-would need that redefinition, in general. They are those that have a parameter of generic type and
-that generic type is restricted in the subclass. For instance, method \<accept> in
+compilation of generics, restricted to a specific method and forced manually.
+It is interesting to consider which methods
+would need that redefinition, in general. They are those that have a parameter of a generic type
+that is restricted in a subclass. For instance, method \<accept> in
 Fig.~\ref{fig:shared_entity} has parameters \<buyer> and \<offer> of generic type
@@
 \<accept> in the subclass with the more specific types for the \<buyer> and \<offer>
-parameters. In general, a compiler might perform this automatically or a static analysis
+parameters. In the future, a compiler might perform this automatically or a static analysis
 tool might issue a warning when such redefinition is needed. Currently, however, this is
-left to the programmer, which might be the origin of security issues, as shown
-in Sec.~\ref{sec:attack}.
+left to the programmer of the smart contracts, who might overlook the issue
+and give rise to security issues, as shown in Sec.~\ref{sec:attack}.

### 2021-07-29 050_related_work.tex
Final revision

@@
 
-Given the plurality of verification tools that have been developed for indentifying security bugs
-in smart contracts, a unified and standard approach for evaluating their accuracy and performances
-have been developed in~\cite{effectiveness_analysis_tools}. This automated and systematic approach,
-called SolidiFI, consists in injecting bugs into all potential locations in a smart contract,
-in order to introduce different kinds of vulnerabilities, and then using several static analysis
-tools to identify which bugs each of them is able to detect.
+%Given the plurality of verification tools that have been developed for indentifying security bugs
+%in smart contracts, a unified and standard approach for evaluating their accuracy and performances
+%have been developed in~\cite{effectiveness_analysis_tools}. This automated and systematic approach,
+%called SolidiFI, consists in injecting bugs into all potential locations in a smart contract,
+%in order to introduce different kinds of vulnerabilities, and then using several static analysis
+%tools to identify which bugs each of them is able to detect.
 
@@
 the Neo Savant online IDE. Currently, neither Neo Savant IDE nor the block explorer
-allow one to inspect the compiled bytecode.
+allow one to inspect the compiled bytecode, in order to undertsnad how generic types are compiled.

### 2021-07-29 090_conclusion.tex
Final revision

@@
 programming languages, including Java, they get erased at compile time
-into types that might be too permissive for low level calls, such as
+into types that might be too permissive for low-level calls, such as
 those that start blockchain transactions. Note that using a programming
@@
 languages have no direct low-level
-counterpart but are implemented via instrumentation, starting from
-inner classes and closures. This is fine at source level, but allows
+counterpart but are implemented via instrumentation, from
+inner classes to closures. This is fine at source level, but allows
 low-level calls to easily circumvent

### 2021-09-23 015_generics_in_java.tex
Applied comments by BCCA2021 reviewers

@@
 in resource-constrained applications, because the code size may dramatically increase
-because of duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+because of duplication~\cite{generics_embedded_systems}. For code in blockchain,
 the heterogeneous approach obliges one to reinstall all instantiations of the generic code,

### 2021-09-23 040_solution.tex
Applied comments by BCCA2021 reviewers

@@
 \noindent
-The first \<accept> method is the compilation of that from Fig.~\ref{fig:solution}:
+The first \<accept> method above is the compilation of that from Fig.~\ref{fig:solution}:
 it delegates to the \<accept> method of the superclass \<AbstractValidators>. The second
-\<accept> method is a \emph{bridge method} that the compiler generates in order to guarantee
+\<accept> method above is a \emph{bridge method} that the compiler generates in order to guarantee
 that all calls to the erased signature \<accept(BigInteger,PayableContract,Offer)> actually

### 2021-09-23 050_related_work.tex
Applied comments by BCCA2021 reviewers

@@
 a rule-based representation of the bytecode.
+%
+Considering the specific case of the Java language,
+formal techniques for static analysis can be built, for instance, over
+the Featherweight Java calculus~\cite{IgarashiPW01}, or by abstract
+interpretation~\cite{Spoto16}. Currently, however, we are not aware
+of formal verifications for generics, at bytecode level.
 

### 2021-09-27 020_shared_entities_with_generics.tex
minor_camera_ready_changes

@@
   \caption{A simplified part of our shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}.}\label{fig:shared_entity}
+  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}}.}\label{fig:shared_entity}
 \end{figure}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[t]
   \begin{center}
@@
   \caption{A simplified part of our implementation of the shared entity interface.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}.}\label{fig:simple_shared_entity}
+  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}}.}\label{fig:simple_shared_entity}
 \end{figure}

### 2021-09-27 040_solution.tex
minor_camera_ready_changes

@@
   \caption{The shared entity of the validators of a Hotmoka blockchain built over Tendermint.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}.}\label{fig:solution}
+  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}}.}\label{fig:solution}
 \end{figure}

### 2022-02-03 bcca_2021/010_introduction.tex
directory refactoring

--- 2021-07-29 010_introduction.tex
@@
 type variables. Generics are a powerful and very useful facility for programming
-smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness\,\cite{ebp}.
+smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}.
 In Java source code, generics are strongly typed, if no \emph{unchecked operations}
 are used~\cite{NaftalinW06}, as it will always be the case in this paper.
+However, generics might have security issue at the level of compiled Java code and this paper originated from a
+real issue that we found in our code.
 

### 2022-02-03 bcca_2021/030_generics_attack.tex
directory refactoring

--- 2021-07-29 030_generics_attack.tex
@@
   \caption{The representation, in Hotmoka, of a validator of a Hotmoka blockchain based on Tendermint.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintED25519Validator.java}.}\label{fig:validator}
+  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintED25519Validator.java}}.}\label{fig:validator}
 \end{figure}
@@
   \caption{The shared entity of the validators of a Hotmoka blockchain.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/AbstractValidators.java}.}\label{fig:abstract_validators}
+  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/AbstractValidators.java}}.}\label{fig:abstract_validators}
 \end{figure}

### 2022-02-03 cluster_computing/001_abstract.tex
cluster computing package

--- 2021-07-29 001_abstract.tex
@@
 %
-This paper investigates the use of generics in the
-context of smart contracts for blockchain, in order to implement
+This paper originated from a real security issue that we found
+while using generics for writing
+smart contracts for blockchain, in order to implement
 a contract for \emph{shared entities} (such as a company shared
 by its shareholders),
-by using the Hotmoka blockchain whose contracts are written in Java.
+for the Hotmoka blockchain, whose contracts are written in Java.
 %
 The considered case study is particularly important since
-a special case of shared entities could be the validators' set of
-the blockchain itself. The analysis shows that the power of generics comes at the risk of a too permissive
+the validators' set of the blockchain itself is
+a special case of shared entities.
+The analysis shows that the power of generics comes at the risk of a too permissive
 typing of the compiled code, due to the erasure mechanism, with a consequent possible attack

### 2022-02-03 cluster_computing/010_introduction.tex
cluster computing package

--- 2022-02-03 bcca_2021/010_introduction.tex
@@
 
-\IEEEPARstart{B}{lockchains} exploit the redundant, concurrent execution of the same
+Blockchains exploit the redundant, concurrent execution of the same
 transactions on a decentralized network of many machines,

### 2022-02-03 cluster_computing/050_related_work.tex
cluster computing package

--- 2021-09-23 050_related_work.tex
@@
 the Neo Savant online IDE. Currently, neither Neo Savant IDE nor the block explorer
-allow one to inspect the compiled bytecode, in order to undertsnad how generic types are compiled.
+allow one to inspect the compiled bytecode, in order to understand how generic types are compiled.

### 2022-02-15 cluster_computing/017_takamaka.tex
Started section of Takamaka

\section{The Takamaka Language for Smart Contracts}\label{sec:takamaka}

This section gives a short introduction to the Takamaka subset of
Java that we use for writing smart contracts in this paper.
This language has been introduced in~\cite{Spoto19}. A full tutorial
is available online, as part of the documentation of the Hotmoka
blockchain that runs smart contracts written
in Takamaka~\cite{hotmoka_repository}. This section introduces only
the essential notions that are needed to understand the subsequent sections.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=12.2cm]{entities-hierarchy.png}
  \end{center}
  \caption{The hierarchy of Takamaka classes that implement accounts, shared entities and validators. Their source code can be found inside the Java project \textsf{\url{https://github.com/Hotmoka/hotmoka/tree/master/io-takamaka-code}}.}\label{fig:hierarchy-entities}
\end{figure}

### 2022-02-15 cluster_computing/017_takamaka.tex
Started describing accounts

@@
 the essential notions that are needed to understand the subsequent sections.
+The hierarchy of the classes described in this section is in Fig.~\ref{fig:hierarchy-entities}.
+In the following we will report a simplified presentation of the code of some
+of such classes. The full code is in the Github repository of Hotmoka.
 
@@
 \end{figure}
+
+Takamaka implements objects persisted in blockchain as subclasses of the
+\<io.takamaka.code.lang.Storage> class.
+This is the main difference with other attemps at using Java for writing smart contracts:
+the programmer does not code the serialization and deserialization of objects into
+a \emph{keeper} or a key/value map, but simply extends \<Storage> and objects get
+persisted automatically \emph{out of magic}. In this sense, Takamaka follows the apporach of
+Solidity, but using Java.
+
+The \<io.takamaka.code.lang.Contract> class implements objects that can be
+persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
+Their balance is available through a \<balance> method. This is a \<@View> method, meaning
+that it can be called without paying gas (the measure of execution cost), since such methods
+cannot have side-effects and consequently do not modify the storage of the blockchain.
+Payments can be received only through methods annotated as \<@Payable>.
+The \<Contract> superclass has no such methods, but subclasses may have.
+For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
+a method \<receive> to receive payments from its caller. Many methods (inclusing all
+\<@Payable> methods) need to identify their caller. This is done by adding the
+\<@FromContract> annotation, that guarantees that the caller is a contract, that is available
+inside the method as \<caller()>.
+
+Method calls started from outside the blockchain (for instance, from a client such as a wallet
+or from a web application), must specify an already existing \<ExternallyOwnedAccount>
+as caller. This account will pay for the gas of the execution. The blockchain will accept
+the call only if is signed with the private key that matches the public key provided to
+the constructor of the account when it was created. Method \<publicKey> allows one to
+recover that public key and method \<nonce> allows one to get a progressive identifier that
+can be used to distinguish successive calls with the same account, to force their order
+of execution and to avoid replaying.
+All that is very similar to Solidity, except for the fact that externally owned accounts
+are actual Java objects inside the blockchain, not just an abstraction of a public key.
+
+Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
+Both proof of work and proof of stacke can be used, for instance. In particular, if proof of stake
+is used, then each validator node of the blockchain must specify a
+\<io.takamaka.code.governance.Validator> object, that plays the role of the
+banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
+It is a special externally owned account, with an extra \<id> method that provides the
+identifier of the validator node inside the blockchain network. This identifier depends from the
+specific blockchain implementation.
+For instance, the subclass
+\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
+implements \<id> as for the Tendermint blockchain engine, that is,
+as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class Validator extends ExternallyOwnedAccount {
+
+  public Validator(String publicKey) {
+    super(publicKey);
+  }
+
+  public @View String id() {
+    return publicKey(); // subclasses may redefine
+  }
+}
+
+public final class TendermintED25519Validator extends Validator {
+  private final String id;
+
+  public TendermintED25519Validator(String publicKey) {
+    super(publicKey);
+
+    MessageDigest sha256 = MessageDigest.getInstance("SHA-256");
+    sha256.update(Base64.getDecoder().decode(publicKey()));
+    this.id = bytesToHex(sha256.digest()).substring(0, 40);
+  }
+
+  @Override public @View final String id() {
+    return id;
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{The account of a validator and its specialization for a Hotmoka blockchain based on Tendermint.}\label{fig:validator}
+\end{figure}

### 2022-02-15 cluster_computing/030_generics_attack.tex
Started describing accounts

--- 2022-02-03 bcca_2021/030_generics_attack.tex
@@
 Tendermint engine.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public abstract class Validator 
-       extends ExternallyOwnedAccount {
-
-  public Validator(String publicKey) {
-    super(publicKey);
-  }
-
-  @View
-  public abstract String id();
-}
-
-public final class TendermintED25519Validator 
-       extends Validator {
-
-  // the first 40 characters of the hexadecimal representation
-  // of the sha256 hashing of the public key bytes
-  private final String id;
-
-  public TendermintED25519Validator(String publicKey) {
-    super(publicKey);
-    this.id = /* derived from publicKey as in the comment for id above */
-  }
-
-  @Override @View
-  public final String id() {
-    return id;
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{The representation, in Hotmoka, of a validator of a Hotmoka blockchain based on Tendermint.
-  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintED25519Validator.java}}.}\label{fig:validator}
-\end{figure}
 

### 2022-02-15 cluster_computing/020_shared_entities_with_generics.tex
cluster computing package
Edited the shared entity description

--- 2021-09-27 020_shared_entities_with_generics.tex
@@
 A \emph{shared entity} is concept that often arises in blockchain
-applications. Namely, a it is something divided into \emph{shares}. Participants,
+applications. Namely, a shared entity is something divided into \emph{shares}. Participants,
 that hold shares, are called \emph{shareholders} and can dynamically
@@
     \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract,
-                              O extends Offer<S>> {
+public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
+      extends SharedEntityView<S> {
 
   @View
-  BigInteger sharesOf(S shareholder);
+  StorageSetView<O> getOffers();
+
+  @View
+  BigInteger sharesOnSaleOf(S shareholder);
 
   @FromContract(PayableContract.class) @Payable
-  void place(BigInteger amount, O offer);
+  void place(BigInteger ticket, O offer);
 
   @FromContract(PayableContract.class) @Payable
-  void accept(BigInteger amount, S buyer, O offer);
-
-  List<S> getShareholders();
+  void accept(BigInteger ticket, S buyer, O offer);
 
@@
   \end{center}
-  \caption{A simplified part of our shared entity interface.
-  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SharedEntity.java}}.}\label{fig:shared_entity}
+  \caption{A simplified part of our shared entity interface.}\label{fig:shared_entity}
 \end{figure}
@@
 Therefore, one can parameterize the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the offers for selling shares.
+\<S> is the type of the shareholders and \<O> is the type of the sale offers of shares.
 
-We are developing for the Hotmoka blockchain, that uses smart contracts written in a subset
-of Java called Takamaka.
-Fig.~\ref{fig:shared_entity} shows a simplification of our interface for shared entities.
+The \<SharedEntityView> interface at the top of the hierarchy in Fig.~\ref{fig:hierarchy-entities}
+defines the read-only operations on a shared entity. This view is \emph{static}, in the sense that it
+does not specify the operations for transfers of shares. Therefore, its only type parameter is \<S>:
+any contract can play the role of the type for the shareholders of the entity.
+Method \<getShares> yields a snapshot of the
+current shares of the entity (who owns how much). Method \<getShareholders> yields the shareholders.
+It is not \<@View>, since it creates a new stream, which is a side-effect.
+Method \<isShareholder> checks if an object is a shareholder. Method \<sharesOf> yields
+the number of shares of a shareholder. As typical in Takamaka, a \<snapshot> method allows one
+to create a frozen read-only copy of an entity (in constant time), useful when an enity must be queried from
+a client without the risk of race conditions if another client is modifying the same entity concurrently.
+
+The \<SharedEntity> subinterface adds methods for transfer of shares
+(see Fig.~\ref{fig:shared_entity}).
 It includes an inner class \<Offer> that models sale offers:
@@
 Implementations can subclass \<Offer> if they need more specific offers.
-By using class \<Offer>, the \<SharedEntity> interface specifies four methods.
-Method \<sharesOf> allows one to know how many shares a potential \<shareholder> (of type \<S>) holds.
-It is annotated as \<@View>. In Takamaka, this means that its execution can be performed
-\emph{for free}, without paying gas, since it has no side-effects. Who wants to sell
-(some of) its shares calls method \<place> with a sale \<offer>.
-This method is annotated as \<@Payable> since
-implementations are allowed to require a payment of \<amount> coins for managing the sale.
-Who buys the shares calls method \<accept> with the accepted \<offer>
+Offers can be placed on sale
+by calling the \<place> method with a sale \<offer>.
+This method is annotated as \<@FromContract> since the caller must be
+identified (or otherwise anybody could sell the shares of anybody else) and
+as \<@Payable> so that implementations can require
+to pay a \<ticket> to place shares on sale.
+The sale offer is passed as a parameter to \<place>, hence it must have been created before calling that method.
+The set of all sale offers is available through \<getOffers>. Method \<sharesOnSale> yields the
+cumulative number of shares on sale for a given shareholder.
+Who wants to buy shares calls method \<accept> with the accepted \<offer>
 and with itself as \<buyer> (the reason will be explained soon)
@@
 its cumulative number of shares (if it was a shareholder already).
-Also this method is \<@Payable>, since its caller must pay \<amount> $\ge$ \<offer.cost>
+Also this method is \<@Payable>, since its caller must pay \<ticket> $\ge$ \<offer.cost>
 coins to the seller.
 This means that shareholders must be able to receive payments and that
-is why \<S extends PayableContract>: the \<PayableContract>s, in Takamaka, are those that can receive
-payments.
-Method \<getShareholders> yields the list of the current shareholders of the entity.
-It is not \<@View>, since it creates a new list, which is a side-effect.
+is why \<S extends PayableContract>: only \<PayableContract>s are guaranteed to have a
+\<receive> method in Takamaka.
 
-The annotation \<@FromContract> on both \<place> and \<accept> enforces that only
+As said before, the annotation \<@FromContract> on both \<place> and \<accept> enforces that only
 contracts can call these methods.
 These callers must be (old or new) shareholders,
-hence they must have type \<S>. In Takamaka, this could be written
-as \<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
-in \<S.class>. Due to this syntactical limitation of Java,
+hence they must have type \<S>. Therefore, one would like to write
+\<@FromContract(S.class)>. Unfortunately, Java does not allow a generic type variable \<S>
+in the syntax \<S.class>. Due to this syntactical limitation of Java,
 the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
@@
 Let us state an important property about shared entities:
-
-\begin{tcolorbox}
+%
+\vspace{2ex}
+\begin{mdframed}[leftmargin=10pt,rightmargin=10pt]
   \begin{center}\emph{Consistency of Shareholders}\end{center}
+  \noindent
   If \<se> is a {\codesize\texttt{SharedEntity<S,O>}}
-  then the elements contained in the list \<se.getShareholders()> have type \<S>.
-\end{tcolorbox}
-
-\noindent
+  then \<se.getShareholders()> contains only elements of type \<S>.
+\end{mdframed}
+\vspace{2ex}
+%
 This property is important since it states that we can trust the type \<S> of

### 2022-02-15 cluster_computing/020_shared_entities_with_generics.tex
Reorganized sections on validators and attack

@@
 a voting community, where shares represent the voting power of each given voter.
-A further example are the validator nodes of a proof of stake blockchain,
+A further example is the set of the validator nodes of a proof of stake blockchain,
 where shares represent their voting power and remuneration percentage.
@@
     \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
+public interface SharedEntity
+      <S extends PayableContract, O extends Offer<S>>
       extends SharedEntityView<S> {
@@
 
-    public Offer
-        (S seller, BigInteger sharesOnSale, 
-         BigInteger cost, long duration) {
-
+    public Offer(S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
       this.seller = seller;
@@
 
-Let us state an important property about shared entities:
-%
-\vspace{2ex}
-\begin{mdframed}[leftmargin=10pt,rightmargin=10pt]
-  \begin{center}\emph{Consistency of Shareholders}\end{center}
-  \noindent
-  If \<se> is a {\codesize\texttt{SharedEntity<S,O>}}
-  then \<se.getShareholders()> contains only elements of type \<S>.
-\end{mdframed}
-\vspace{2ex}
-%
-This property is important since it states that we can trust the type \<S> of
-the shareholders: if we create a \<SharedEntity> and fix a specific type \<S>
-for its shareholders, then only instances of \<S> will actually manage to become shareholders.
-
 \begin{figure}[t]
@@
     <S extends PayableContract,O extends Offer<S>>
-      extends Contract
-      implements SharedEntity<S,O> {
+    extends PayableContract implements SharedEntity<S,O> {
 
-  private final StorageTreeMap<S,BigInteger> shares = 
-     new StorageTreeMap<>();
-  private final StorageSet<O> offers = 
-     new StorageTreeSet<>();
+  private StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
+  private StorageSet<O> offers = new StorageTreeSet<>();
 
   public SimpleSharedEntity
-    (S[] shareholders, BigInteger[] shares) {
-    require(shareholders.length == shares.length, 
-            "length mismatch");
-    for (int pos = 0; pos < shareholders.length; pos++)
-      addShares(shareholders[pos], shares[pos]);
+      (S shareholder, BigInteger share) {
+    addShares(shareholder, share);
   }
@@
   public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault
-      (shareholder, BigInteger.ZERO);
+    return shares.getOrDefault(shareholder, ZERO);
   }
@@
   public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), 
-            "not authorized to sell");
-    require(shares.containsKey(offer.seller), 
-            "only shareholders can sell");
-    require(sharesOf(offer.seller)
-        .subtract(sharesOnSaleOf(offer.seller))
-        .compareTo(offer.sharesOnSale) >= 0, 
-        "not enough shares to sell");
+    require(offer.seller == caller(), "unauthorized");
+    require(shares.containsKey(offer.seller), "unauthor.");
+    require(sharesOf(offer.seller) - sharesOnSaleOf(offer.seller) >= offer.sharesOnSale, "not enough shares");
     offers.add(offer);
@@
   public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, 
-            "only the future owner can buy the shares");
+    require(caller() == buyer, "unauthorized");
     require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), 
-            "the sale offer is not ongoing anymore");
-    require(offer.cost.compareTo(amount) <= 0, 
-            "not enough money");
+    require(offer.isOngoing(), "the offer is not ongoing");
+    require(offer.cost <= amount, "not enough money");
     offers.remove(offer);
@@
 
-  private BigInteger sharesOnSaleOf(S shareholder) {
+  @Override @View
+  public final BigInteger sharesOnSaleOf(S shareholder) {
     return offers.stream()
-      .filter(offer -> offer.seller == shareholder && 
-              offer.isOngoing())
+     .filter(o -> o.seller == shareholder && o.isOngoing())
      .map(offer -> offer.sharesOnSale)
      .reduce(ZERO, BigInteger::add);
-  }
-
-  @Override
-  public List<S> getShareholders() {
-    return shares.keyList();
   }
@@
   \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.
-  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/dao/SimpleSharedEntity.java}}.}\label{fig:simple_shared_entity}
+  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
 \end{figure}
 
-Fig.~\ref{fig:simple_shared_entity} shows our implementation of the \<SharedEntity> interface
-in Fig.~\ref{fig:shared_entity}
-by using two fields: \<shares> maps each shareholder to its amount of shares and
+Fig.~\ref{fig:simple_shared_entity} shows a portion of our \<SimpleSharedEntity>
+implementation of the \<SharedEntity> interface in Fig.~\ref{fig:shared_entity}, that
+uses two fields: \<shares> maps each shareholder to the amount of shares the it holds and
 \<offers> collects the offers that have been placed.
-The constructor initially populates the map \<shares>.
+The constructor initially populates the map \<shares> with the initial shareholder.
+Other shareholders can be added later, by buying shares\footnote{In the real code, the class has
+  constructors to create shared entities with a \emph{set} of initial shareholders. This paper
+  reports a simplification of the actual code.}.
 Method \<sharesOf> simply accesses
@@
 the seller identified in the \<offer>. This forbids shareholders to sell shares on behalf of others.
-Moreover, this guarantees that the caller has type \<S>, like \<offer.seller>.
-As we said before, this cannot be expressed with the syntax of the language.
+Moreover, this guarantees that the caller has type \<S>, the type of \<offer.seller>.
+As it has been said before, this cannot be expressed with the syntax of the language.
 Method \<place> further requires the seller to be a shareholder with at least \<offer.sharesOnSale>
@@
 receives the required price \<offer.cost>.
-Method \<getShareholders()> yields the list of the keys in the domain of map \<shares>.
-
-It turns out that the \emph{Consistency of Shareholders} property holds if Java source code
-creates and populates \<SimpleSharedEntity>s.
-Namely, the code in Fig.~\ref{fig:simple_shared_entity}
-does not use unchecked casts, hence it is strongly-typed~\cite{NaftalinW06} and
-the map \<shares> actually holds values of type \<S> in its domain, only.
-For this consistency result, we had to pay the price
-of the dummy \<buyer> argument for method \<accept>. Without that argument,
-\emph{Consistency of Shareholders} would not hold, since we could only write
-\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
-Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes the code
-non-strongly-typed. In that case, also contracts not of type \<S> could call \<accept>
-and become shareholders.
-
-There is, however, a problem with the reasoning
-in the previous paragraph. Namely, absence of unchecked
-operations guarantees strong typing of Java \emph{source} code. But what is installed
-and executed in blockchain is the Java bytecode that has been derived from
-the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
-Malicious users might install in blockchain some manually crafted bytecode,
-not derived from its Java source code compiled together with the source code
-in Fig.~\ref{fig:simple_shared_entity}.
-That crafted code might
-call the methods of \<SimpleSharedEntity>s in order to attack
-that contract. In particular,
-the signature of method \<accept> declares a parameter \<buyer> of type \<S> at source code level, but
-its compilation into Java bytecode declares an erased
-parameter \<buyer> of type \<PayableContract> instead.
-It follows that an attacker can install in blockchain a snippet of bytecode that calls
-\<accept> and passes \emph{any} \<PayableContract>, not only those that are instances of \<S>:
-the \emph{Consistency of Shareholders} property is easily violated at bytecode level.
-
-In general, the problem arises whenever a method declares a formal parameter of generic type,
-such as \<buyer> in method \<accept> of Fig.~\ref{fig:shared_entity}. Next section shows
-the actual significance of the consequent security risk.

### 2022-02-15 cluster_computing/025_validators.tex
Reorganized sections on validators and attack

--- 2022-02-15 cluster_computing/030_generics_attack.tex
@@
-\section{An Attack to the Shared Entities Contract}\label{sec:attack}
+\section{Blockchain Validators Set as a Shared Entity}\label{sec:validators}
 
-This paper originated from an actual security issue that we found in our code
-that used shared entities in order to model the validators of a blockchain.
-Namely, the Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
+The Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
 generic engine for replicating an application over a network of nodes. In our case,
@@
 consensus, which means that a selected dynamic subset of the nodes is in charge of
-validating the transactions and voting their acceptance. In general, Hotmoka
-models validator nodes as \<Validator> objects, that are externally owned accounts
+validating the transactions and voting their acceptance. As we have said,
+Hotmoka models validator nodes as \<Validator> objects, that are externally owned accounts
 with an extra identifier. In the specific case of a Hotmoka blockchain built over Tendermint,
@@
 policy for rewarding or changing the validators' set dynamically.
+
+The set of the validator nodes of a blockchain network is an example of a shared entity.
+Namely, each such validator owns an amount of validation power, that corresponds to the shares
+of a shareholder. Validation power can be sold and bought, exactly as shares.
+Consequently, the \<Validators> interface in Fig.~\ref{fig:hierarchy-entities}
+(reported in Fig.~\ref{fig:validators})
+extends the \<SharedEntity> interface, fixes the shareholders to be
+instances of \<Validator> and adds two methods: \<getStake> yields the money
+at stake for each given validator (if the validator misbehaves, its stake will be reduced or
+\emph{slashed}); and \<reward>, that is called by the blockchain itself at the end of each
+block creation: it distributes the cost of the gas consumed by the transactions of the block,
+to the well-behaving validators, and slashes the stakes of the misbehaving validators.
+
 The \<AbstractValidators> class implements the validators' set
 and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
-(see Fig.~\ref{fig:abstract_validators}).
+(see Fig.~\ref{fig:validators}).
 Shares are voting power in this case.
-Its subclass \<TendermintValidators> restricts the type \<S> of the shareholders
-to \<TendermintED25519Validator>.
-At each block committed, Hotmoka calls the \<reward> method of \<AbstractValidators>
+Its subclass \<TendermintValidators> restricts the type of the validators
+to be \<TendermintED25519Validator>.
+At each block committed, Hotmoka calls the \<reward> method of \<Validators>
 in order to reward the validators that behaved correctly
-and punish those that misbehaved, possibly removing them from the validators' set.
+and slash those that misbehaved, possibly removing them from the validators' set.
 They are specified by two strings
@@
 Tendermint engine.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public abstract class 
-    AbstractValidators<V extends Validator>
-    extends SimpleSharedEntity<V, Offer<V>> {
-
-  public AbstractValidators
-    (V[] validators, BigInteger[] powers) {
-    super(validators, powers);
-  }
-
-  public void reward
-    (String behaving, String misbehaving, 
-     BigInteger gasConsumed) { ... }
-}
-
-public class TendermintValidators
-    extends AbstractValidators<TendermintED25519Validator> {
-
-  public TendermintValidators
-        (TendermintED25519validator[] validators, 
-         BigInteger[] powers) {
-    super(validators, powers);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{The shared entity of the validators of a Hotmoka blockchain.
-  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/AbstractValidators.java}}.}\label{fig:abstract_validators}
-\end{figure}
 
@@
 
-It is important that the \emph{Consistency of Shareholders} property holds
-for \<TendermintValidators>: its shareholders
-must be \<TendermintED25519Validator>s (as declared in the generic
-signature of \<TendermintValidators>) that enforce a
-match between their public key, that identifies who can spend the rewards sent to the validator,
-and their Tendermint identifier, that identifies which node of the blockchain
-must do the validation work (see how the constructor initializes \<this.id> in
-Fig.~\ref{fig:validator}).
-If it were possible to add a shareholder of another
-type \<Attacker>, the code of \<Attacker> could decouple the node identifier from its
-public key (see Fig.~\ref{fig:attacker}):
-Tendermint would expect the node (belonging to the \emph{victim}) to do
-the validation work while the owner
-of the private key of the \<Attacker> could just wait for accrued rewards to spend.
-A sort of validator's slavery.
-Sec.~\ref{sec:shared_entities} asserted that the \emph{Consistency of Shareholders}
-property holds, at source level.
-Namely, an \<attacker> (of type \<Attacker>) can only become shareholder
-by accepting an ongoing sale \<offer> of shares through a call to
-\<tv.accept(offer.cost, attacker, offer)> (see Fig.~\ref{fig:simple_shared_entity}).
-This is impossible at source level, where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
-is not an instance of \<S>, which has been set to \<TendermintED25519Validator>.
-But a Hotmoka blockchain contains only the bytecode of \<SimpleSharedEntity>,
-where the signature of \<accept> has been erased into
-\<accept(BigInteger amount, PayableContract buyer, Offer offer)>.
-Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
-at bytecode level
-\emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
-are instances of \<PayableContract>.
-That transaction adds \<attacker> to the shareholders of \<tv>,
-therefore violating the \emph{Consistency of Shareholders} property and allowing
-validator's slavery.
-
 \begin{figure}[t]
@@
     \begin{lstlisting}[language=Takamaka]
-public class Attacker extends ExternallyOwnedAccount {
+public interface Validators<V extends Validator>
+    extends SharedEntity<V,Offer<V>> {
 
-  public Validator(String publicKey) {
-    super(publicKey);
+  @View
+  BigInteger getStake(V validator);
+
+  @FromContract @Payable
+  void reward(String behaving, String misbehaving, BigInteger gasConsumed);
 }
 
-  @View
-  public String id() {
-    return /* id of the victim blockchain node, unrelated to publicKey */
+public abstract class AbstractValidators
+    <V extends Validator>
+    extends SimpleSharedEntity<V,Offer<V>> {
+
+  public AbstractValidators
+      (V validator, BigInteger power) {
+    super(validator, power);
+  }
+
+  public void reward(String behaving, String misbehaving, BigInteger gasConsumed) { ... }
+}
+
+public class TendermintValidators
+   extends AbstractValidators<TendermintED25519Validator> {
+
+  public TendermintValidators(TendermintED25519validator validator, BigInteger power) {
+    super(validator, power);
   }
@@
   \end{center}
-  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the rewards of the work of that node.}\label{fig:attacker}
+  \caption{The shared entity of the validators set of a Hotmoka blockchain.}\label{fig:validators}
 \end{figure}

### 2022-02-15 cluster_computing/030_generics_attack.tex
Reorganized sections on validators and attack

@@
 
-This paper originated from an actual security issue that we found in our code
-that used shared entities in order to model the validators of a blockchain.
-Namely, the Hotmoka blockchain is built over Tendermint~\cite{Kwon14}, a
-generic engine for replicating an application over a network of nodes. In our case,
-the application is the executor of smart contracts in Java, such as that in
-Fig.~\ref{fig:simple_shared_entity}. Tendermint is based on a proof of stake
-consensus, which means that a selected dynamic subset of the nodes is in charge of
-validating the transactions and voting their acceptance. In general, Hotmoka
-models validator nodes as \<Validator> objects, that are externally owned accounts
-with an extra identifier. In the specific case of a Hotmoka blockchain built over Tendermint,
-validators are \<TendermintED25519Validator> objects whose
-identifier is derived from their ed25519 public key (see Fig.~\ref{fig:validator}).
-This identifier is public information, reported in the blocks or easily eavesdropped.
-Tendermint applications can implement their own
-policy for rewarding or changing the validators' set dynamically.
-The \<AbstractValidators> class implements the validators' set
-and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
-(see Fig.~\ref{fig:abstract_validators}).
-Shares are voting power in this case.
-Its subclass \<TendermintValidators> restricts the type \<S> of the shareholders
-to \<TendermintED25519Validator>.
-At each block committed, Hotmoka calls the \<reward> method of \<AbstractValidators>
-in order to reward the validators that behaved correctly
-and punish those that misbehaved, possibly removing them from the validators' set.
-They are specified by two strings
-that contain the identifiers of the validators, as provided by the underlying
-Tendermint engine.
+Let us state an important property about shared entities:
+%
+\vspace{2ex}
+\begin{mdframed}[leftmargin=10pt,rightmargin=10pt]
+  \begin{center}\emph{Consistency of Shareholders}\end{center}
+  \noindent
+  If \<se> is a {\codesize\texttt{SharedEntity<S,O>}}
+  then \<se.getShareholders()> contains only elements of type \<S>.
+\end{mdframed}
+\vspace{2ex}
+%
+This property is important since it states that we can trust the type \<S> of
+the shareholders: if one creates a \<SharedEntity> and fixes a specific type \<S>
+for its shareholders, then only instances of \<S> will actually manage to become shareholders.
 
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public abstract class 
-    AbstractValidators<V extends Validator>
-    extends SimpleSharedEntity<V, Offer<V>> {
+It turns out that the \emph{Consistency of Shareholders} property holds
+for instances of the class \<SimpleSharedEntity> in Fig.~\ref{fig:simple_shared_entity}.
+Namely, that class does not use unchecked casts, hence it is strongly-typed~\cite{NaftalinW06} and
+its map \<shares> actually holds values of type \<S> in its domain, only.
+For this consistency result, we had to pay the price
+of the dummy \<buyer> argument for the method \<accept>
+of the shared entities. Without that argument, the
+\emph{Consistency of Shareholders} property would not hold, since we could only write
+\<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
+Fig.~\ref{fig:simple_shared_entity}, with an unchecked cast that makes its code
+non-strongly-typed. In that case, also contracts not of type \<S> could call \<accept>
+and become shareholders.
 
-  public AbstractValidators
-    (V[] validators, BigInteger[] powers) {
-    super(validators, powers);
-  }
+There is, however, a problem with the reasoning
+in the previous paragraph. Namely, absence of unchecked
+operations guarantees strong typing of Java \emph{source} code. But what is installed
+and executed in blockchain is the Java bytecode that has been derived from
+the compilation of the code in Fig.~\ref{fig:simple_shared_entity}.
+Malicious users might install in blockchain some manually crafted bytecode,
+not derived from its Java source code compiled together with the source code
+in Fig.~\ref{fig:simple_shared_entity}.
+That crafted code might
+call the methods of \<SimpleSharedEntity>s in order to attack
+that contract. In particular,
+the signature of method \<accept> declares a parameter \<buyer> of type \<S> at source code level, but
+its compilation into Java bytecode declares an erased
+parameter \<buyer> of type \<PayableContract> instead.
+It follows that an attacker can install in blockchain a snippet of bytecode that calls
+\<accept> and passes \emph{any} \<PayableContract>, not only those that are instances of \<S>:
+the \emph{Consistency of Shareholders} property is easily violated at bytecode level.
 
-  public void reward
-    (String behaving, String misbehaving, 
-     BigInteger gasConsumed) { ... }
-}
+In general, the problem arises whenever a method declares a formal parameter of generic type,
+such as \<buyer> in method \<accept> of Fig.~\ref{fig:shared_entity}. Next section shows
+the actual significance of the consequent security risk.
 
-public class TendermintValidators
-    extends AbstractValidators<TendermintED25519Validator> {
-
-  public TendermintValidators
-        (TendermintED25519validator[] validators, 
-         BigInteger[] powers) {
-    super(validators, powers);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{The shared entity of the validators of a Hotmoka blockchain.
-  Its full code is available at \textsf{\url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/AbstractValidators.java}}.}\label{fig:abstract_validators}
-\end{figure}
-
-Since \<SimpleSharedEntity> allows shares to be sold and bought, this holds for
-its \<TendermintValidators> subclass as well: the set of validators
-is dynamic and it is possible to sell and buy voting power in order to invest in the blockchain
-and earn rewards at each block committed. At block creation time,
-Hotmoka calls method \<getShareholders> inherited from
-\<SimpleSharedEntity> and informs the
-underlying Tendermint engine about the identifiers of the validator nodes for the next blocks.
-Tendermint expects such validators to mine and vote the subsequent blocks, until a change in the
-validators' set occurs.
-
-It is important that the \emph{Consistency of Shareholders} property holds
-for \<TendermintValidators>: its shareholders
+In particular, it is important that the \emph{Consistency of Shareholders} property holds
+for the subclass \<TendermintValidators>: its shareholders
 must be \<TendermintED25519Validator>s (as declared in the generic
-signature of \<TendermintValidators>) that enforce a
+signature of \<TendermintValidators> in Fig.~\ref{fig:validators}) that enforce a
 match between their public key, that identifies who can spend the rewards sent to the validator,
@@
 This is impossible at source level, where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
-is not an instance of \<S>, which has been set to \<TendermintED25519Validator>.
+is not an instance of \<V>, which has been set to \<TendermintED25519Validator>.
 But a Hotmoka blockchain contains only the bytecode of \<SimpleSharedEntity>,
@@
 \emph{does} succeed, since \<attacker> is an externally owned account and all such accounts
-are instances of \<PayableContract>.
+are instances of \<PayableContract> (Fig.~\ref{fig:hierarchy-entities}).
 That transaction adds \<attacker> to the shareholders of \<tv>,
@@
   \end{center}
-  \caption{An attacker that exploits a blockchain node for validation and fraudolently earns the rewards of the work of that node.}\label{fig:attacker}
+  \caption{An attacker that exploits the work of a blockchain validator node and fraudolently earns the rewards of that work.}\label{fig:attacker}
 \end{figure}

### 2022-02-15 cluster_computing/001_abstract.tex
Edits in abstract, conclusions and related works

@@
 \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
+This is safe at source level, since the compiler takes care of checking that
+types are correct, before erasure. However, the types of the generated bytecode
+are erased and consequently weaker. In a permissionless blockchain, where
+every user can call the bytecode of smart contracts installed by other users,
+these weaker types pose a risk of attack.
 %
-This paper originated from a real security issue that we found
+Namely, this paper originated from a real security issue that we found
 while using generics for writing
-smart contracts for blockchain, in order to implement
-a contract for \emph{shared entities} (such as a company shared
-by its shareholders),
-for the Hotmoka blockchain, whose contracts are written in Java.
+smart contracts that implement
+\emph{shared entities} (such as a company shared by its shareholders),
+for the Hotmoka blockchain, in a subset of Java called Takamaka.
 %
@@
 a special case of shared entities.
-The analysis shows that the power of generics comes at the risk of a too permissive
-typing of the compiled code, due to the erasure mechanism, with a consequent possible attack
+The analysis shows that the power of generics comes at the risk of
+allowing a validator of an illegal type, that exploits the work of another
+validator in order to earn validation rewards, an actual attack
 to the validators' set. This paper proposes a solution
 that forces the compiler to generate more precise type information than
-those arising by erasure.
+that generated by erasure.

### 2022-02-15 cluster_computing/040_solution.tex
Reorganized sections on validators and attack
Edits in abstract, conclusions and related works

--- 2021-09-23 040_solution.tex
@@
 over-permissive erasure of the signature of method \<accept>,
-where \<buyer> is given type \<PayableContract>.
+where the compiler gives \<buyer> the type \<PayableContract>.
 Therefore, a solution is to oblige the compiler to generate a more
@@
   \begin{center}
-    \begin{lstlisting}[language=Java]
+    \begin{lstlisting}[language=Takamaka]
 public class TendermintValidators
@@
 
-  public TendermintValidators
-        (TendermintED25519validator[] validators, 
-         BigInteger[] powers) {
-    super(validators, powers);
+  public TendermintValidators(TendermintED25519validator validator, BigInteger power) {
+    super(validator, power);
   }
@@
   public void accept
-    (BigInteger amount, 
-     TendermintED25519Validator buyer, 
+    (BigInteger amount, TendermintED25519Validator buyer, 
      Offer<TendermintED25519Validator> offer) {
@@
   \end{center}
-  \caption{The shared entity of the validators of a Hotmoka blockchain built over Tendermint.
-  Its full code is available at \url{https://github.com/Hotmoka/hotmoka/blob/master/io-takamaka-code/src/main/java/io/takamaka/code/governance/tendermint/TendermintValidators.java}.}\label{fig:solution}
+  \caption{The fixed code of the shared entity of the validators of a Hotmoka blockchain built over Tendermint.}\label{fig:solution}
 \end{figure}
@@
 The fixed code is shown in Fig.~\ref{fig:solution}. The only difference is that method
-\<accept> has been redefined to enforce the correct type for \<buyer>. For the rest, that method
+\<accept> has been redefined to enforce the correct type for \<buyer>
+(see that redefined method also in Fig.~\ref{fig:hierarchy-entities}). For the rest, that method
 delegates to its implementation inherited from \<AbstractValidators>, through a call
@@
 Fig.~\ref{fig:shared_entity} has parameters \<buyer> and \<offer> of generic type
-\<S> and \<O>, respectively. The subclass in Fig.~\ref{fig:abstract_validators} restricts
+\<S> and \<O>, respectively. The subclass in Fig.~\ref{fig:validators} restricts
 \<S> to be a \<TendermintED25519Validator> and \<O> to be

### 2022-02-15 cluster_computing/050_related_work.tex
Edits in abstract, conclusions and related works

@@
 
-% \todo{Forse questa parte va spostata.}
-% 
-% 
-% \textbf{Compilation of generics --}
-% Generics is a way for writing code parametrized by type, in other words 
-% a same piece of code can be written by using a type placehorder 
-% (a \emph{type parameter}) and then instantiated for specific,
-% concrete types. There exists two common ways to implement generics in 
-% a programming language, that are often described in literature~\cite{generics_categories}
-%  as \emph{heterogeneous} and \emph{homogeneous}. In the heterogeneous 
-% approach the code is specialized for each instance
-% of the generic parameter; this is the approach adopted by C++ \emph{templates}.
-% Conversely, the homogeneous approach is the one provided by Java and .Net; 
-% in this case, only one instance of the code is maintained and shared by 
-% all generic instances.
-% This implementation is based on the type \emph{erasure} mechanism, where the 
-% generic parameter is replaced by the upwards bound of each instance, most 
-% often \<Object>.
-% Even if the heterogeneous approach is the safest, it is rarely applied in particular
-% in resource-constrained applications, because the code size may dramatically increase
-% as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
-% the heterogeneous approach obliges one to reinstall all instantiations of the 
-% generic code, with extra costs of gas.
-% Conversely, the homogeneous approach ensures a smaller consumption
-% of resources but can lead to the problems described in this paper. \\
-% 
-% \textbf{Smart contracts vulnerabilities --}
+Generics is a way for writing code parametrized by type, in other words 
+a same piece of code can be written by using a type placehorder 
+(a \emph{type parameter}) and then instantiated for specific,
+concrete types. There exists two common ways to implement generics in 
+a programming language, that are often described in literature~\cite{generics_categories}
+as \emph{heterogeneous} and \emph{homogeneous}. In the heterogeneous 
+approach, the code is specialized for each instance
+f the generic parameter; this is the approach adopted by C++ \emph{templates}.
+Conversely, the homogeneous approach is the one provided by Java and .Net; 
+in this case, only one instance of the code is maintained and shared by 
+all generic instances.
+This implementation is based on the type \emph{erasure} mechanism, where the 
+generic parameter is replaced by the upwards bound of each instance, most 
+often \<Object>.
+Even if the heterogeneous approach is the safest, it is rarely applied in particular
+in resource-constrained applications, because the code size may dramatically increase
+as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+the heterogeneous approach obliges one to reinstall all instantiations of the 
+generic code, with extra costs of gas.
+Conversely, the homogeneous approach ensures a smaller consumption
+of resources but can lead to the problems described in this paper.
 
@@
 costs of faulty software in the US is about tens of billions of dollars per year and represent
-approximately just under 1 percent of the Nation's gross domestic product.
+approximately just under one percent of the Nation's gross domestic product.
 The effects induced by errors in software development are even worse when such pieces of software
@@
 manage critical data such as money, digital assets and identities. For this reason, smart contracts
-vulnerabilities and correctness are becoming important in literature\,\cite{smart_contracts_verification}. Possible solutions can be
+vulnerabilities and correctness are becoming important in literature~\cite{smart_contracts_verification}. Possible solutions can be
 classified into three main categories: (i) static analysis of EVM bytecode,

### 2022-02-15 cluster_computing/090_conclusion.tex
Edits in abstract, conclusions and related works

--- 2021-07-29 090_conclusion.tex
@@
 This paper has shown that generic types can be useful in the definition
-of smart contracts, but introduce risks of security since in many
-programming languages, including Java, they get erased at compile time
+of smart contracts and can simplify the development of rather complex
+code such as those for shared entities, and support code reuse, for
+instance to implement the validators set of a blockchain network.
+However, this paper has shown that generic types
+introduce risks of security as well.
+Namely, many
+programming languages, including Java, erase them at compile time
 into types that might be too permissive for low-level calls, such as
-those that start blockchain transactions. Note that using a programming
+those that are started by blockchain transactions. Note that the use of a programming
 language without generics is not the solution: Solidity has no generics
@@
 (see \<accept> in Fig.~\ref{fig:solution}). This fixes the security risk,
-but cannot be regarded as an elegant solution. It is just a trick that
-works because it forces the compiler to generate some kind of bytecode.
+but cannot be regarded as the definite solution to the problem. It is just a trick that
+works because it forces the compiler to generate some specific kind of bytecode.
 A \emph{smarter} compiler might recognize the redefined \<accept> as
@@
 languages have no direct low-level
-counterpart but are implemented via instrumentation, from
-inner classes to closures. This is fine at source level, but allows
+counterpart but are implemented via instrumentation. Examples are
+inner classes and closures (lambda expressions). This is fine at source level, but allows
 low-level calls to easily circumvent

### 2022-02-15 cluster_computing/010_introduction.tex
Avoided repetition

@@
 The remainder of this paper is organized as follows.
-Sec..~\ref{sec:java_generics} discusses the management of generics in Java.
-Sec.~\ref{sec:shared_entities} introduces our real-life Java smart
-contract that uses generics. Sec.~\ref{sec:attack} shows that a na\"{i}ve
-deployment of a subclass of that contract leads to a code vulnerability.
+Sec.~\ref{sec:java_generics} discusses the management of generics in Java.
+Sec.~\ref{sec:takamaka} presents the basic notions about the Takamaka language for smart contracts in Java.
+Sec.~\ref{sec:shared_entities} shows our real-life Java smart
+contracts shared entities, that use generic types.
+Sec.~\ref{sec:validators} shows the instantiation of the shared entities to implement the validators' set
+of a proof of stake blockchain.
+Sec.~\ref{sec:attack} shows that a na\"{i}ve
+deployment of a subclass of the validators' set leads to a code vulnerability.
 Sec.~\ref{sec:fix} presents a fix to that vulnerability.
-Sec.~\ref{sec:related_work} discusses some related work and, finally,
+Sec.~\ref{sec:related_work} discusses some related work.
 Sec.~\ref{sec:conclusion} concludes.

### 2022-02-15 cluster_computing/015_generics_in_java.tex
Avoided repetition

--- 2021-07-29 015_generics_in_java.tex
@@
 is replaced by the upwards bound of each instance, mostly often \<Object>.
-Even if the heterogeneous approach is the safest, it is rarely applied in particular
+Even though the heterogeneous approach is the safest, it is rarely applied in particular
 in resource-constrained applications, because the code size may dramatically increase
-because of duplicated~\cite{generics_embedded_systems}. For code in blockchain,
+because of duplication~\cite{generics_embedded_systems}. For code in blockchain,
 the heterogeneous approach obliges one to reinstall all instantiations of the generic code,

### 2022-02-15 cluster_computing/050_related_work.tex
Avoided repetition

@@
 \section{Related Work}\label{sec:related_work}
-
-Generics is a way for writing code parametrized by type, in other words 
-a same piece of code can be written by using a type placehorder 
-(a \emph{type parameter}) and then instantiated for specific,
-concrete types. There exists two common ways to implement generics in 
-a programming language, that are often described in literature~\cite{generics_categories}
-as \emph{heterogeneous} and \emph{homogeneous}. In the heterogeneous 
-approach, the code is specialized for each instance
-f the generic parameter; this is the approach adopted by C++ \emph{templates}.
-Conversely, the homogeneous approach is the one provided by Java and .Net; 
-in this case, only one instance of the code is maintained and shared by 
-all generic instances.
-This implementation is based on the type \emph{erasure} mechanism, where the 
-generic parameter is replaced by the upwards bound of each instance, most 
-often \<Object>.
-Even if the heterogeneous approach is the safest, it is rarely applied in particular
-in resource-constrained applications, because the code size may dramatically increase
-as some code is duplicated~\cite{generics_embedded_systems}. For code in blockchain,
-the heterogeneous approach obliges one to reinstall all instantiations of the 
-generic code, with extra costs of gas.
-Conversely, the homogeneous approach ensures a smaller consumption
-of resources but can lead to the problems described in this paper.
 

### 2022-02-21 cluster_computing/010_introduction.tex
figures

@@
 Since, at run time, it is the bytecode that gets executed,
-everything can be passed for \<p>, not just a \<Person> instance.
+everything can be passed for \<p>, not just a \<Person> instance, as illustrated in Fig.\,\ref{figure.solidity_problem}.
 The compiler cannot even enforce strong typing
@@
 costed millions of dollars.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/solidity_problem}
+\caption{Example of possible problem that can occur in Solidity due to the absence of a strong typing mechanism.}
+\label{figure.solidity_problem}
+\end{figure}
 

### 2022-02-21 cluster_computing/015_generics_in_java.tex
figures

@@
 the two type variables \<S> and \<O> have been \emph{erased} and replaced with their
-respective upper bound.
+respective upper bound, as illustrated in Fig.\,\ref{figure.java_generics_erasure}.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/java_generics_erasure}
+\caption{Example of generics implementation by erasure in Java.}
+\label{figure.java_generics_erasure}
+\end{figure}
 

### 2022-02-21 cluster_computing/017_takamaka.tex
Edited the shared entity description
figures

@@
 In the following we will report a simplified presentation of the code of some
-of such classes. The full code is in the Github repository of Hotmoka.
+of such classes. The full code is in the Github repository of Hotmoka \footnote{\url{https://github.com/Hotmoka}}.
 
@@
 
-The \<io.takamaka.code.lang.Contract> class implements objects that can be
-persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
-Their balance is available through a \<balance> method. This is a \<@View> method, meaning
-that it can be called without paying gas (the measure of execution cost), since such methods
-cannot have side-effects and consequently do not modify the storage of the blockchain.
-Payments can be received only through methods annotated as \<@Payable>.
-The \<Contract> superclass has no such methods, but subclasses may have.
-For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
-a method \<receive> to receive payments from its caller. Many methods (inclusing all
-\<@Payable> methods) need to identify their caller. This is done by adding the
-\<@FromContract> annotation, that guarantees that the caller is a contract, that is available
-inside the method as \<caller()>.
-
-Method calls started from outside the blockchain (for instance, from a client such as a wallet
-or from a web application), must specify an already existing \<ExternallyOwnedAccount>
-as caller. This account will pay for the gas of the execution. The blockchain will accept
-the call only if is signed with the private key that matches the public key provided to
-the constructor of the account when it was created. Method \<publicKey> allows one to
-recover that public key and method \<nonce> allows one to get a progressive identifier that
-can be used to distinguish successive calls with the same account, to force their order
-of execution and to avoid replaying.
-All that is very similar to Solidity, except for the fact that externally owned accounts
-are actual Java objects inside the blockchain, not just an abstraction of a public key.
-
-Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
-Both proof of work and proof of stacke can be used, for instance. In particular, if proof of stake
-is used, then each validator node of the blockchain must specify a
-\<io.takamaka.code.governance.Validator> object, that plays the role of the
-banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
-It is a special externally owned account, with an extra \<id> method that provides the
-identifier of the validator node inside the blockchain network. This identifier depends from the
-specific blockchain implementation.
-For instance, the subclass
-\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
-implements \<id> as for the Tendermint blockchain engine, that is,
-as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key.
-
-\begin{figure}[t]
+\begin{figure}[th]
   \begin{center}
@@
 \end{figure}
+
+The \<io.takamaka.code.lang.Contract> class implements objects that can be
+persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
+Their balance is available through a \<balance> method. This is a \<@View> method, meaning
+that it can be called without paying gas (the measure of execution cost), since such methods
+cannot have side-effects and consequently do not modify the storage of the blockchain.
+Payments can be received only through methods annotated as \<@Payable>.
+The \<Contract> superclass has no such methods, but subclasses may have.
+For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
+a method \<receive> to receive payments from its caller. Many methods (inclusing all
+\<@Payable> methods) need to identify their caller. This is done by adding the
+\<@FromContract> annotation, that guarantees that the caller is a contract, that is available
+inside the method as \<caller()>.
+
+Method calls started from outside the blockchain (for instance, from a client such as a wallet
+or from a web application), must specify an already existing \<ExternallyOwnedAccount>
+as caller. This account will pay for the gas of the execution. The blockchain will accept
+the call only if is signed with the private key that matches the public key provided to
+the constructor of the account when it was created. Method \<publicKey> allows one to
+recover that public key and method \<nonce> allows one to get a progressive identifier that
+can be used to distinguish successive calls with the same account, to force their order
+of execution and to avoid replaying.
+All that is very similar to Solidity, except for the fact that externally owned accounts
+are actual Java objects inside the blockchain, not just an abstraction of a public key.
+An exemplification of a call made to a \<PayableContract> is reported in Fig.\,\ref{figure.takamaka_payable_contract}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
+\caption{}
+\label{figure.takamaka_payable_contract}
+\end{figure}
+
+Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
+Both proof of work and proof of stake can be used, for instance. In particular, if proof of stake
+is used, then each validator node of the blockchain must specify a
+\<io.takamaka.code.governance.Validator> object, that plays the role of the
+banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
+It is a special externally owned account, with an extra \<id> method that provides the
+identifier of the validator node inside the blockchain network. This identifier depends from the
+specific network.
+For instance, the subclass
+\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
+implements \<id> as for the Tendermint blockchain engine, that is,
+as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key.

### 2022-02-22 cluster_computing/020_shared_entities_with_generics.tex
figures

@@
 
-A \emph{shared entity} is concept that often arises in blockchain
+A \emph{shared entity} is a concept that often arises in blockchain
 applications. Namely, a shared entity is something divided into \emph{shares}. Participants,
@@
 
-    public Offer(S seller, BigInteger sharesOnSale, BigInteger cost, long duration) {
+    public Offer(S seller, BigInteger sharesOnSale, 
+                 BigInteger cost,long duration) {
       this.seller = seller;
@@
 
+\begin{figure}[tp]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class SimpleSharedEntity
+    <S extends PayableContract,O extends Offer<S>>
+    extends PayableContract implements SharedEntity<S,O> {
+
+  private StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
+  private StorageSet<O> offers = new StorageTreeSet<>();
+
+  public SimpleSharedEntity
+      (S shareholder, BigInteger share) {
+    addShares(shareholder, share);
+  }
+
+  @Override @View
+  public final BigInteger sharesOf(S shareholder) {
+    return shares.getOrDefault(shareholder, ZERO);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void place(BigInteger amount, O offer) {
+    require(offer.seller == caller(), "unauthorized");
+    require(shares.containsKey(offer.seller), "unauthor.");
+    require(sharesOf(offer.seller) - sharesOnSaleOf(offer.seller) >= 
+            offer.sharesOnSale, "not enough shares");
+    offers.add(offer);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void accept(BigInteger amount, S buyer, O offer) {
+    require(caller() == buyer, "unauthorized");
+    require(offers.contains(offer), "unknown offer");
+    require(offer.isOngoing(), "the offer is not ongoing");
+    require(offer.cost <= amount, "not enough money");
+    offers.remove(offer);
+    removeShares(offer.seller, offer.sharesOnSale);
+    addShares(buyer, offer.sharesOnSale);
+    offer.seller.receive(offer.cost);
+  }
+
+  @Override @View
+  public final BigInteger sharesOnSaleOf(S shareholder) {
+    return offers.stream()
+     .filter(o -> o.seller == shareholder && o.isOngoing())
+     .map(offer -> offer.sharesOnSale)
+     .reduce(ZERO, BigInteger::add);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
+\end{figure}
+
 The \<SharedEntityView> interface at the top of the hierarchy in Fig.~\ref{fig:hierarchy-entities}
@@
 the number of shares of a shareholder. As typical in Takamaka, a \<snapshot> method allows one
-to create a frozen read-only copy of an entity (in constant time), useful when an enity must be queried from
+to create a frozen read-only copy of an entity (in constant time), useful when an entity must be queried from
 a client without the risk of race conditions if another client is modifying the same entity concurrently.
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/shared_entity}
+\caption{Exemplification of a Takamaka shared entity and of its connections with the objects persisted in the blockchain.}
+\label{figure.shared_entity}
+\end{figure}
+
 As said before, the annotation \<@FromContract> on both \<place> and \<accept> enforces that only
@@
 parameter \<buyer> in \<accept>.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class SimpleSharedEntity
-    <S extends PayableContract,O extends Offer<S>>
-    extends PayableContract implements SharedEntity<S,O> {
-
-  private StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private StorageSet<O> offers = new StorageTreeSet<>();
-
-  public SimpleSharedEntity
-      (S shareholder, BigInteger share) {
-    addShares(shareholder, share);
-  }
-
-  @Override @View
-  public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, ZERO);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "unauthorized");
-    require(shares.containsKey(offer.seller), "unauthor.");
-    require(sharesOf(offer.seller) - sharesOnSaleOf(offer.seller) >= offer.sharesOnSale, "not enough shares");
-    offers.add(offer);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "unauthorized");
-    require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the offer is not ongoing");
-    require(offer.cost <= amount, "not enough money");
-    offers.remove(offer);
-    removeShares(offer.seller, offer.sharesOnSale);
-    addShares(buyer, offer.sharesOnSale);
-    offer.seller.receive(offer.cost);
-  }
-
-  @Override @View
-  public final BigInteger sharesOnSaleOf(S shareholder) {
-    return offers.stream()
-     .filter(o -> o.seller == shareholder && o.isOngoing())
-     .map(offer -> offer.sharesOnSale)
-     .reduce(ZERO, BigInteger::add);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
-\end{figure}
 

### 2022-03-01 cluster_computing/001_abstract.tex
Reorganized figures

@@
 to produce parametrized code, that can be instantiated for each
-concrete type provided for them. 
+concrete type provided for the placeholders.
 %
-In many programming languages, such as Java, they are implemented by
+In many programming languages, such as Java, generics are implemented by
 \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
@@
 the validators' set of the blockchain itself is
-a special case of shared entities.
+a special case of shared entity.
 The analysis shows that the power of generics comes at the risk of
-allowing a validator of an illegal type, that exploits the work of another
+allowing the intrusion of
+a validator of an illegal type, that exploits the work of another
 validator in order to earn validation rewards, an actual attack

### 2022-03-01 cluster_computing/010_introduction.tex
Reorganized figures

@@
 are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics might have security issue at the level of compiled Java code and this paper originated from a
+However, generics might have security issues
+at the level of compiled Java code and this paper originated from a
 real issue that we found in our code.
@@
 library of the Takamaka language, and to demonstrate that a na\"{i}ve use
-of Java generics can lead to a code security vulnerability which
+of Java generics can lead to a code security vulnerability that
 allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
@@
 Sec.~\ref{sec:shared_entities} shows our real-life Java smart
-contracts shared entities, that use generic types.
+contracts for shared entities, that use generic types.
 Sec.~\ref{sec:validators} shows the instantiation of the shared entities to implement the validators' set
@@
 Sec.~\ref{sec:conclusion} concludes.
+
+This paper is a revised and extended version of~\cite{BeniniGMS21}.
+In comparison to that paper, Sec.~\ref{sec:takamaka} and
+Sec.~\ref{sec:validators} are new; eight figures have been added;
+all other sections have been expanded.

### 2022-03-01 cluster_computing/015_generics_in_java.tex
Reorganized figures

@@
 
-There exists two common ways to implement generics in a programming language,
+There exist two common ways to implement generics in a programming language,
 that are often described in literature as \emph{heterogeneous}
@@
 is replaced by the upwards bound of each instance, mostly often \<Object>.
-Even though the heterogeneous approach is the safest, it is rarely applied in particular
+Even though the heterogeneous approach is the safest, it is rarely applied, in particular
 in resource-constrained applications, because the code size may dramatically increase
-because of duplication~\cite{generics_embedded_systems}. For code in blockchain,
+as a consequence of duplication~\cite{generics_embedded_systems}. For code in blockchain,
 the heterogeneous approach obliges one to reinstall all instantiations of the generic code,
@@
 However, this guarantee applies to Java source code compiled by the Java compiler,
-not to bytecode that can be generated manually, in order to attack class \<SharedEntity>,
-as we will show later.
+not to bytecode that can be generated manually, in order to attack instance of
+the \<SharedEntity> class, as we will show later.

### 2022-03-01 cluster_computing/017_takamaka.tex
figures
Reorganized figures

@@
 In the following we will report a simplified presentation of the code of some
-of such classes. The full code is in the Github repository of Hotmoka \footnote{\url{https://github.com/Hotmoka}}.
+of such classes. The full code is in the Github repository of Hotmoka~\cite{hotmoka_repository}.
 
@@
 \<io.takamaka.code.lang.Storage> class.
-This is the main difference with other attemps at using Java for writing smart contracts:
+This is the main difference with other attempts at using Java for writing smart contracts:
 the programmer does not code the serialization and deserialization of objects into
 a \emph{keeper} or a key/value map, but simply extends \<Storage> and objects get
-persisted automatically \emph{out of magic}. In this sense, Takamaka follows the apporach of
+persisted automatically \emph{out of magic}. In this sense, Takamaka follows the approach of
 Solidity, but using Java.
+
+The \<io.takamaka.code.lang.Contract> class implements objects that can be
+persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
+Their balance is available through a \<balance> method. This is a \<@View> method, meaning
+that it can be called without paying gas (the measure of execution cost), since such methods
+cannot have side-effects and consequently do not modify the storage of the blockchain.
+Payments can be received only through methods annotated as \<@Payable>.
+The \<Contract> superclass has no such methods, but subclasses may have.
+For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
+a method \<receive> to receive payments from its caller. Many methods (inclusing all
+\<@Payable> methods) need to identify their caller. This is done by adding the
+\<@FromContract> annotation, that guarantees that the caller is a contract, available
+inside the method as \<caller()>.
+
+Method calls started from outside the blockchain (for instance, from a client such as a wallet
+or from a web application), must specify an already existing \<ExternallyOwnedAccount>
+as caller. This account will pay for the gas of the execution. The blockchain will accept
+the call only if is signed with the private key that matches the public key provided to
+the constructor of the account when it was created. Method \<publicKey> allows one to
+recover that public key and method \<nonce> allows one to get a progressive identifier that
+can be used to distinguish successive calls with the same account, to force their order
+of execution and to avoid replaying.
+All that is very similar to Solidity, except for the fact that externally owned accounts
+are actual Java objects inside the blockchain, not just an abstraction of a public key.
+An exemplification of a call made to a \<PayableContract> is reported in Fig.\,\ref{figure.takamaka_payable_contract}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
+\caption{Exemplification of the Takamaka persistent objects stored in the blockchain.}
+\label{figure.takamaka_payable_contract}
+\end{figure}
 
@@
 
-  @Override public @View final String id() {
+  @Override @View public final String id() {
     return id;
@@
   \caption{The account of a validator and its specialization for a Hotmoka blockchain based on Tendermint.}\label{fig:validator}
-\end{figure}
-
-The \<io.takamaka.code.lang.Contract> class implements objects that can be
-persisted in blockchain \emph{and} have a balance. Therefore, they can receive and provide payments.
-Their balance is available through a \<balance> method. This is a \<@View> method, meaning
-that it can be called without paying gas (the measure of execution cost), since such methods
-cannot have side-effects and consequently do not modify the storage of the blockchain.
-Payments can be received only through methods annotated as \<@Payable>.
-The \<Contract> superclass has no such methods, but subclasses may have.
-For instance, its \<io.takamaka.code.lang.PayableContract> subclass has
-a method \<receive> to receive payments from its caller. Many methods (inclusing all
-\<@Payable> methods) need to identify their caller. This is done by adding the
-\<@FromContract> annotation, that guarantees that the caller is a contract, that is available
-inside the method as \<caller()>.
-
-Method calls started from outside the blockchain (for instance, from a client such as a wallet
-or from a web application), must specify an already existing \<ExternallyOwnedAccount>
-as caller. This account will pay for the gas of the execution. The blockchain will accept
-the call only if is signed with the private key that matches the public key provided to
-the constructor of the account when it was created. Method \<publicKey> allows one to
-recover that public key and method \<nonce> allows one to get a progressive identifier that
-can be used to distinguish successive calls with the same account, to force their order
-of execution and to avoid replaying.
-All that is very similar to Solidity, except for the fact that externally owned accounts
-are actual Java objects inside the blockchain, not just an abstraction of a public key.
-An exemplification of a call made to a \<PayableContract> is reported in Fig.\,\ref{figure.takamaka_payable_contract}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
-\caption{}
-\label{figure.takamaka_payable_contract}
 \end{figure}
@@
 implements \<id> as for the Tendermint blockchain engine, that is,
-as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key.
+as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key
+(see its code in Fig.~\ref{fig:validator}).

### 2022-03-01 cluster_computing/020_shared_entities_with_generics.tex
Reorganized figures

@@
 
-  @View
-  StorageSetView<O> getOffers();
+  @View StorageSetView<O> getOffers();
 
-  @View
-  BigInteger sharesOnSaleOf(S shareholder);
+  @View BigInteger sharesOnSaleOf(S shareholder);
 
@@
 
-    @View
-    public boolean isOngoing() {
+    @View public boolean isOngoing() {
       return now() <= expiration;
@@
 \<S> is the type of the shareholders and \<O> is the type of the sale offers of shares.
-
-\begin{figure}[tp]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class SimpleSharedEntity
-    <S extends PayableContract,O extends Offer<S>>
-    extends PayableContract implements SharedEntity<S,O> {
-
-  private StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
-  private StorageSet<O> offers = new StorageTreeSet<>();
-
-  public SimpleSharedEntity
-      (S shareholder, BigInteger share) {
-    addShares(shareholder, share);
-  }
-
-  @Override @View
-  public final BigInteger sharesOf(S shareholder) {
-    return shares.getOrDefault(shareholder, ZERO);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void place(BigInteger amount, O offer) {
-    require(offer.seller == caller(), "unauthorized");
-    require(shares.containsKey(offer.seller), "unauthor.");
-    require(sharesOf(offer.seller) - sharesOnSaleOf(offer.seller) >= 
-            offer.sharesOnSale, "not enough shares");
-    offers.add(offer);
-  }
-
-  @Override @FromContract(PayableContract.class) @Payable
-  public void accept(BigInteger amount, S buyer, O offer) {
-    require(caller() == buyer, "unauthorized");
-    require(offers.contains(offer), "unknown offer");
-    require(offer.isOngoing(), "the offer is not ongoing");
-    require(offer.cost <= amount, "not enough money");
-    offers.remove(offer);
-    removeShares(offer.seller, offer.sharesOnSale);
-    addShares(buyer, offer.sharesOnSale);
-    offer.seller.receive(offer.cost);
-  }
-
-  @Override @View
-  public final BigInteger sharesOnSaleOf(S shareholder) {
-    return offers.stream()
-     .filter(o -> o.seller == shareholder && o.isOngoing())
-     .map(offer -> offer.sharesOnSale)
-     .reduce(ZERO, BigInteger::add);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
-\end{figure}
 
@@
 Offers can be placed on sale
-by calling the \<place> method with a sale \<offer>.
+by calling the \<place> method with a sale \<offer>
+(see Fig.~\ref{figure.shared_entity}).
 This method is annotated as \<@FromContract> since the caller must be
@@
 
+\begin{figure}[tp]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class SimpleSharedEntity
+    <S extends PayableContract,O extends Offer<S>>
+    extends PayableContract implements SharedEntity<S,O> {
+
+  private StorageTreeMap<S,BigInteger> shares = new StorageTreeMap<>();
+  private StorageSet<O> offers = new StorageTreeSet<>();
+
+  public SimpleSharedEntity(S shareholder, BigInteger share) {
+    addShares(shareholder, share);
+  }
+
+  @Override @View public final BigInteger sharesOf(S shareholder) {
+    return shares.getOrDefault(shareholder, ZERO);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void place(BigInteger amount, O offer) {
+    require(offer.seller == caller(), "unauthorized");
+    require(shares.containsKey(offer.seller), "unauthor.");
+    require(sharesOf(offer.seller) - sharesOnSaleOf(offer.seller) >= 
+            offer.sharesOnSale, "not enough shares");
+    offers.add(offer);
+  }
+
+  @Override @FromContract(PayableContract.class) @Payable
+  public void accept(BigInteger amount, S buyer, O offer) {
+    require(caller() == buyer, "unauthorized");
+    require(offers.contains(offer), "unknown offer");
+    require(offer.isOngoing(), "the offer is not ongoing");
+    require(offer.cost <= amount, "not enough money");
+    offers.remove(offer);
+    removeShares(offer.seller, offer.sharesOnSale);
+    addShares(buyer, offer.sharesOnSale);
+    offer.seller.receive(offer.cost);
+  }
+
+  @Override @View public final BigInteger sharesOnSaleOf(S shareholder) {
+    return offers.stream()
+      .filter(o -> o.seller == shareholder && o.isOngoing())
+      .map(offer -> offer.sharesOnSale)
+      .reduce(ZERO, BigInteger::add);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
+\end{figure}
+
 Fig.~\ref{fig:simple_shared_entity} shows a portion of our \<SimpleSharedEntity>

### 2022-03-01 cluster_computing/025_validators.tex
figures
Reorganized figures

@@
 generic engine for replicating an application over a network of nodes. In our case,
-the application is the executor of smart contracts in Java, such as that in
+the application is an executor of smart contracts in Java, such as that in
 Fig.~\ref{fig:simple_shared_entity}. Tendermint is based on a proof of stake
 consensus, which means that a selected dynamic subset of the nodes is in charge of
-validating the transactions and voting their acceptance. As we have said,
+validating the transactions and voting their acceptance. 
+As we have said,
 Hotmoka models validator nodes as \<Validator> objects, that are externally owned accounts
@@
 policy for rewarding or changing the validators' set dynamically.
+
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public interface Validators<V extends Validator>
+      extends SharedEntity<V,Offer<V>> {
+
+  @View BigInteger getStake(V validator);
+
+  @FromContract @Payable
+  void reward(String behaving, String misbehaving, BigInteger gasConsumed);
+}
+
+public abstract class AbstractValidators<V extends Validator>
+    extends SimpleSharedEntity<V,Offer<V>> {
+
+  public AbstractValidators(V validator, BigInteger power) {
+    super(validator, power);
+  }
+
+  public void reward(String behaving, String misbehaving,
+                     BigInteger gasConsumed) { ... }
+}
+
+public class TendermintValidators
+   extends AbstractValidators<TendermintED25519Validator> {
+
+  public TendermintValidators(TendermintED25519validator validator, BigInteger power) {
+    super(validator, power);
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{The shared entity of the validators set of a Hotmoka blockchain.}\label{fig:validators}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.9\linewidth]{figures/takamaka_validators}
+\caption{Hierarchy of classes for implementing Hotmoka Validators.}
+\label{figure.takamaka_validators}
+\end{figure}
 
@@
 and the distribution of the reward and is a subclass of \<SimpleSharedEntity>
-(see Fig.~\ref{fig:validators}).
+(see Fig.~\ref{fig:validators} and Fig.~\ref{figure.takamaka_validators}).
 Shares are voting power in this case.
@@
 validators' set occurs.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public interface Validators<V extends Validator>
-    extends SharedEntity<V,Offer<V>> {
-
-  @View
-  BigInteger getStake(V validator);
-
-  @FromContract @Payable
-  void reward(String behaving, String misbehaving, BigInteger gasConsumed);
-}
-
-public abstract class AbstractValidators
-    <V extends Validator>
-    extends SimpleSharedEntity<V,Offer<V>> {
-
-  public AbstractValidators
-      (V validator, BigInteger power) {
-    super(validator, power);
-  }
-
-  public void reward(String behaving, String misbehaving, BigInteger gasConsumed) { ... }
-}
-
-public class TendermintValidators
-   extends AbstractValidators<TendermintED25519Validator> {
-
-  public TendermintValidators(TendermintED25519validator validator, BigInteger power) {
-    super(validator, power);
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{The shared entity of the validators set of a Hotmoka blockchain.}\label{fig:validators}
-\end{figure}

### 2022-03-01 cluster_computing/030_generics_attack.tex
figures
Reorganized figures

@@
 
-In general, the problem arises whenever a method declares a formal parameter of generic type,
-such as \<buyer> in method \<accept> of Fig.~\ref{fig:shared_entity}. Next section shows
-the actual significance of the consequent security risk.
+\begin{figure}[t]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class Attacker extends ExternallyOwnedAccount {
+
+  public Validator(String publicKey) {
+    super(publicKey);
+  }
+
+  @View
+  public String id() {
+    return /* id of the victim blockchain node, unrelated to publicKey */
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{An attacker that exploits the work of a blockchain validator node and fraudolently earns the rewards of that work.}\label{fig:attacker}
+\end{figure}
+
+\begin{figure}[b]
+\centering
+\includegraphics[width=.9\linewidth]{figures/attack}
+\caption{Example of possible attack to a smart contract using Java generics.}
+\label{figure.attack}
+\end{figure}
 
@@
 by accepting an ongoing sale \<offer> of shares through a call to
-\<tv.accept(offer.cost, attacker, offer)> (see Fig.~\ref{fig:simple_shared_entity}).
-This is impossible at source level, where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
+\<tv.accept(offer.cost, attacker, offer)> (Fig.~\ref{fig:simple_shared_entity}).
+This is impossible at source level (left part of Fig.~\ref{figure.attack}),
+where that call does \emph{not} compile, since \<attacker> has type \<Attacker> that
 is not an instance of \<V>, which has been set to \<TendermintED25519Validator>.
@@
 where the signature of \<accept> has been erased into
-\<accept(BigInteger amount, PayableContract buyer, Offer offer)>.
+\<accept(BigInteger amount, PayableContract buyer, Offer offer)>
+(see Fig.~\ref{figure.java_generics_erasure} and the right part of
+Fig.~\ref{figure.attack}).
 Hence a blockchain transaction that invokes \<tv.accept(offer.cost, attacker, offer)>
@@
 validator's slavery.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class Attacker extends ExternallyOwnedAccount {
-
-  public Validator(String publicKey) {
-    super(publicKey);
-  }
-
-  @View
-  public String id() {
-    return /* id of the victim blockchain node, unrelated to publicKey */
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{An attacker that exploits the work of a blockchain validator node and fraudolently earns the rewards of that work.}\label{fig:attacker}
-\end{figure}

### 2022-03-01 cluster_computing/040_solution.tex
solution figures
Reorganized figures

@@
-\section{Fixing the Compilation of the Contract} \label{sec:fix}
+\section{Fixing the Compilation of the Contract}\label{sec:fix}
 
@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
   \begin{center}
@@
 
-  public void accept
-   (BigInteger,TendermintED25519Validator,Offer)
+  public void accept(BigInteger,TendermintED25519Validator,Offer)
    aload_0
@@
    aload_3
-   invokespecial AbstractValidators
-     .accept(BigInteger,PayableContract,Offer)
+   invokespecial AbstractValidators.accept
+                        (BigInteger,PayableContract,Offer)
    return
@@
    aload_3
-   invokevirtual accept
-      (BigInteger,TendermintED25519Validator,Offer)
+   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
    return
@@
 bridge method and its checked cast guarantee that only \<TendermintED25519Validator>s
-can become validators. For instance, an instance of \<Attacker> (Fig.~\ref{fig:attacker})
+can become validators. As shown in Fig.~\ref{figure.solution},
+an instance of \<Attacker> (Fig.~\ref{fig:attacker})
 cannot be passed to the first \<accept> (type mismatch) and makes the second \<accept>
@@
 parameters. In the future, a compiler might perform this automatically or a static analysis
-tool might issue a warning when such redefinition is needed. Currently, however, this is
-left to the programmer of the smart contracts, who might overlook the issue
+tool might issue a warning when such redefinition is needed. Currently, however, that is
+left to the programmer of the smart contracts, who might overlook the problem
 and give rise to security issues, as shown in Sec.~\ref{sec:attack}.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=.95\linewidth]{figures/solution}
+\caption{Example of how the proposed solution works at run time when the \<accept> method is called by a \<Validator> or an \<Attacker>.}
+\label{figure.solution}
+\end{figure}

### 2022-03-02 cluster_computing/001_abstract.tex
title, abstract and introduction rewriting + biblio fix

--- 2022-02-15 cluster_computing/001_abstract.tex
@@
-Generics are a powerful feature of programming languages that allows one
-to write highly reusable code.
-%
-More specifically, they are based on the use of type placeholders in order
-to produce parametrized code, that can be instantiated for each
-concrete type provided for them. 
-%
-In many programming languages, such as Java, they are implemented by
-\emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
-This is safe at source level, since the compiler takes care of checking that
-types are correct, before erasure. However, the types of the generated bytecode
-are erased and consequently weaker. In a permissionless blockchain, where
-every user can call the bytecode of smart contracts installed by other users,
-these weaker types pose a risk of attack.
-%
-Namely, this paper originated from a real security issue that we found
-while using generics for writing
-smart contracts that implement
-\emph{shared entities} (such as a company shared by its shareholders),
-for the Hotmoka blockchain, in a subset of Java called Takamaka.
-%
-The considered case study is particularly important since
-the validators' set of the blockchain itself is
-a special case of shared entities.
-The analysis shows that the power of generics comes at the risk of
-allowing a validator of an illegal type, that exploits the work of another
-validator in order to earn validation rewards, an actual attack
-to the validators' set. This paper proposes a solution
-that forces the compiler to generate more precise type information than
-that generated by erasure.
+Smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order automatically enforce some effects when particularly events occur.
+The writing of smart contracts is a complex and critical activity which can benefit from the use of high-level features of programming languages, like generics.
+In many programming languages, such as Java, generics are implemented by \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
+This is safe at source level, since the compiler takes care of checking that types are correct, before erasure.
+However, the types of the generated bytecode are erased and consequently weaker.
+In a permissionless blockchain, where every user can call the bytecode of smart contracts installed by other users, these weaker types pose a risk of attack.
+In this paper we start from a real security issue that we found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain, and we propose a possible solution based on an appropriate code rewriting.

### 2022-03-02 cluster_computing/010_introduction.tex
title, abstract and introduction rewriting + biblio fix

@@
 
-Blockchains exploit the redundant, concurrent execution of the same
-transactions on a decentralized network of many machines,
-in order to enforce their execution in accordance with
-a set of predefined rules. Namely, blockchains make it hard, for a single machine,
-to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
-gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
-has been the first blockchain's success story. Here
-transactions are programmed in a non-Turing complete bytecode language,
-almost exclusively used to implement transfers of units of coins between \emph{accounts}.
-
-A few years after Bitcoin, another blockchain, called
-Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
-transactions in an actual, imperative and Turing-complete programming language, called Solidity.
-Solidity's code is organized in \emph{smart contracts}, that can be seen as
-objects that control money. A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.
-Ethereum's transactions can hence execute much more than coin transfers. Namely,
-they run object constructors and methods, which results in a sort
-of \emph{world computer} that persists the same objects inside the memory of all the
-computers composing the blockchain's network.
-
-In Solidity's bytecode,
-non-primitive values are referenced through a very general
-\<address> type. For instance, a Solidity method
-\<child(Person p, uint256 n) returns Person> actually compiles
-into \<child(address p, uint256 n) returns address>, losing most
-type information~\cite{CrafaPZ19}.
-Since, at run time, it is the bytecode that gets executed,
-everything can be passed for \<p>, not just a \<Person> instance, as illustrated in Fig.\,\ref{figure.solidity_problem}.
-The compiler cannot even enforce strong typing
-by generating defensive type instance checks and casts, because
-values are unboxed in Ethereum: they have no attached
-type information at run time,
-they are just numerical \emph{addresses}.
-It follows that inside the \<child> method, an eventual call to a \<Person>'s method
-on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
-In other words, Solidity is not strongly typed.
-Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
-to another method, such as on \<p> passed to \<child>, since an attacker can pass crafted
-objects for \<p>, with arbitrary implementations for their methods,
-which can result in the unexpected execution of
-dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
-costed millions of dollars.
-
-\begin{figure}[ht]
+In recent years the Blockchain technology has rapidily emerged as a powerfull tool for supporting the development of many and innovative services and infrastructures.
+Blockchain-enabled applications are spreading across diverse sectors such as supply chain, business, healthcare, IoT, privacy, and data management \cite{blockchain_applications}.
+A blockchain is essentially a \textit{distributed ledger}, namely a database replicated across different locations and synchronized by multiple independent participants.
+Blockchains exploit the redundant, concurrent execution of the same transactions on a decentralized network of many machines, in order to enforce their execution in accordance with a set of predefined rules.
+Namely, blockchains make it hard, for a single machine, to disrupt the semantics of transactions or their ordering: a misbehaving single machine gets immediately put out of consensus and isolated.
+%
+\begin{figure*}[th]
 \centering
-\includegraphics[width=.9\linewidth]{figures/solidity_problem}
+\includegraphics[width=0.8\linewidth]{figures/solidity_problem}
 \caption{Example of possible problem that can occur in Solidity due to the absence of a strong typing mechanism.}
 \label{figure.solidity_problem}
-\end{figure}
+\end{figure*}
+%
+The key innovation introduced by this technology is a mechanism  able to reach an emergent agreement about a global state without the need for a central authority.
+Morever, another peculiarity is that the consensus is not explicit, because there is not a fixed moment when it occurs.
+
+Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}  has been the first blockchain's success story.
+Here transactions are programmed in a non-Turing complete bytecode language, almost exclusively used to implement transfers of units of coins between \emph{accounts}, providing a totally decentralized P2P digital cash system based on a distributed public ledger.
+%
+A few years after Bitcoin, another blockchain, called Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming transactions in an actual, imperative and Turing-complete programming language, called Solidity.
+The major innovation of Ethereum is the contrustruction through its nodes of a distributed \emph{world computer} that can run general purpose code.
+Indee, if the term ``distributed ledger'' is usually used to describe blockchains like Bitcoin, Ethereum is often defined as a ``distributed state machine''.
+Solidity's code is organized in \emph{smart contracts}, namely pieces of code which are stored in the blockchain and are executed when a particular event occurs, e.g.
+when a transaction is scheduled.
+From a theoretical point of view, a smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.
+Through smart contracts, Ethereum's transactions can hence execute much more than coin transfers.
+In this case the global shared state is given by a set of objects which are persisted and manipulated in the same way by all nodes in the blockchain through the execution of the same object constructors and methods.
+
+The world computer built by Ethereum is known as Ethereum Virtual Machine (EVM) which is the platform where accounts and smart contracts live and are executed.
+Solidity is high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
+At this regards, we can observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
+For instance, a Solidity method \<child(Person p, uint256 n) returns Person> actually compiles into \<child(address p, uint256 n) returns address>, losing most type information~\cite{CrafaPZ19}.
+Since, at run time, it is the bytecode that gets executed, everything can be passed for \<p>, not just a \<Person> instance, as illustrated in Fig.\,\ref{figure.solidity_problem}.
+The compiler cannot even enforce strong typing by generating defensive type instance checks and casts, because values are unboxed in Ethereum: they have no attached type information at run time, they are just numerical \emph{addresses}.
+It follows that inside the \<child> method, an eventual call to a \<Person>'s method on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
+In other words, Solidity is not strongly typed.
+Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed to another method, such as on \<p> passed to \<child>, since an attacker can pass crafted objects for \<p>, with arbitrary implementations for their methods, which can result in the unexpected execution of dangerous code.
+This actually happened in the case of the infamous DAO hack~\cite{dao16}, that costed millions of dollars.
 
@@
 smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
-In Java source code, generics are strongly typed, if no \emph{unchecked operations}
-are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics might have security issues
-at the level of compiled Java code and this paper originated from a
-real issue that we found in our code.
+%
+Through the use of generics, it is possible to provide to users a set of predefined contract templates which can be extended and specialized by users with lower programming skills, but higher knowledge about the specific application domain.
+They are based on the use of type placeholders in order
+to produce parametrized code, that can be instantiated for each
+concrete type provided for the placeholders.
+%
+However, strong typing and generics are two intertwined language features that have to be carefully considered when smart contracts are implemented and their bytecode is subsequently deployed on a blockchain.
+For instance, in Java source code, generics are strongly typed, if no \emph{unchecked operations} are used~\cite{NaftalinW06}, as it will always be the case in this paper.
+However, generics might have security issues at the level of compiled Java code and this paper originated from a real issue that we found in our code.
 
@@
 In comparison to that paper, Sec.~\ref{sec:takamaka} and
-Sec.~\ref{sec:validators} are new; eight figures have been added;
-all other sections have been expanded.
+Sec.~\ref{sec:validators} are new; while all other sections have been expanded and the enriched with several explanatory figures.

### 2022-03-02 cluster_computing/017_takamaka.tex
latex style
title, abstract and introduction rewriting + biblio fix

@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
   \begin{center}
@@
 Their source code can be found inside the Java project \textsf{\url{https://github.com/Hotmoka/hotmoka/tree/master/io-takamaka-code}}.}\label{fig:hierarchy-entities}
-\end{figure}
+\end{figure*}
 
@@
 \centering
-\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
+\includegraphics[width=\linewidth]{figures/takamaka_payable_contract}
 \caption{Exemplification of the Takamaka persistent objects stored in the blockchain.}
@@
 
-public final class TendermintED25519Validator extends Validator {
+public final class TendermintED25519Validator 
+       extends Validator {
   private final String id;
 
-  public TendermintED25519Validator(String publicKey) {
+  public TendermintED25519Validator
+    (String publicKey) {
     super(publicKey);
 
-    MessageDigest sha256 = MessageDigest.getInstance("SHA-256");
-    sha256.update(Base64.getDecoder().decode(publicKey()));
-    this.id = bytesToHex(sha256.digest()).substring(0, 40);
+    MessageDigest sha256 = 
+      MessageDigest.getInstance("SHA-256");
+    sha256.update(Base64.getDecoder().
+      decode(publicKey()));
+    this.id = bytesToHex(sha256.digest()).
+      substring(0, 40);
   }
@@
 For instance, the subclass
-\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
+\<io.takamaka.code.governance.tendermint.Tender\-mintED25519Validator>
 implements \<id> as for the Tendermint blockchain engine, that is,

### 2022-03-02 cluster_computing/020_shared_entities_with_generics.tex
title, abstract and introduction rewriting + biblio fix

@@
-\section{Shared Entities using Generics}\label{sec:shared_entities}
+\section{A Generics Shared Entities Implementation}\label{sec:shared_entities}
 
@@
 
+In general, two concepts are specific to each implementation of shared entities:
+who are the potential shareholders and how offers for selling shares work.
+Therefore, one can parameterize the interface of a shared entity with two type variables:
+\<S> is the type of the shareholders and \<O> is the type of the sale offers of shares.
+
 \begin{figure}[ht]
+\centering
+\includegraphics[width=\linewidth]{figures/shared_entity}
+\caption{Exemplification of a Takamaka shared entity and of its connections with the objects persisted in the blockchain.}
+\label{figure.shared_entity}
+\end{figure}
+
+The \<SharedEntityView> interface at the top of the hierarchy in Fig.~\ref{fig:hierarchy-entities}
+defines the read-only operations on a shared entity. This view is \emph{static}, in the sense that it
+does not specify the operations for transfers of shares. Therefore, its only type parameter is \<S>:
+any contract can play the role of the type for the shareholders of the entity.
+Method \<getShares> yields a snapshot of the
+current shares of the entity (who owns how much). Method \<getShareholders> yields the shareholders.
+It is not \<@View>, since it creates a new stream, which is a side-effect.
+Method \<isShareholder> checks if an object is a shareholder. Method \<sharesOf> yields
+the number of shares of a shareholder. As typical in Takamaka, a \<snapshot> method allows one
+to create a frozen read-only copy of an entity (in constant time), useful when an entity must be queried from
+a client without the risk of race conditions if another client is modifying the same entity concurrently.
+
+\begin{figure*}[ht]
   \begin{center}
     \begin{lstlisting}[language=Takamaka]
-public interface SharedEntity
-      <S extends PayableContract, O extends Offer<S>>
+public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
   extends SharedEntityView<S> {
@@
   \caption{A simplified part of our shared entity interface.}\label{fig:shared_entity}
-\end{figure}
-
-In general, two concepts are specific to each implementation of shared entities:
-who are the potential shareholders and how offers for selling shares work.
-Therefore, one can parameterize the interface of a shared entity with two type variables:
-\<S> is the type of the shareholders and \<O> is the type of the sale offers of shares.
-
-The \<SharedEntityView> interface at the top of the hierarchy in Fig.~\ref{fig:hierarchy-entities}
-defines the read-only operations on a shared entity. This view is \emph{static}, in the sense that it
-does not specify the operations for transfers of shares. Therefore, its only type parameter is \<S>:
-any contract can play the role of the type for the shareholders of the entity.
-Method \<getShares> yields a snapshot of the
-current shares of the entity (who owns how much). Method \<getShareholders> yields the shareholders.
-It is not \<@View>, since it creates a new stream, which is a side-effect.
-Method \<isShareholder> checks if an object is a shareholder. Method \<sharesOf> yields
-the number of shares of a shareholder. As typical in Takamaka, a \<snapshot> method allows one
-to create a frozen read-only copy of an entity (in constant time), useful when an entity must be queried from
-a client without the risk of race conditions if another client is modifying the same entity concurrently.
+\end{figure*}
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=.9\linewidth]{figures/shared_entity}
-\caption{Exemplification of a Takamaka shared entity and of its connections with the objects persisted in the blockchain.}
-\label{figure.shared_entity}
-\end{figure}
-
 As said before, the annotation \<@FromContract> on both \<place> and \<accept> enforces that only
@@
 
-\begin{figure}[tp]
+\begin{figure*}[tp]
   \begin{center}
@@
   \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
-\end{figure}
+\end{figure*}
 

### 2022-03-02 cluster_computing/040_solution.tex
title, abstract and introduction rewriting + biblio fix

@@
-\section{Fixing the Compilation of the Contract}\label{sec:fix}
+\section{A Solution for Fixing the Compilation of the Contract}\label{sec:fix}
 
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \begin{center}
@@
   \caption{The fixed code of the shared entity of the validators of a Hotmoka blockchain built over Tendermint.}\label{fig:solution}
-\end{figure}
+\end{figure*}
 
@@
    aload_3
-   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
+   invokevirtual accept
+       (BigInteger,
+        TendermintED25519Validator,
+        Offer)
    return
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \centering
-\includegraphics[width=.95\linewidth]{figures/solution}
+\includegraphics[width=0.9\linewidth]{figures/solution}
 \caption{Example of how the proposed solution works at run time when the \<accept> method is called by a \<Validator> or an \<Attacker>.}
 \label{figure.solution}
-\end{figure}
+\end{figure*}

### 2022-03-03 cluster_computing/001_abstract.tex
corrections of typos

@@
-Smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order automatically enforce some effects when particularly events occur.
-The writing of smart contracts is a complex and critical activity which can benefit from the use of high-level features of programming languages, like generics.
+Smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order to automatically enforce some effects when particular events occur.
+The writing of smart contracts is a complex and critical activity that can benefit from the use of high-level features of programming languages, like generics.
 In many programming languages, such as Java, generics are implemented by \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.

### 2022-03-03 cluster_computing/010_introduction.tex
corrections of typos

@@
 
-In recent years the Blockchain technology has rapidily emerged as a powerfull tool for supporting the development of many and innovative services and infrastructures.
+In recent years, blockchain technology has rapidly emerged as a powerful tool for supporting the development of many and innovative services and infrastructures.
 Blockchain-enabled applications are spreading across diverse sectors such as supply chain, business, healthcare, IoT, privacy, and data management \cite{blockchain_applications}.
@@
 A few years after Bitcoin, another blockchain, called Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming transactions in an actual, imperative and Turing-complete programming language, called Solidity.
-The major innovation of Ethereum is the contrustruction through its nodes of a distributed \emph{world computer} that can run general purpose code.
-Indee, if the term ``distributed ledger'' is usually used to describe blockchains like Bitcoin, Ethereum is often defined as a ``distributed state machine''.
-Solidity's code is organized in \emph{smart contracts}, namely pieces of code which are stored in the blockchain and are executed when a particular event occurs, e.g.
-when a transaction is scheduled.
+The major innovation of Ethereum is the construction through its nodes of a distributed \emph{world computer} that can run general-purpose code.
+Indeed, if the term \emph{distributed ledger} is usually used to describe blockchains like Bitcoin, Ethereum is often defined as a \emph{distributed state machine}.
+Solidity's code is organized in \emph{smart contracts}, namely pieces of code that are stored in the blockchain and are executed when a particular event occurs, e.g.\ when a transaction is scheduled.
 From a theoretical point of view, a smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.
 Through smart contracts, Ethereum's transactions can hence execute much more than coin transfers.
-In this case the global shared state is given by a set of objects which are persisted and manipulated in the same way by all nodes in the blockchain through the execution of the same object constructors and methods.
+In this case the global shared state is given by a set of objects that are persisted and manipulated in the same way by all nodes in the blockchain through the execution of the same object constructors and methods.
 
 The world computer built by Ethereum is known as Ethereum Virtual Machine (EVM) which is the platform where accounts and smart contracts live and are executed.
-Solidity is high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
-At this regards, we can observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
+Solidity is a high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
+At this regard, we can observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
 For instance, a Solidity method \<child(Person p, uint256 n) returns Person> actually compiles into \<child(address p, uint256 n) returns address>, losing most type information~\cite{CrafaPZ19}.
@@
 type variables. Generics are a powerful and very useful facility for programming
-smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
+smart contracts, since they allow one to personalize the behavior of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
 %
-Through the use of generics, it is possible to provide to users a set of predefined contract templates which can be extended and specialized by users with lower programming skills, but higher knowledge about the specific application domain.
+Through the use of generics, it is possible to provide to users a set of predefined contract templates that can be extended and specialized by users with lower programming skills, but higher knowledge about the specific application domain.
 They are based on the use of type placeholders in order

### 2022-03-03 cluster_computing/015_generics_in_java.tex
title, abstract and introduction rewriting + biblio fix
corrections of typos

@@
-\section{Generics Management in Java}\label{sec:java_generics}
+\section{Generics Implementation in Programming Languages}\label{sec:java_generics}
 
@@
 
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=0.8\linewidth]{figures/java_generics_erasure}
+\caption{Example of generics implementation by erasure in Java.}
+\label{figure.java_generics_erasure}
+\end{figure*}
+
 In order to understand the mechanism of erasure, consider for instance
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=.9\linewidth]{figures/java_generics_erasure}
-\caption{Example of generics implementation by erasure in Java.}
-\label{figure.java_generics_erasure}
-\end{figure}
-
 Erasure weakens the type information of the compiled code. It is the responsibility of the
@@
 However, this guarantee applies to Java source code compiled by the Java compiler,
-not to bytecode that can be generated manually, in order to attack instance of
+not to bytecode that can be generated manually, in order to attack instances of
 the \<SharedEntity> class, as we will show later.

### 2022-03-03 cluster_computing/017_takamaka.tex
corrections of typos

@@
 \<io.takamaka.code.governance.tendermint.Tender\-mintED25519Validator>
-implements \<id> as for the Tendermint blockchain engine, that is,
+implements \<id> as for the Tendermint blockchain engine~\cite{Kwon14},
+that is,
 as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key

### 2022-03-03 cluster_computing/050_related_work.tex
corrections of typos

@@
 it has been deployed, the immutability being one of its main characteristics, so that
-errors are treated as intended behaviours. Moreover, smart contracts often store and
+errors are treated as intended behaviors. Moreover, smart contracts often store and
 manage critical data such as money, digital assets and identities. For this reason, smart contracts

### 2022-03-05 cluster_computing/010_introduction.tex
cover letters + biogra

@@
 In comparison to that paper, Sec.~\ref{sec:takamaka} and
-Sec.~\ref{sec:validators} are new; while all other sections have been expanded and the enriched with several explanatory figures.
+Sec.~\ref{sec:validators} are new; while all other sections have been expanded with several additional details and enriched with many explanatory figures.

### 2022-03-05 cluster_computing/cover_letter.tex
cover letters + biogra

%
\documentclass[journal,onecolumn, 11pt]{IEEEtran}

\usepackage{times,epsf}
\usepackage{graphicx}
\usepackage{color}
\usepackage{url}

\usepackage{amsfonts}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}

\setlength{\textwidth}{5.5in}
\setlength{\hoffset}{0.4in}

\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}

\newcommand{\todo}[1]{\textcolor{red}{#1}}

%\makeatletter

% doublespace
% \def\baselinestretch{0.93}
%\def\baselinestretch{1.4}

% more than .95 of text and figures

\def\topfraction{.95}
\def\floatpagefraction{.95}
\def\textfraction{.05}

\newcommand{\BOX}[1]
{
  {\it
%    \medskip
    \begin{center}
      \begin{tabular}{|c|}
        \hline
        \parbox{0.97\columnwidth}{
          \medskip
          #1
          \medskip} \\
        \hline
      \end{tabular}
    \end{center}
%    \medskip
  }
}

\begin{document}

\title{{\LARGE On the Use of Generic Types for Smart Contracts}}

\author{{\normalsize F. Spoto, S. Migliorini, M. Gambini, A. Benini}}

\date{}

\maketitle

Dear Editors,

We are pleased to submit for your consideration our paper: ``On the Use of Generic Types for Smart Contracts''. This work addresses the benefits and weaknesses in using high-level programming featuare, like generics, in the development of blockchain-based smart contracts.

This work is an extended version of our previous paper ``Power and Pitfalls of Generic Smart Contracts'' published in the third IEEE International Conference on Blockchain Computing and Applications. In comparison to that paper, Sec. 3 and Sec. 5 are new; while all other sections have been expanded with several additional details and enriched with many explanatory figures.

\hspace{7cm}

Kinds regards.

The authors

\end{document}

### 2022-04-25 cluster_computing/001_abstract.tex
Started replying to cluster computing reviewers

@@
 In a permissionless blockchain, where every user can call the bytecode of smart contracts installed by other users, these weaker types pose a risk of attack.
-In this paper we start from a real security issue that we found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain, and we propose a possible solution based on an appropriate code rewriting.
+This paper focuses on a real security issue found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain, that can lead to stealing the remuneration of validator nodes, and proposes a patch based on appropriate code rewriting.

### 2022-04-25 cluster_computing/010_introduction.tex
Started replying to cluster computing reviewers

@@
 Solidity is a high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
-At this regard, we can observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
+At this regard, observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
 For instance, a Solidity method \<child(Person p, uint256 n) returns Person> actually compiles into \<child(address p, uint256 n) returns address>, losing most type information~\cite{CrafaPZ19}.
@@
 For instance, in Java source code, generics are strongly typed, if no \emph{unchecked operations} are used~\cite{NaftalinW06}, as it will always be the case in this paper.
-However, generics might have security issues at the level of compiled Java code and this paper originated from a real issue that we found in our code.
+However, generics might have security issues at the level of compiled Java code and this paper originated from a real issue that has been found in our code.
 

### 2022-04-25 cluster_computing/015_generics_in_java.tex
Started replying to cluster computing reviewers

@@
 The functionality of \<SharedEntity> will be discussed later (Sec.~\ref{sec:shared_entities}).
-Here, we only consider how its generic type parameters get compiled.
+Here, it is relevant to consider only how its generic type parameters get compiled.
 Namely, \<SharedEntity> uses two generic type parameters \<S> and \<O>, that must be provided
@@
 Such generic parameters have an upper bound: \<S> can only be a subtype of \<PayableContract>,
-while \<O> can only be a subtype of {\codesize\texttt{Offer<S>}}. If we check the bytecode
-generated for \<SharedEntity>, we will see that \<accept> is declared, in bytecode, as
+while \<O> can only be a subtype of {\codesize\texttt{Offer<S>}}. If one checks the bytecode
+generated for \<SharedEntity>, she will see that \<accept> is declared, in bytecode, as
 \<void accept(BigInteger amount, PayableContract buyer, Offer offer)>, that is,
@@
 not to bytecode that can be generated manually, in order to attack instances of
-the \<SharedEntity> class, as we will show later.
+the \<SharedEntity> class, as shown later.

### 2022-04-25 cluster_computing/017_takamaka.tex
Started replying to cluster computing reviewers

@@
 This section gives a short introduction to the Takamaka subset of
-Java that we use for writing smart contracts in this paper.
+Java that this paper uses for writing smart contracts.
 This language has been introduced in~\cite{Spoto19}. A full tutorial
@@
 The hierarchy of the classes described in this section is in Fig.~\ref{fig:hierarchy-entities}.
-In the following we will report a simplified presentation of the code of some
-of such classes. The full code is in the Github repository of Hotmoka~\cite{hotmoka_repository}.
+In the following, a simplified presentation of the code of some
+of such classes will be reported. The full code is in the Github repository of Hotmoka~\cite{hotmoka_repository}.
 

### 2022-04-25 cluster_computing/020_shared_entities_with_generics.tex
corrections of typos
Started replying to cluster computing reviewers

@@
 
-    public Offer(S seller, BigInteger sharesOnSale, 
-                 BigInteger cost,long duration) {
+    public Offer(S seller, BigInteger sharesOnSale, BigInteger cost,long duration) {
       this.seller = seller;
@@
 in the syntax \<S.class>. Due to this syntactical limitation of Java,
-the best we could write in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
+the best that can be written in Fig.~\ref{fig:shared_entity} is \<@FromContract(PayableContract.class)>,
 which allows \emph{any} \<PayableContract> to call these methods, not just those of type \<S>.
-Since the syntax of the language does not support our abstraction, we will have to
+Since the syntax of the language does not support the needed abstraction, one has to
 program explicit dynamic checks in code, as shown later, and this will be the reason of the
@@
 caller to have type \<S>, since the syntax of the language does not allow one to express it,
-as we explained before. Then \<accept> requires the \<offer> to exist, to be still ongoing
+as explained before. Then \<accept> requires the \<offer> to exist, to be still ongoing
 and to cost no more than the \<amount> of money provided to \<accept>. If that is the case,

### 2022-04-25 cluster_computing/030_generics_attack.tex
title, abstract and introduction rewriting + biblio fix
Started replying to cluster computing reviewers

@@
 %
-This property is important since it states that we can trust the type \<S> of
+This property is important since it states that one can trust the type \<S> of
 the shareholders: if one creates a \<SharedEntity> and fixes a specific type \<S>
@@
 its map \<shares> actually holds values of type \<S> in its domain, only.
-For this consistency result, we had to pay the price
-of the dummy \<buyer> argument for the method \<accept>
+For this consistency result, one needs
+the dummy \<buyer> argument for the method \<accept>
 of the shared entities. Without that argument, the
-\emph{Consistency of Shareholders} property would not hold, since we could only write
+\emph{Consistency of Shareholders} property would not hold, since one could only write
 \<addShares((S) caller(), offer.sharesOnSale)> in the implementation of \<accept> in
@@
 and become shareholders.
+
+\begin{figure}[ht]
+  \begin{center}
+    \begin{lstlisting}[language=Takamaka]
+public class Attacker extends ExternallyOwnedAccount {
+
+  public Validator(String publicKey) {
+    super(publicKey);
+  }
+
+  @View
+  public String id() {
+    return /* id of the victim blockchain node, unrelated to publicKey */
+  }
+}
+    \end{lstlisting}
+  \end{center}
+  \caption{An attacker that exploits the work of a blockchain validator node and fraudolently earns the rewards of that work.}\label{fig:attacker}
+\end{figure}
+
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=0.9\linewidth]{figures/attack}
+\caption{Example of possible attack to a smart contract using Java generics.}
+\label{figure.attack}
+\end{figure*}
 
@@
 the \emph{Consistency of Shareholders} property is easily violated at bytecode level.
-
-\begin{figure}[t]
-  \begin{center}
-    \begin{lstlisting}[language=Takamaka]
-public class Attacker extends ExternallyOwnedAccount {
-
-  public Validator(String publicKey) {
-    super(publicKey);
-  }
-
-  @View
-  public String id() {
-    return /* id of the victim blockchain node, unrelated to publicKey */
-  }
-}
-    \end{lstlisting}
-  \end{center}
-  \caption{An attacker that exploits the work of a blockchain validator node and fraudolently earns the rewards of that work.}\label{fig:attacker}
-\end{figure}
-
-\begin{figure}[b]
-\centering
-\includegraphics[width=.9\linewidth]{figures/attack}
-\caption{Example of possible attack to a smart contract using Java generics.}
-\label{figure.attack}
-\end{figure}
 

### 2022-04-25 cluster_computing/090_conclusion.tex
Started replying to cluster computing reviewers

@@
 \emph{useless} and just remove it. This would recreate the issue that
-we have just solved. That is, our solution works only for the way
+has been just solved. That is, the solution in this paper works only for the way
 compilers compile \emph{today}.
@@
 the encapsulation guarantees of the language. When embedded in a permissionless
-blockchain, such features become dangerous attack surfaces.
+blockchain, such features become dangerous attack surfaces. This paper has shown
+the attack surface due to redefinition of methods with a ganeric parameter.
+But another example is the use of instrumented methods to allow the access to
+private state from inner classes: since inner classes are compiled into distinct
+bytecode classes, the compiler adds non-private accessors to the
+private state. These accessors cannot be used
+at source level, but can be called at bytecode level to gain access to private state.
+This paper does not provide a solution to this other issue, but this further example
+makes it clear that the attack surface is larger than what described here.

### 2022-04-25 cluster_computing/010_introduction.tex
More replies to Clusetr Computing reviews

@@
 
-Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}  has been the first blockchain's success story.
-Here transactions are programmed in a non-Turing complete bytecode language, almost exclusively used to implement transfers of units of coins between \emph{accounts}, providing a totally decentralized P2P digital cash system based on a distributed public ledger.
+The rules of blockchain transactions are specified by \emph{smart contracts},
+that are code written in a variety of programming languages. To the best of our knowledge, none of them allows generic types and, in any case,
+nothing has been published about the opportunity but also the risks of using generic types for writing smart contracts.
+The contribution of this paper is exactly to show a real-life
+use of generic types for an actual smart contract contained in the support
+library of the Takamaka language for writing smart contracts~\cite{Spoto19,Spoto20},
+and to demonstrate that a na\"{i}ve use
+of Java generics can lead to a code security vulnerability that
+allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
+This paper provides a fix to that specific issue,
+by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
+More generally, this paper can be useful for the definition of
+bytecode languages for future smart contract languages, by
+learning from the weaknesses of Java bytecode, in particualr those related to the
+compilation of generics.
+
+Historically, programming languages for specifying blockchain transactions started with
+Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}, the first blockchain's success story.
+Here transactions are programmed in a non-Turing complete bytecode language, with no notion of generic types, almost exclusively used to implement transfers of units of coins between \emph{accounts}, providing a totally decentralized P2P digital cash system based on a distributed public ledger.
 %
-A few years after Bitcoin, another blockchain, called Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming transactions in an actual, imperative and Turing-complete programming language, called Solidity.
+A few years after Bitcoin, another blockchain, called Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming transactions in an actual, imperative and Turing-complete programming language, called Solidity, also missing generic types.
 The major innovation of Ethereum is the construction through its nodes of a distributed \emph{world computer} that can run general-purpose code.
@@
 Solidity is a high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
-At this regard, observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
+At this regard, observe that, in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type.
 For instance, a Solidity method \<child(Person p, uint256 n) returns Person> actually compiles into \<child(address p, uint256 n) returns address>, losing most type information~\cite{CrafaPZ19}.
@@
 The compiler cannot even enforce strong typing by generating defensive type instance checks and casts, because values are unboxed in Ethereum: they have no attached type information at run time, they are just numerical \emph{addresses}.
-It follows that inside the \<child> method, an eventual call to a \<Person>'s method on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
+It follows that, inside the \<child> method, an eventual call to a \<Person>'s method on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
 In other words, Solidity is not strongly typed.
@@
 
-Strong typing is one of the reasons that push towards the adoption
+Strong typing is one of the reasons that pushed towards the adoption
 of \emph{traditional} programming languages for smart contracts. For instance,
@@
 such as \emph{generics}\ie the possibility of using
-type variables. Generics are a powerful and very useful facility for programming
+generic type variables. Generics are a powerful and very useful facility for programming
 smart contracts, since they allow one to personalize the behavior of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
@@
 However, generics might have security issues at the level of compiled Java code and this paper originated from a real issue that has been found in our code.
-
-The contribution of this paper is to show a real-life
-use of generics for an actual smart contract contained in the support
-library of the Takamaka language, and to demonstrate that a na\"{i}ve use
-of Java generics can lead to a code security vulnerability that
-allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
-This paper will provide a fix to that specific issue,
-by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
-More generally, this paper can be useful for the definition of
-bytecode languages for future smart contract languages, by
-learning from the weaknesses of Java bytecode.
 

### 2022-04-27 cluster_computing/001_abstract.tex
Further edits to the Cluster Computing version

--- 2022-03-03 cluster_computing/001_abstract.tex
@@
-Smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order to automatically enforce some effects when particular events occur.
-The writing of smart contracts is a complex and critical activity that can benefit from the use of high-level features of programming languages, like generics.
+This paper shows that generic types (\emph{generics}) are useful for writing more abstract and more general smart contracts, but that this comes with some security risk, reporting a concrete security issue found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain.
+That issue can be used to steal the remuneration of validator nodes.
+This paper proposes a patch based on appropriate code rewriting.
+Namely, smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order to automatically enforce some effects when particular events occur.
+The writing of smart contracts is a complex and critical activity that can benefit from the use of high-level features of programming languages, and generics is one of them.
 In many programming languages, such as Java, generics are implemented by \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
 This is safe at source level, since the compiler takes care of checking that types are correct, before erasure.
-However, the types of the generated bytecode are erased and consequently weaker.
+However, the erased types of the generated bytecode are consequently weaker.
 In a permissionless blockchain, where every user can call the bytecode of smart contracts installed by other users, these weaker types pose a risk of attack.
-In this paper we start from a real security issue that we found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain, and we propose a possible solution based on an appropriate code rewriting.

### 2022-04-27 cluster_computing/010_introduction.tex
Further edits to the Cluster Computing version

@@
 %
-The key innovation introduced by this technology is a mechanism  able to reach an emergent agreement about a global state without the need for a central authority.
+That is, the key innovation introduced by this technology is a mechanism  able to reach an emergent agreement about a global state without the need for a central authority.
 Morever, another peculiarity is that the consensus is not explicit, because there is not a fixed moment when it occurs.
@@
 The rules of blockchain transactions are specified by \emph{smart contracts},
-that are code written in a variety of programming languages. To the best of our knowledge, none of them allows generic types and, in any case,
-nothing has been published about the opportunity but also the risks of using generic types for writing smart contracts.
+that are code written in a variety of programming languages. To the best of our knowledge, none of them allows generic types (\emph{generics})
+and, in any case,
+nothing has been published about the opportunity but also the risks of using generics for writing smart contracts.
 The contribution of this paper is exactly to show a real-life
-use of generic types for an actual smart contract contained in the support
-library of the Takamaka language for writing smart contracts~\cite{Spoto19,Spoto20},
+use of generics for an actual smart contract contained in the support
+library of the Takamaka language~\cite{Spoto19,Spoto20},
 and to demonstrate that a na\"{i}ve use
@@
 More generally, this paper can be useful for the definition of
-bytecode languages for future smart contract languages, by
-learning from the weaknesses of Java bytecode, in particualr those related to the
+future bytecode languages for smart contract languages, by
+learning from the weaknesses of Java bytecode, in particular from those related to the
 compilation of generics.
@@
 
-The world computer built by Ethereum is known as Ethereum Virtual Machine (EVM) which is the platform where accounts and smart contracts live and are executed.
+The world computer built by Ethereum is known as Ethereum Virtual Machine (EVM) and is the platform where accounts and smart contracts live and are executed.
 Solidity is a high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM.
@@
 language features, that are missing in Solidity,
-such as \emph{generics}\ie the possibility of using
-generic type variables. Generics are a powerful and very useful facility for programming
+such as generics. They are a powerful and very useful facility for programming
 smart contracts, since they allow one to personalize the behavior of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
 %
-Through the use of generics, it is possible to provide to users a set of predefined contract templates that can be extended and specialized by users with lower programming skills, but higher knowledge about the specific application domain.
-They are based on the use of type placeholders in order
+Through the use of generics, it is possible to provide to users a set of predefined contract templates that they can extend and specialize with lower programming skills, but higher knowledge about the specific application domain.
+Generics are based on the use of type placeholders in order
 to produce parametrized code, that can be instantiated for each
@@
 Sec.~\ref{sec:shared_entities} shows our real-life Java smart
-contracts for shared entities, that use generic types.
+contracts for shared entities, that use generics.
 Sec.~\ref{sec:validators} shows the instantiation of the shared entities to implement the validators' set
@@
 Sec.~\ref{sec:attack} shows that a na\"{i}ve
-deployment of a subclass of the validators' set leads to a code vulnerability.
+deployment of a subclass of the validators' set leads to a code vulnerability
+due to the way generics are compiled.
 Sec.~\ref{sec:fix} presents a fix to that vulnerability.

### 2022-04-27 cluster_computing/020_shared_entities_with_generics.tex
Further edits to the Cluster Computing version

@@
-\section{A Generics Shared Entities Implementation}\label{sec:shared_entities}
+\section{A Generic Shared Entities Implementation}\label{sec:shared_entities}
 
@@
 
-Fig.~\ref{fig:simple_shared_entity} shows a portion of our \<SimpleSharedEntity>
+Fig.~\ref{fig:simple_shared_entity} shows a portion of the code of our \<SimpleSharedEntity>
 implementation of the \<SharedEntity> interface in Fig.~\ref{fig:shared_entity}, that

### 2022-04-27 cluster_computing/030_generics_attack.tex
Further edits to the Cluster Computing version

@@
 
-Let us state an important property about shared entities:
+Let us state an important, expected property about shared entities:
 %
@@
   \noindent
-  If \<se> is a {\codesize\texttt{SharedEntity<S,O>}}
+  If \<se> is a {\codesize\texttt{SharedEntity<S,O>} object,}
   then \<se.getShareholders()> contains only elements of type \<S>.
@@
 \includegraphics[width=0.9\linewidth]{figures/attack}
-\caption{Example of possible attack to a smart contract using Java generics.}
+\caption{Example of possible attack to a smart contract that uses Java generics.}
 \label{figure.attack}

### 2022-04-27 cluster_computing/050_related_work.tex
Further edits to the Cluster Computing version

@@
 
-It has been estimated that, on average, software developers make from 100 to 150 errors
+Programming languages specific to smart contracts (such as Solidity)
+do not have generic types. General-purpose programming languages do have
+generic types in most cases, but are much less frequently used for writing smart contracts.
+In any case, we are not aware of any scientific work on the use of generic
+types for writing more generic smart contracts, nor of any study on the security
+risks, and their solutions, that this implies for the resulting smart contracts. From this point of view,
+the present paper has no direct literature to compare with.
+
+It is possible instead to insert this paper in the broader context of software correctness and security.
+Namely, it has been estimated that, on average, software developers make from 100 to 150 errors
 for every thousand lines of code~\cite{software_engineering}.

### 2022-04-27 cluster_computing/090_conclusion.tex
Further edits to the Cluster Computing version

@@
 
-This paper has shown that generic types can be useful in the definition
+This paper has shown that generics are useful in the definition
 of smart contracts and can simplify the development of rather complex
-code such as those for shared entities, and support code reuse, for
+code such as that for shared entities, and support code reuse, for
 instance to implement the validators set of a blockchain network.
@@
 
-The solution in this paper has been to redefine a method with an argument
-of generic type in such a way to call its superclass
-(see \<accept> in Fig.~\ref{fig:solution}). This fixes the security risk,
+The solution in this paper has been to redefine the methods that have an argument
+of generic type, in such a way to call their superclass
+(see the case of \<accept> in Fig.~\ref{fig:solution}). This fixes the security risk,
 but cannot be regarded as the definite solution to the problem. It is just a trick that
@@
 blockchain, such features become dangerous attack surfaces. This paper has shown
-the attack surface due to redefinition of methods with a ganeric parameter.
-But another example is the use of instrumented methods to allow the access to
+the attack surface due to redefinition of methods with a generic parameter.
+But another example is the use of instrumented methods to allow access to
 private state from inner classes: since inner classes are compiled into distinct

### 2022-05-02 cluster_computing/000_answers_to_reviews.tex
minor_revision_figure_related_work_check

%
\documentclass[journal,onecolumn, 11pt]{IEEEtran}

\usepackage{times,epsf}
\usepackage{graphicx}
\usepackage{color}
\usepackage{url}

\usepackage{amsfonts}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}

\setlength{\textwidth}{5.5in}
\setlength{\hoffset}{0.4in}

\setlength{\oddsidemargin}{0.0in}
\setlength{\evensidemargin}{0.0in}

\newcommand{\todo}[1]{\textcolor{red}{#1}}

%\makeatletter

% doublespace
% \def\baselinestretch{0.93}
%\def\baselinestretch{1.4}

% more than .95 of text and figures

\def\topfraction{.95}
\def\floatpagefraction{.95}
\def\textfraction{.05}

\newcommand{\BOX}[1]
{
  {\it
%    \medskip
    \begin{center}
      \begin{tabular}{|c|}
        \hline
        \parbox{0.97\columnwidth}{
          \medskip
          #1
          \medskip} \\
        \hline
      \end{tabular}
    \end{center}
%    \medskip
  }
}

\begin{document}

\title{{\LARGE On the Use of Generic Types for Smart Contracts}}

\author{{\normalsize Fausto Spoto, Sara Migliorini, Mauro Gambini and Andrea Benini
\\}\vspace{5mm}{\normalsize {\bf Answers to Reviewer's Comments}}}

\date{}

\maketitle

Dear Editor,

\vspace{2mm}

we would like to submit a new version of our paper ``On the Use of Generic Types for Smart Contracts''. We thank you and the reviewers for the comments to our submission. In particular, in this revised manuscript, we have carefully considered the reviewer comments and we have provide a detailed answer to their comments in what follows.

\vspace{5mm}

We thank you and the reviewers in advance for your work.
\vspace{5mm}

        \hspace{7cm} The authors:
\vspace{5mm}

                \hspace{7cm} Fausto Spoto,

                \hspace{7cm} Sara Migliorini,

                \hspace{7cm} Mauro Gambini

                \hspace{7cm} Andrea Benini.

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{\textbf{Reviewer 1}}

\BOX{
Thank you for your submission. I highly advice you to explain a little bit more about the attack surfaces.
}

We have expanded the Conclusion section with the following paragraph:
``This paper has shown the attack surface due to redefinition of methods with a generic parameter. But another example is the use of instrumented methods to allow access to private state from inner classes: since inner classes are compiled into distinct
bytecode classes, the compiler adds non-private accessors to the private state. These accessors cannot be used at source level, but can be called at bytecode level to gain access to private state.
This paper does not provide a solution to this other issue, but this further example
makes it clear that the attack surface is larger than what described here.''

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{5mm}
\hrule
\section*{\textbf{Reviewer 3}}

\BOX{
This manuscript present a real security issue that the authors found while using generics for writing smart contracts that implement shared entities for the Hotmoka blockchain, and they proposed a possible solution based on an appropriate code rewriting.

They are many major comments in here. I bring as much as I can.
}

We thank the reviewer for the comments. We tried to answer all of them in the following.

\BOX{
1 -- The abstract section should be rewritten in order to clearly state the manuscript main focus.
}

We have rewritten the abstract, stating its focus from its very first line:
``This paper shows that generic types (\emph{generics}) are useful for writing more abstract and more general smart contracts, but this comes with some security risks, reporting a concrete security issue found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain. That issue can be used to steal the remuneration of validator nodes. This paper proposes a patch based on appropriate code rewriting. Namely, smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order to automatically enforce some effects when particular events occur. The writing of smart contracts is a complex and critical activity that can benefit from the use of high-level features of programming languages, and generics is one of them. In many programming languages, such as Java, generics are implemented by \emph{erasure}, i.e., replaced by their upper bound type during compilation into bytecode. This is safe at source level, since the compiler takes care of checking that types are correct, before erasure. However, the erased types of the generated bytecode are consequently weaker. In a permissionless blockchain, where every user can call the bytecode of smart contracts installed by other users, these weaker types pose a risk of attack.''

\BOX{
2 -- The manuscript should be more organized to make it more interesting for the readers.
}

We agree that there is always space for improvement and better presentation of the technical material. Still, this comment does not guide us to what exactly should be organized and made more interesting. A less generic comment would help us. In any case, we have reorganized abstract and introduction following the other comments.

\BOX{
3 -- The introduction should be rewritten to clarify its message. Drawbacks of former proposals should be clearly indicated and innovations and new ideas highlighted.
}

We have reorganized the introduction and placed the following sentence at its beginning, to clarify the message. Note that there are no former proposals, since generics have not been used to write smart contracts up to now, to be best of our knowledge. This is expressed clearly now:
``The rules of blockchain transactions are specified by \emph{smart contracts},
that are code written in a variety of programming languages. To the best of our knowledge, none of them allows generic types and, in any case,
nothing has been published about the opportunity but also the risks of using generic types for writing smart contracts.
The contribution of this paper is exactly to show a real-life
use of generic types for an actual smart contract contained in the support
library of the Takamaka language,
and to demonstrate that a naive use
of Java generics can lead to a code security vulnerability that
allows an attacker to earn money by exploiting someone else's work, with both economical and legal side effects.
This paper provides a fix to that specific issue,
by proposing a re-engineering of the code that forces the compiler to generate defensive checks.
More generally, this paper can be useful for the definition of
future bytecode languages for smart contract languages, by
learning from the weaknesses of Java bytecode, in particular from those related to the
compilation of generics.''

\BOX{
4 -- Problem statement, objective \& contribution are not clear.
}

The same change to the introduction highlighted at point 3 above clarifies the statemente objective and contribution of the paper now.

\BOX{
5 -- If no one has proposed before a solution like the proposed solution, this claim should be highlighted much more. Else, it should be indicated who has done this, and it should be indicated what the innovations of the current paper are.
}

The same change to the introduction highlighted at point 3 above clarifies that generic types have not been used for writing smart contracts up to now and most programming
languages for smart contracts do not even allow the use of generics. We have also added a paragraph
at the beginning of the related work section that states that generics have not been used
for smart contracts up to now and their related risks have not been studied.

\BOX{6 -- For explaining the related work, the authors need to simply mentions their strategy in generic types or smart contracts, mention the advantages and disadvantages. I did not get the general view of the methods or solutions based the explanation that the authors have provided in this part. \\

\noindent 7 -- In my view, the literature survey of introduction is quite weak, unfocused and insufficient. The literature review needs to have a flow that leads to the objectives. What is the essential problem of this work? The authors should really explain the drawback of approaches in the related works especially instead of simply stating what they have done. Besides, please explain the main contribution related to previous approaches or solutions.\\

\noindent 8 -- Section "Related Work" make sure you (i.e., the authors) discuss and cite the most relevant works.}

All these comments about the related work section have pushed us to add a paragraph at its beginning, that clarifies that we have really nothing to compare with and we can only put our paper in the context of engineering techniques for strengthening the correctness of software. The new section now starts as follows: ``Programming languages specific to smart contracts (such as Solidity) do not have generic types. General-purpose programming languages do have generic types in most cases, but are much less frequently used for writing smart contracts. In any case, we are not aware of any scientific work on the use of generic types for writing more generic smart contracts, nor of any study on the security risks, and their solutions, that this implies for the resulting smart contracts. From this point of view, the present paper has no direct literature to compare with. It is possible instead to insert this paper in the broader context of software correctness and security. Namely...''.

Moreover, we have tried to contextualize the proposed solution with respect to the described technique, asserting that it could be used in the future to automatically overcome the vulnerability or guiding the definition of the next generation of programming languages for smart contracts.

\BOX{9 -- Avoid using single person pronouns like ``we'' throughout the manuscript.} We have chased and removed as many as possible of them, in the whole paper. Currently, only three ``we'' remain in the manuscript, in Section 8 ``Related Work'', since they state our personal view.

\BOX{10 -- Some terms are just used without explaining the meaning. It is not enough to bring a concept and just bring a reference. The authors need to explain a little bit about it.}

We have tried to imagine which are such terms. The manuscript is expected to be read by somebody working in the blockchain community, hence we do not explain all terms in that area. Yet, if there are specific terms to explain, we are happy to do it but this comment does not help us identifying the exact explanations that are missing.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{5mm}
\hrule
\section*{\textbf{Reviewer 4}}

\BOX{The manuscript is well written.  I accept the paper in its current form.
}

We would like to thank the reviewer for this comment.

\end{document}

### 2022-05-02 cluster_computing/001_abstract.tex
minor_revision_figure_related_work_check

@@
-This paper shows that generic types (\emph{generics}) are useful for writing more abstract and more general smart contracts, but that this comes with some security risk, reporting a concrete security issue found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain.
+This paper shows that generic types (\emph{generics}) are useful for writing more abstract and more general smart contracts, but this comes with some security risks, reporting a concrete security issue found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain.
 That issue can be used to steal the remuneration of validator nodes.

### 2022-05-02 cluster_computing/010_introduction.tex
minor_revision_figure_related_work_check

@@
 and, in any case,
-nothing has been published about the opportunity but also the risks of using generics for writing smart contracts.
+nothing has been published about the opportunity, but also the risks of using generics for writing smart contracts.
 The contribution of this paper is exactly to show a real-life

### 2022-05-02 cluster_computing/050_related_work.tex
minor_revision_figure_related_work_check

--- 2022-03-03 cluster_computing/050_related_work.tex
@@
 \section{Related Work}\label{sec:related_work}
+
+Programming languages specific to smart contracts (such as Solidity)
+do not have generic types. Conversely general-purpose programming languages do have
+generic types in most cases, but are much less frequently used for writing smart contracts.
+In any case, we are not aware of any scientific work on the use of generic
+types for writing more generic smart contracts, nor of any study on the security
+risks, and their solutions, that this implies for the resulting smart contracts. From this point of view,
+the present paper has no direct literature to compare with.
+%
+Anyway, it is possible instead to insert this paper in the broader context of software correctness and security.
 
@@
 
+\todo{The solution proposed in this paper could be implemented through an automatic bytecode rectification mechanism.
+Indeed, the additional method with a more restrictive signature could be automatically added in the bytecode without the need for an explicit method redefinition at the source code level.}
+
 %Given the plurality of verification tools that have been developed for indentifying security bugs
@@
 
-Finally, as regards to the definition of new programming languages for safe smart contracs,
+Finally, as regards to the definition of new programming languages for safe smart contracts,
 Scilla~\cite{scilla} has been tailored by taking System~F as a foundational calculus.
@@
 allow one to inspect the compiled bytecode, in order to understand how generic types are compiled.
+
+\todo{As regards to this last solution which is based on the definition of new programming languages specific for safe smart contracts writing, the proposed solution could guide a more sophisticated and consious bytecode generation.
+Indeed, the next generation of programming language for smart contracts should take in mind that the generated bytecode should be called directly, without passing from the source code and the compiler checks.
+Therefore, any checks which are possible at source code level compilation, like type checking, should remain possible also during bytecode execution.}

### 2022-05-02 cluster_computing/090_conclusion.tex
minor_revision_figure_related_work_check

@@
 makes it clear that the attack surface is larger than what described here.
+
+\section{Compliance with Ethical Standards}
+
+This work was partially supported by ``Progetto di Eccellenza'' of the Computer Science Department of University of Verona, Italy.
+
+The authors declare that they have no conflicts of interest.
+
+This article does not contain any studies involving animals performed by any of the authors.
+
+This article does not contain any studies involving human participants performed by any of the authors
+
+\section{Data Availability Statements}
+
+The source code are freely available at \url{https://github.com/Hotmoka/hotmoka/tree/master/io-takamaka-code}

### 2022-05-02 cluster_computing/000_answers_to_reviews.tex
Corrected typos

@@
 We thank you and the reviewers for the comments to our submission.
-In particular, in this revised manuscript, we have carefully considered the reviewer comments and we have provide a detailed answer to their comments in what follows.
+In particular, in this revised manuscript, we have carefully considered the reviewer comments and we have provided a detailed answer to their comments in what follows.
 
@@
 \BOX{
-This manuscript present a real security issue that the authors found while using generics for writing smart contracts that implement shared entities for the Hotmoka blockchain, and they proposed a possible solution based on an appropriate code rewriting.
+This manuscript presents a real security issue that the authors found while using generics for writing smart contracts that implement shared entities for the Hotmoka blockchain, and they proposed a possible solution based on an appropriate code rewriting.
 
-They are many major comments in here. I bring as much as I can.
+There are many major comments in here. I bring as much as I can.
 }
@@
 
-The same change to the introduction highlighted at point 3 above clarifies the statemente objective and contribution of the paper now.
+The same change to the introduction highlighted at point 3 above clarifies statement, objective and contribution of the paper now.
 
@@
 
-All these comments about the related work section have pushed us to add a paragraph at its beginning, that clarifies that we have really nothing to compare with and we can only put our paper in the context of engineering techniques for strengthening the correctness of software.
+All these comments about the related work section have pushed us to add a paragraph at its beginning, that clarifies that we have really nothing to compare with directly and we can only put our paper in the context of engineering techniques for strengthening the correctness of software.
 The new section now starts as follows: ``Programming languages specific to smart contracts (such as Solidity) do not have generic types.

### 2022-05-02 cluster_computing/050_related_work.tex
Corrected typos

@@
 
-\todo{The solution proposed in this paper could be implemented through an automatic bytecode rectification mechanism.
-Indeed, the additional method with a more restrictive signature could be automatically added in the bytecode without the need for an explicit method redefinition at the source code level.}
+The solution proposed in this paper could be implemented through an automatic bytecode rectification mechanism.
+Indeed, the additional method with a more restrictive signature could be automatically added in the bytecode without the need for an explicit method redefinition at the source code level.
 
@@
 
-\todo{As regards to this last solution which is based on the definition of new programming languages specific for safe smart contracts writing, the proposed solution could guide a more sophisticated and consious bytecode generation.
-Indeed, the next generation of programming language for smart contracts should take in mind that the generated bytecode should be called directly, without passing from the source code and the compiler checks.
-Therefore, any checks which are possible at source code level compilation, like type checking, should remain possible also during bytecode execution.}
+As regards to this last solution, which is based on the
+definition of new programming languages specific for writing
+safe smart contracts, the proposed solution could guide
+a more sophisticated and conscious bytecode generation.
+Indeed, the next generation of programming languages for smart contracts
+should take in mind that the generated bytecode should be called directly,
+without passing from the source code and the compiler checks.
+Therefore, any checks that are possible at source-code level,
+such as type checking, should remain possible also during bytecode execution.
